---
title: "Setup: getting the data onto your machine"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

This notebook introduces the dataset the tutorial uses, shows how to get the raw
files onto your own machine, and sets out what running everything costs in time,
memory, and disk. It assumes the array background from
[Background](00a_background.qmd).

::: {.callout-important}
## Where to type each block
Every code block in this chapter carries a grey label telling you where it runs. There are
only two places, and mixing them up is the most common way to get stuck:

- **RStudio Console** — the pane with the `>` prompt, where R lives. Every block of R code
  goes here.
- **Terminal** / **PowerShell** — a system shell, for the handful of non-R commands such as
  `git clone`. On macOS that is Terminal.app, on Windows it is PowerShell, on Linux it is
  whatever shell you use.

One trap worth naming: RStudio has a **Terminal** tab sitting right next to the Console,
and it is a system shell, not R. R code pasted there will fail with a string of syntax
errors — so when a block says *RStudio Console*, make sure you have clicked the Console
tab. If you are not using RStudio, `R` at a terminal gives you the same `>` prompt and
works fine.
:::

::: {.callout-note collapse="true"}
## One thing that runs before each chapter, and what it is not allowed to do

Every chapter begins with a hidden chunk that does `source("_setup.R")`. That file is
deliberately limited to two housekeeping jobs: it puts the library holding the EPIC
annotation packages on R's search path, and it registers the figure font and ggplot theme
so the plots share the site's look. It also quietly attaches the packages the theme needs,
but no chapter relies on that: every chapter was checked by running its code without
`_setup.R` at all.

What it does **not** do is supply anything the code you read depends on. Each chapter
loads the packages it uses with its own visible `library()` call, writes its file paths out
in full as `data/<file>`, and names colors by hex code. So if you copy a block out of a
chapter into a fresh R session, it runs — you are never silently relying on something
defined in a file you cannot see. If a chapter's code ever fails for you with
`object not found` or `could not find function`, that is a bug in the chapter, not
something you are missing.
:::

## Building the software environment {#sec-env}

Everything except the pipeline chapter needs **only R**. If you are new to computing,
start there and ignore conda entirely — you will still be able to work through
preprocessing, QC, cell composition, batch effects, the EWAS itself, and annotation.

::: {.panel-tabset}

### R only (recommended if you are getting started)

You need **R 4.3 or newer** ([download from CRAN](https://cran.r-project.org/)) — install
the current release — and, if you want,
[RStudio](https://posit.co/download/rstudio-desktop/). Everything below is typed
into the **Console** pane — the one with the `>` prompt.

One note on versions: the published numbers and the Zenodo checkpoints were produced on
R 4.2.3 (Bioconductor 3.16), so a value may differ in its last digits on a newer stack.
The code itself is unchanged. See
[`SESSIONINFO.md`](https://github.com/krferrier/Methylation-EWAS-tutorial/blob/main/tutorial/SESSIONINFO.md).

You need four CRAN and eighteen Bioconductor packages. The whole installation is short
enough to read, so **option 1 is to select the block below, copy it, and paste it into the
Console.** Nothing to download first, nothing to configure, and you can see exactly what
it does before you run it. Press Enter and wait.

```{.r filename="RStudio Console"}
repos <- "https://cloud.r-project.org"
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager", repos = repos)

BiocManager::install(c(
  "data.table", "ggplot2", "knitr", "DT",
  "minfi", "illuminaio", "wateRmelon", "GEOquery", "Biobase",
  "IlluminaHumanMethylationEPICmanifest",
  "IlluminaHumanMethylationEPICanno.ilm10b4.hg19",
  "FlowSorted.Blood.EPIC", "genefilter",
  "sva", "limma", "bacon",
  "sesame", "sesameData", "GenomicRanges", "rtracklayer",
  "missMethyl", "methylGSA"
), ask = FALSE, update = FALSE, upgrade = "never")
```

Expect 20-40 minutes the first time. On Windows and macOS most of these arrive as
pre-built binaries — there are simply a lot of them — while on Linux they build from
source, which is slower. You will see a great deal of output scroll past; that is normal.

::: {.callout-note collapse="true"}
## Other ways to run the same installation
The block above is all you need. These alternatives do exactly the same thing, and exist
only if you prefer them.

**Option 2 — run the shipped script without downloading it.** The same commands live in
[`install_packages.R`](https://github.com/krferrier/Methylation-EWAS-tutorial/blob/main/tutorial/install_packages.R),
which also checks your R version first. One line in the Console fetches and runs it:

```{.r filename="RStudio Console"}
source("https://raw.githubusercontent.com/krferrier/Methylation-EWAS-tutorial/main/tutorial/install_packages.R")
```

**Option 3 — run your local copy of that script.** If you cloned the repository (see
@sec-sample-sheet-get), set the working directory to `tutorial/` first — **Session → Set
Working Directory → Choose Directory…**, or open `tutorial/` as an RStudio Project so it
is set for you every time — then:

```{.r filename="RStudio Console"}
source("install_packages.R")
```

**Option 4 — from a terminal.** The same script runs outside RStudio, from `tutorial/`:

```{.bash filename="Terminal"}
Rscript install_packages.R
```
:::

The non-R sections: **chapter 07** runs the EWAS through a Snakemake pipeline, and that
genuinely needs conda. Chapter 06 already runs the same association test directly in R,
so nothing analytical is lost — chapter 07 shows how to scale it up. The `comb-p`
region-level section of chapter 08 also needs conda; the rest of that chapter does not.

### conda

If you already use conda, or you want to run chapter 07, four environments are defined in
the repository's `envs/` directory. Three of them exist because their toolchains cannot
coexist: Snakemake needs Python 3, `comb-p` runs on Python 2.7, and `methylGSA` is only
packaged for an older R than the rest of the stack.

```{.bash filename="Terminal"}
conda env create -f envs/methyl.yml       # chapters 01-06, 08   -- R 4.5
conda env create -f envs/smk.yml          # chapter 07           -- Snakemake
conda env create -f envs/combp.yml        # chapter 08 DMRs      -- comb-p
conda env create -f envs/methylgsa.yml    # chapter 08 methylGSA -- R 4.2
```

| Environment | Chapters | Core tool |
|---|---|---|
| `ewas-methyl` | 01-06, 08 | R 4.5, Bioconductor 3.22 |
| `ewas-smk` | 07 | Snakemake 9.26.1 |
| `ewas-combp` | 08 (DMR section) | comb-p on Python 2.7 |
| `ewas-methylgsa` | 08 (methylGSA section) | methylGSA on R 4.2 |

If you only want the R chapters, `envs/methyl.yml` is the only one you need. `methylGSA`
is one of three probe-bias corrections chapter 08 compares, and the other two —
`gometh` and the sesame-based enrichment — run in the main environment.

Each file lists only the packages the tutorial uses directly and lets conda resolve the
rest. `envs/smk.yml` deliberately holds nothing but the workflow engine: the pipeline
declares its own per-rule R environment and Snakemake builds it on first run, which is
also where METAL is compiled from source. `envs/methylgsa.yml` is split out for a
packaging reason rather than a scientific one: bioconda's newest `methylGSA` build is for
R 4.2, so keeping it in `envs/methyl.yml` would hold that whole environment a
Bioconductor generation back. Readers on the R-only route never see this — Bioconductor
3.23 ships a current `methylGSA`, so `BiocManager` installs it alongside everything else.

If you do not have conda, [Miniforge](https://github.com/conda-forge/miniforge) is a
minimal installer that defaults to the conda-forge channel; its README covers macOS,
Windows, and Linux.

:::

The exact software state that produced every number in this tutorial — full
`sessionInfo()` with all chapters' packages loaded, plus troubleshooting for two
version-compatibility problems you may hit — is recorded in
[`SESSIONINFO.md`](https://github.com/krferrier/Methylation-EWAS-tutorial/blob/main/tutorial/SESSIONINFO.md).

## Raw Methylation data files {#sec-idat-files}

Each sample on an Illumina array is stored as **two IDAT files** — one per color
channel:

```
GSM3853168_200932680028_R01C01_Grn.idat.gz   (green channel)
GSM3853168_200932680028_R01C01_Red.idat.gz   (red channel)
```

The filename encodes the physical location of the sample: `200932680028` is the
**Sentrix barcode** (the BeadChip / "slide"), and `R01C01` is the **array
position** (row 1, column 1) on that chip.

For this tutorial, we are going to use a subset of 96 samples from the Grady Trauma Project
(GSE132203), which is available from the NCBI GEO database. GEO stores each sample's IDAT
pair in its own per-sample supplementary FTP directory, e.g.:

```
https://ftp.ncbi.nlm.nih.gov/geo/samples/GSM3853nnn/GSM3853168/suppl/
```

Note the `GSM3853nnn` part: GEO buckets samples into directories of a thousand by
replacing the last three digits of the accession with `nnn`. `GSM3853168` therefore
lives under `GSM3853nnn`. The download scripts below construct this for you.

The 96-sample subset is **192 files, 1.42 GB** — two IDATs per sample. Do not download
these by hand from the GEO website; use one of the three options below.

## First, get the sample sheet {#sec-sample-sheet-get}

All three download options below read `data/sample_sheet.csv`, so you need that file
before you can fetch anything. It is a 96-row table, one row per array, that pairs each
GEO accession with the phenotype data for that sample. The download scripts use two of
its columns: `sample_id` (the `GSMnnnnnnn` accession) and `Basename` (the filename stem
of the array, without the `_Grn.idat.gz` / `_Red.idat.gz` suffix).

The sample sheet ships with this tutorial's repository, so there are two ways to get it.

::: {.panel-tabset}

#### Clone the whole repository

This gets you the sample sheet, the notebooks, and the helper scripts in one step, which
is what you want if you plan to work through the tutorial yourself. Run it in a
**terminal** — not in the R Console:

```{.bash filename="Terminal"}
git clone https://github.com/krferrier/Methylation-EWAS-tutorial.git
```

That creates a folder called `Methylation-EWAS-tutorial`, and the sample sheet lands at
`Methylation-EWAS-tutorial/tutorial/data/sample_sheet.csv`.

**Now point R at the `tutorial` folder inside it.** Every path in the chapters is written
relative to that folder, so this one step decides whether the rest of the tutorial works.
In RStudio: **Session → Set Working Directory → Choose Directory…**, then pick
`Methylation-EWAS-tutorial/tutorial`. Better still, open that folder as an RStudio Project
(**File → Open Project…**) and it is set for you every time you come back.

Check it before moving on:

```{.r filename="RStudio Console"}
getwd()                               # should end in .../Methylation-EWAS-tutorial/tutorial
file.exists("data/sample_sheet.csv")  # must print TRUE
```

#### Download just the one file

If you only want the sample sheet, download it straight from the repository:
[sample_sheet.csv](https://raw.githubusercontent.com/krferrier/Methylation-EWAS-tutorial/main/tutorial/data/sample_sheet.csv)
(in your browser: right-click → Save As).

Taking this route means **you have to build the folder structure yourself**, because the
chapters look for `data/sample_sheet.csv` — the file inside a folder called `data`, not
loose in your Downloads folder. Make a folder for the tutorial wherever you keep your
work, put a `data` folder inside it, and save the file in there. You can do that from the
Console:

```{.r filename="RStudio Console"}
dir.create("~/ewas-tutorial/data", recursive = TRUE)   # make the folders
setwd("~/ewas-tutorial")                               # work from the top one
download.file(
  "https://raw.githubusercontent.com/krferrier/Methylation-EWAS-tutorial/main/tutorial/data/sample_sheet.csv",
  "data/sample_sheet.csv")
file.exists("data/sample_sheet.csv")                   # must print TRUE
```

Note that this gets you the sample sheet only — not the notebooks or the helper scripts.
:::

Either way, check that you have 96 samples plus a header row before continuing.

### The layout every chapter assumes {#sec-layout}

No chapter uses a path like `C:/Users/you/Downloads/...`. They all use short relative
paths — `data/sample_sheet.csv`, `data/idats/` — which R resolves against your **working
directory**. So the working directory has to be the folder that contains `data`:

```
<your working directory>          <- getwd() prints this; "tutorial/" if you cloned
├── data/
│   ├── sample_sheet.csv          <- you have this now
│   └── idats/                    <- created for you by the download step below
├── 00_setup.qmd                  <- these only exist if you cloned
├── 01_qc.qmd
└── ...
```

Two lines confirm you are in the right place, and they are worth running before every
download step:

```{.r filename="RStudio Console"}
getwd()                               # where R currently is
file.exists("data/sample_sheet.csv")  # TRUE means the paths below will resolve
```

::: {.callout-warning}
## If you see `cannot open the connection`
This is the most common error in the whole tutorial, and it is worth recognising on sight:

```
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file 'data/sample_sheet.csv': No such file or directory
```

You may see that short path or a longer one naming a folder that does not exist — either
way the fix is the same. Nothing is wrong with the file and nothing is wrong with the
code: R is looking in the wrong folder, because your working directory is not the one that
contains `data`.

**Step 1 — find the folder the file is actually in.** Search for `sample_sheet.csv` (or for
the repository folder, `Methylation-EWAS-tutorial`) in Finder on macOS or File Explorer on
Windows. What you need is the folder *containing* `data`, not the file itself — if you
cloned the repository, that is the `tutorial` folder. Then copy that folder's path:

- **macOS** — click the folder, then press Cmd + Option + C. (Or right-click the folder,
  hold Option, and choose "Copy … as Pathname".)
- **Windows** — right-click the folder and choose "Copy as path".

**Step 2 — point R at it.** Paste the path into `setwd()`:

```{.r filename="RStudio Console"}
setwd("<paste the folder path here>")
```

On Windows the path you copied will use backslashes, which R reads as escape characters.
Swap them for forward slashes — `C:/Users/you/Methylation-EWAS-tutorial/tutorial` — or
double them. Or skip the typing altogether: **Session → Set Working Directory → Choose
Directory…** in RStudio does the same thing and cannot get the slashes wrong.

**Step 3 — check that it worked.**

```{.r filename="RStudio Console"}
getwd()                               # confirm where R thinks it is
file.exists("data/sample_sheet.csv")  # should print TRUE now
```

Once that last line prints `TRUE`, re-run whatever failed and it will work. For a fuller
walkthrough of working directories, see section 6 of
[Jacinta Kong's Setup R page](https://jacintak.github.io/teaching/introR/#working-directories-in-r).
:::


::: {.callout-note}
## The `Basename` column is a relative path
`Basename` reads `idats/GSM3853168_200932680028_R01C01` — a path relative to `data/`,
not an absolute one. That is why the download options below all write into
`data/idats/`, and why `read.metharray()` further down composes the path as
`file.path("data/idats", basename(ss$Basename))`. If you put the IDATs somewhere else,
adjust that path rather than editing the sample sheet.
:::

## Downloading the 96-sample subset {#sec-download}

::: {.callout-tip}
## Which option should I pick?
Start with **option 1**: it is R code you paste into the Console, it needs nothing
installed beyond the packages above, and it behaves the same on macOS, Windows, and Linux.
Option 2 pulls one archive from Zenodo instead of 192 separate files, so it is faster — but
it needs either a terminal or a manual download in your browser. Option 3 fetches the same
GEO files as option 1, from a terminal instead of the Console.

:::

### Option 1 — from GEO, using R (paste this into the Console)

This reads the sample sheet, builds each GEO URL, and downloads all 192 files.
It is safe to re-run: files already present are skipped, so an interrupted download
resumes where it stopped.

```{.r filename="RStudio Console"}
ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)
dir.create("data/idats", recursive = TRUE, showWarnings = FALSE)

# R's default download timeout is 60 seconds, which is not enough for a 7 MB
# file on a slow connection. Raise it or downloads will fail part-way through.
options(timeout = max(600, getOption("timeout")))

for (k in seq_len(nrow(ss))) {
  gsm    <- ss$sample_id[k]                       # e.g. GSM3853168
  stub   <- basename(ss$Basename[k])              # GSM3853168_200932680028_R01C01
  bucket <- paste0(substr(gsm, 1, nchar(gsm) - 3), "nnn")   # GSM3853nnn

  for (channel in c("Grn", "Red")) {
    fname <- sprintf("%s_%s.idat.gz", stub, channel)
    dest  <- file.path("data/idats", fname)
    if (file.exists(dest)) next                   # already have it
    url <- sprintf("https://ftp.ncbi.nlm.nih.gov/geo/samples/%s/%s/suppl/%s",
                   bucket, gsm, fname)
    # mode = "wb" is required on Windows: without it R corrupts binary files.
    download.file(url, dest, mode = "wb", quiet = TRUE)
  }
  cat(sprintf("[%3d/%d] %s\n", k, nrow(ss), gsm))
}

cat("files downloaded:", length(list.files("data/idats")), "of 192\n")
```

### Option 2 — one archive from Zenodo (one file instead of 192)

The subset is deposited as a single archive alongside the other tutorial data, so this is
one download and one extraction, with no per-sample loop and no GEO path construction. You can either 
access and download the archive from Zenodo manually from the
[tutorial data record](https://doi.org/10.5281/zenodo.22135215) — that link always
resolves to the newest version — or 
you can use the provided bash script `get_data.sh` as follows:

```{.bash filename="Terminal"}
./get_data.sh A_idats
```

If using the bash script, the 192 IDATs will be located in `data/idats/` relative to wherever the script was run. So, make 
sure to run the script from the location you want the dataset to live. If you are downloading the archive manually, I recommend 
following the same filepath structure such that the samples are all located in `data/idats/`, which is where the sample sheet's
`Basename` column expects them.

### Option 3 — from GEO, using the command line

::: {.panel-tabset}

#### macOS / Linux

`curl` is preinstalled on macOS. `-C -` resumes a partial file, so re-running is safe.

```{.bash filename="Terminal"}
mkdir -p data/idats

# Pull the accession (column 1) and Basename (column 11) out of the sample
# sheet, stripping the quotes the CSV uses, then fetch both channels for each.
awk -F, 'NR > 1 { gsub(/"/, "", $1); gsub(/"/, "", $11); print $1, $11 }' \
    data/sample_sheet.csv |
while read -r gsm basename_path; do
  stub="$(basename "$basename_path")"        # GSM3853168_200932680028_R01C01
  bucket="${gsm%???}nnn"                     # GSM3853nnn
  for channel in Grn Red; do
    f="${stub}_${channel}.idat.gz"
    [ -s "data/idats/$f" ] && continue       # already downloaded
    url="https://ftp.ncbi.nlm.nih.gov/geo/samples/${bucket}/${gsm}/suppl/${f}"
    if curl -C - -sfL -o "data/idats/$f" "$url"; then
      echo "ok    $f"
    else
      echo "FAIL  $f"
    fi
  done
done

ls data/idats/*.idat.gz | wc -l      # expect 192
```

#### Windows (PowerShell)

Open **PowerShell** (not Command Prompt) and run:

```{.powershell filename="PowerShell"}
New-Item -ItemType Directory -Force -Path data\idats | Out-Null
$ss = Import-Csv data\sample_sheet.csv

foreach ($row in $ss) {
    $gsm    = $row.sample_id
    $stub   = Split-Path $row.Basename -Leaf
    $bucket = $gsm.Substring(0, $gsm.Length - 3) + "nnn"

    foreach ($channel in @("Grn", "Red")) {
        $fname = "${stub}_${channel}.idat.gz"
        $dest  = "data\idats\$fname"
        if (Test-Path $dest) { continue }
        $url = "https://ftp.ncbi.nlm.nih.gov/geo/samples/$bucket/$gsm/suppl/$fname"
        Invoke-WebRequest -Uri $url -OutFile $dest
    }
    Write-Host "$($gsm) done"
}

(Get-ChildItem data\idats\*.idat.gz).Count    # expect 192
```

:::

::: {.callout-note}
## You do not need to unzip the IDATs
`minfi` reads gzipped IDATs (`.idat.gz`) directly. Leave them compressed.
:::

## Reading the IDATs into R {#sec-read-idats}

The sample sheet's `Basename` column holds the path stem of each array — everything
except the `_Grn.idat.gz` / `_Red.idat.gz` suffix. `read.metharray()` takes those stems
and pairs the two channels for you:

```{r}
#| label: read-idats
#| filename: "RStudio Console"
library(minfi)

# The sample sheet does double duty: it lists the files to read, and it holds
# the phenotype row for each array that we attach to the object below.
ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)

# Type the variables we intend to model, now, once. Level order matters: R uses
# the FIRST level as the reference category, so listing "Control" first is what
# makes chapter 06 report the effect of PTSD relative to controls. Left to
# itself, factor() sorts alphabetically, puts "Case" first, and silently
# reverses the sign of every effect estimate downstream.
ss$ptsd            <- factor(ss$ptsd,            levels = c("Control", "Case"))
ss$sex             <- factor(ss$sex,             levels = c("F", "M"))
ss$childhood_abuse <- factor(ss$childhood_abuse, levels = c("neg", "pos"))

# Read the 192 IDATs (two channels x 96 arrays) into one RGChannelSet.
RGset <- read.metharray(
  basenames = file.path("data/idats", basename(ss$Basename)),
  force     = TRUE          # see the callout below -- required for this dataset
)

# read.metharray() reads intensities and nothing else -- it does not carry the
# sample sheet along with it. Attach the phenotype table now so that every later
# chapter can recover it with pData(RGset) rather than re-reading the CSV and
# hoping the row order still matches. It does match here, because the files were
# read in sample-sheet order.
colnames(RGset) <- ss$sample_id                       # label arrays by GEO accession
pData(RGset)    <- DataFrame(ss, row.names = ss$sample_id)   # DataFrame() comes from
                                                             # S4Vectors, attached by minfi

RGset
dim(RGset)                  # 1051943 addresses x 96 arrays

# Save it. Chapters 01 and 02 both start from this object, and re-reading 192
# IDATs costs about two and a half minutes every time. The file is roughly
# 460 MB, so keep it on a disk you have room on.
saveRDS(RGset, "data/01_RGset.rds")
```

The result is an `RGChannelSet`: raw red and green intensities per address, with no
normalization and no probe filtering yet. Saved as `data/01_RGset.rds`, it is the input to
[Quality control](01_qc.qmd) and to [normalization](02_normalization.qmd) — both chapters
open by reading that exact file. It is also what the `B_qc` tier gives you if you skip this
step, so if reading the IDATs is impractical on your machine you can download the object
instead and carry on from chapter 01 unchanged.

::: {.callout-warning}
## `force = TRUE` is required here, and it is not a formality
Without it this dataset fails with:

```
[read.metharray] Trying to parse IDAT files with different array size
but seemingly all of the same type.
```

Ten of the twelve slides in this series were scanned with 1,052,641 addresses; the
remaining two (`201220980045` and `201228780138`, 16 samples) carry 1,051,943.
`read.metharray()` refuses to parse the different array sizes unless you set `force = TRUE`, which restricts the object to
the addresses common to every array. This is why `dim(RGset)` reports the smaller
number.

This is a scanner-decoding difference, not a data-quality problem, and taking the
intersection is the correct way to handle the situation. A probe missing from some arrays cannot be compared
across all of them anyway. But do not automatically use `force = TRUE` on other
data — if the array *types* genuinely differ (EPIC v1 mixed with v2, say), the same
error is telling you something you should not override. Check what the sizes actually
are before you force it:

```{r}
#| label: idat-sizes
#| filename: "RStudio Console"
library(illuminaio)
sizes <- sapply(file.path("data/idats", paste0(basename(ss$Basename), "_Grn.idat.gz")),
                function(f) nrow(readIDAT(f)$Quants))
table(sizes)
```
:::

## Computational cost: time, memory, and disk {#sec-cost}

Everything below was measured on the machine that produced this tutorial — 32 logical
cores (24 physical), 126 GB RAM, local NVMe — running the final filtered 86-sample cohort and 756,251 CpGs.
Your times will differ, but this should provide a general estimate to guide what sample
size you can analyse with your computing resources. The expensive steps are cell-type
deconvolution, normalization, and SVA; The EWAS itself actually uses less resources depending 
on how it's run.

### Benchmarks per chapter

Each computationally intensive step of the tutorial was benchmarked for runtime and max memory needed. The table below shows one row per step. Memory is the peak **total across every R
process**, which is what has to fit in RAM.

| Chapter | Step                                         | Wall time | Peak memory |
|---|----------------------------------------------|---:|---:|
| [00 Setup](00_setup.qmd#sec-read-idats) | Read 96 IDAT pairs (`read.metharray`)        | 2 min 21 s | 5.8 GB |
| [01 QC](01_qc.qmd) | Detection *p*-values (`detectionP`)          | 48 s | 3.8 GB |
| [02 Normalization](02_normalization.qmd) | Functional normalization (`preprocessFunnorm`) | 2 min 4 s | 12.5 GB |
| [03 Probe filtering](03_probe_filtering.qmd) | Mask + detection-*p* + sex chromosomes       | 26 s | 5.6 GB |
| [04 Cell composition](04_cell_composition.qmd) | Deconvolution (`estimateCellCounts2`, IDOL)  | 1 min 54 s | **15.0 GB** |
| [05 Batch effects](05_batch_effects.qmd) | Stratified `ComBat` on slide                 | 57 s | 4.1 GB |
| [05 Batch effects](05_batch_effects.qmd) | `num.sv` + `sva` at *k* = 6                  | 5 min 1 s | 8.9 GB |
| [06 EWAS](06_ewas.qmd) | EWAS using `lmFit` + `eBayes` + BACON        | 18 s | 4.1 GB |
| [Chapter 07](07_pipeline.qmd) | Non-stratified EWAS on 86 samples, 8 workers | 6 min 12 s | **12.0 GB** |
| [Chapter 07](07_pipeline.qmd) | Stratified EWAS: 2 strata (44 and 42) run simultaneously, 4 workers each | 11 min 37 s | 6.3 GB |
| [08 Annotation](08_annotation.qmd) | GENCODE v41 join over 756,251 CpGs           | 11 s | 1.3 GB |

On average, the whole tutorial should take you about 30 minutes of compute. The 
potential limiting resource is memory. The normalization and cell-type deconvolution takes about 12.5GB and 15GB of 
memory to run, respectively. Those using a laptop with 8GB of RAM may not be able to perform these steps on their 
local computer. Saved checkpoints of these results are available in the Zenodo record so that you can still run the rest of the 
steps in these chapters.

For the Chapter 07 EWAS run using a snakemake pipeline, the first row is the whole cohort in one
EWAS. The second splits it into two — here by sex, giving 44 and 42 samples — and runs
those two in parallel. Note that the split run needs *less* memory, not more: peak memory
is driven by **samples × workers**, and each stratum is running half the samples with half
the workers. The amount of memory required for the EWAS using snakemake can be reduced by
reducing the number of workers. The trade-off for reducing the number of workers is that
the wall time will increase. 

::: {.callout-warning}
## Time versus Memory tradeoffs
You may have noticed that the EWAS method shown in Chapter 6 runs significantly faster than the EWAS performed using 
snakemake in Chapter 07, and at this sample size it also uses less memory — 4.1 GB against 12.0 GB for the same 86 
samples at 8 workers. Those two numbers are not fixed in the same way, though. The pipeline's memory is a function of how 
many workers you give it, so it is a dial you control; the Chapter 06 approach holds the whole matrix in one process, so 
its memory is set by the data and grows with the cohort. 86 samples is quite small, and a full-scale EWAS will have 
several hundreds to thousands of samples. The faster method shown in Chapter 06 will very quickly require more memory resources 
than you have available as the number of samples you try to run at once increase. The Snakemake pipleine in Chapter 07 takes a bit longer,
especially for smaller datasets, because part of what it's doing under the hood is breaking the big dataset into smaller, more 
reasonably sized chunks. The Snakemake EWAS method was designed to minimize memory cost for full-scale analyses; the consequence of this 
is longer runtimes.
::: 

Disk space can also be a constraint. The input files require the largest amount of space, but some of the intermediate files produced 
can also be large. For the 86 sample subset, I would make sure you have at least 6 GB of free disk space. 

### Checking what resources your own machine has

Before you plan a run, find out what you are working with. 

#### Cores

We can use R to determine the number of cores available on your machine. This uses base R and should work for Windows, Mac, or Linux OS. 
```{r}
#| label: check-resources
#| filename: "RStudio Console"
# --- cores -----------------------------------------------------------------
parallel::detectCores()                      # logical cores, incl. hyperthreading
parallel::detectCores(logical = FALSE)       # physical cores
```

`detectCores()` reports *logical* cores. With hyperthreading that is often twice the
physical count, but do not count on the factor of two: Intel's hybrid designs (12th
generation and later) pair hyperthreaded performance cores with single-threaded
efficiency cores, so the machine these benchmarks were run on — a Core i9-14900K —
reports **32 logical against 24 physical**. Set `workers` from the physical number.
On Linux, `detectCores(logical = FALSE)` does not always manage this and can return the
logical count instead — as it did for the output above, which was rendered on a Linux
machine — so confirm the physical count with the terminal commands below.

If you would rather not open R, or want to confirm what R is reporting:

::: {.panel-tabset}

##### macOS

```{.bash filename="Terminal (macOS)"}
sysctl -n hw.logicalcpu      # logical cores, incl. hyperthreading
sysctl -n hw.physicalcpu     # physical cores -- use this for `workers`
```

##### Linux

```{.bash filename="Terminal (Linux)"}
nproc     # logical cores, incl. hyperthreading

# physical cores = "Core(s) per socket" x "Socket(s)"
lscpu | awk -F: '/Core\(s\) per socket/{c=$2} /Socket\(s\)/{s=$2} END{print c*s}'
```

Recent `lscpu` versions indent their output under group headings, so anchoring a
pattern to the start of the line (`grep '^Core(s) per socket'`) silently matches
nothing. The `awk` above does not anchor, and works either way.

##### Windows

- Press Ctrl + Shift + Esc to open Task Manager.
- Click on the Performance tab (the speedometer icon on the left or top).
- Select CPU in the left sidebar.
- Read **Cores** and **Logical processors** in the lower-right block of stats. Set
  `workers` from **Cores**.

:::

#### RAM

You can check how much RAM is available to use with Task Manager on Windows, Activity Monitor on Mac, and System Monitor on Linux.

::: {.panel-tabset}

##### macOS

- Press Cmd + Space to open Spotlight, type Activity Monitor, and press Enter.
- Click on the Memory tab near the top of the window.
- Look at the bottom of the window at Memory Pressure and Physical Memory usage graphs to see how much RAM is free or used.

From a terminal:

```{.bash filename="Terminal (macOS)"}
sysctl -n hw.memsize | awk '{ printf "%.1f GB installed\n", $1/1024/1024/1024 }'
vm_stat                      # live breakdown of free / active / wired pages
```

##### Linux

- Open your applications menu and search for System Monitor (or KSysGuard / Plasma System Monitor on KDE desktops).
- Click on the Resources tab (or Performance tab depending on your distribution).
- Look at the Memory section to see a live graph and numbers showing used and free RAM.

From a terminal:

```{.bash filename="Terminal (Linux)"}
free -h                      # 'total' is installed, 'available' is what you can use
```

##### Windows

- Press Ctrl + Shift + Esc to open Task Manager.
- Click on the Performance tab (the speedometer icon on the left or top).
- Select Memory in the left sidebar.
- View your In use, Available, and Free RAM amounts at the bottom of the screen.

From PowerShell:

```{.powershell filename="PowerShell"}
Get-CimInstance Win32_ComputerSystem |
  Select-Object @{n='InstalledGB';e={[math]::Round($_.TotalPhysicalMemory/1GB,1)}}
```

:::

::: {.callout-note}
## "Available" is the number that matters
Installed RAM is the headline figure, but the operating system and everything else
you have open are already using some of it. Compare the peak-memory column in the
table above against **available** memory, not installed.
:::

#### Disk Space

You can check free disk space using a graphical user interface (GUI) by opening your system's built-in file manager or storage settings panel on Windows, Mac, and Linux. 

::: {.panel-tabset}

##### macOS

- Click the Apple logo in the top-left corner of your screen.
- Select System Settings (or System Preferences on older versions).
- Click General in the sidebar, then select Storage.
- View a colorful breakdown of your used and available disk space.

From a terminal, checking the disk holding the current folder:

```{.bash filename="Terminal (macOS)"}
df -h .                      # look at the 'Avail' column
```

##### Linux

- Open your default file manager (such as Nautilus in GNOME or Dolphin in KDE).
- Click on Other Locations or look at the left sidebar under Devices or Hard Disk to see available storage.

From a terminal, checking the disk holding the current folder:

```{.bash filename="Terminal (Linux)"}
df -h .                      # look at the 'Avail' column
```

##### Windows

- Open File Explorer (press Windows Key + E).
- Click on This PC in the left sidebar.
- Look at the Devices and drives section to see a visual bar and remaining free gigabytes for your drives (like the C: drive).
- Alternatively, open the Settings app, go to System, and select Storage for a detailed breakdown.

From PowerShell:

```{.powershell filename="PowerShell"}
Get-PSDrive -PSProvider FileSystem |
  Select-Object Name, @{n='FreeGB';e={[math]::Round($_.Free/1GB,1)}}
```

:::

## Optional: the full 795-sample dataset {#sec-full-series}

If you want to try a full-scale model after finishing the tutorial, GEO also serves the
entire series as one tar archive. It is **11.9 GB** compressed and expands to 1,590 IDAT
files, so budget disk space and time accordingly — and expect the EWAS itself to take longer than the 86-sample walkthrough.

::: {.panel-tabset}

#### macOS / Linux command line

```{.bash filename="Terminal"}
mkdir -p data/idats_full && cd data/idats_full
curl -C - -O "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE132nnn/GSE132203/suppl/GSE132203_RAW.tar"
tar -xvf GSE132203_RAW.tar
ls *.idat.gz | wc -l      # expect 1590
```

#### Windows (PowerShell)

```{.powershell filename="PowerShell"}
New-Item -ItemType Directory -Force -Path data\idats_full | Out-Null
Set-Location data\idats_full
$u = "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE132nnn/GSE132203/suppl/GSE132203_RAW.tar"
Invoke-WebRequest -Uri $u -OutFile GSE132203_RAW.tar
tar -xvf GSE132203_RAW.tar      # tar ships with Windows 10 and later
(Get-ChildItem *.idat.gz).Count
```

#### Web browser
You can also manually download the full sample set by selecting the link for the `GSE132203_RAW.tar` from the Supplementary Files 
section at the bottom of the [GEO Accession page for the Grady Trauma Project](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE132203).

:::

The archive also contains the platform manifest and a `filelist.txt` inventory.

### Building a sample sheet for the full series

The 96-sample sheet was built for you, but for the full series you need to make your own.
The phenotype data lives in the **series matrix file**, which `GEOquery` downloads and
parses for you — you do not need to touch the tar archive to get it.

Two details make this less tidy than it sounds, and both are shown in the code below.
First, GEO packs the phenotype into columns named `characteristics_ch1`,
`characteristics_ch1.1`, and so on, each holding a `"label: value"` *string* rather than a
typed value; you have to split them and recover the label. Second, the Basename is not a
column at all — it has to be recovered from the supplementary-file URL.

```{r}
#| label: full-sample-sheet
#| filename: "RStudio Console"
library(GEOquery)

# The series matrix is ~30 MB; the default 60-second timeout is not enough.
options(timeout = max(1200, getOption("timeout")))

# getGPL = FALSE skips the platform manifest, which we do not need here
# and which is large.
gse <- getGEO("GSE132203", GSEMatrix = TRUE, getGPL = FALSE)[[1]]
pd  <- Biobase::pData(gse)
nrow(pd)                     # 795 samples

# Every "characteristics_ch1*" column holds a "label: value" string. Split on the
# first colon, using the label GEO supplies as the column name.
ch <- grep("^characteristics_ch1", names(pd), value = TRUE)
pheno <- lapply(ch, function(k) {
  v     <- as.character(pd[[k]])
  label <- sub(":.*$", "", v[1])
  value <- trimws(sub("^[^:]*:", "", v))
  setNames(data.frame(value, stringsAsFactors = FALSE), make.names(label))
})
pheno <- do.call(cbind, pheno)

# The Grn supplementary URL ends in "<GSM>_<slide>_<pos>_Grn.idat.gz". Strip that
# suffix to recover the Basename stem minfi wants.
stem <- sub("_Grn\\.idat\\.gz$", "", basename(as.character(pd$supplementary_file)))

full <- data.frame(
  sample_id  = pd$geo_accession,
  series     = "GSE132203",
  sentrix_id = pd$title,                       # e.g. 200928190033_R01C01
  slide      = sub("_.*$", "", pd$title),
  array_pos  = sub("^.*_", "", pd$title),
  Basename   = file.path("idats_full", stem),  # matches the download path above
  stringsAsFactors = FALSE
)
full <- cbind(full, pheno)

# Everything out of the series matrix is text, including the numbers.
for (k in c("age", "mergedcapsandpsswinthin30days",
            "childabphyssexemot_ctq_01modandsev", "pc1", "pc2", "pc3")) {
  full[[k]] <- suppressWarnings(as.numeric(full[[k]]))
}

# Derive the two variables the tutorial models. PTSD is the CAPS/PSS indicator;
# 50 samples have it missing and will drop out of any model that uses it.
full$ptsd <- factor(ifelse(full$mergedcapsandpsswinthin30days == 1, "Case", "Control"),
                    levels = c("Control", "Case"))
full$sex  <- ifelse(full$gender == "Female", "F", "M")

write.csv(full, "data/sample_sheet_full.csv", row.names = FALSE)
table(full$ptsd, full$sex, useNA = "ifany")
```

Running that gives 795 rows and 24 columns, with this breakdown:

|         | Female | Male |
|---------|---:|---:|
| Control | 375 | 182 |
| Case    | 160 | 28  |
| missing | 36  | 14  |

Two things to notice before you model any of it. **PTSD status is more imbalanced
than in the teaching subset** — 188 cases against 557 controls, and among males only 28
cases — whereas the 96-sample subset was balanced on sex and enriched for cases (32
cases against 55 controls).
And **50 samples have no PTSD value at all**, so they will be excluded from any model
using that variable, exactly as 9 of the 96 do in this tutorial.


```{r}
#| label: verify-full-sheet
#| filename: "RStudio Console"
ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)
i  <- match(ss$sample_id, full$sample_id)
sum(!is.na(i))                                   # 96
identical(basename(ss$Basename), basename(full$Basename[i]))   # TRUE
# The three genetic ancestry PCs in the 96-sample sheet were taken from this table.
all.equal(ss[, c("pc1", "pc2", "pc3")], full[i, c("pc1", "pc2", "pc3")],
          check.attributes = FALSE)                                  # TRUE
```

Note that `Basename` above points at `idats_full/`, matching the download commands in
this section. If you put the full series somewhere else, change that one line rather
than editing the sheet afterwards.

## The sample sheet {#sec-sample-sheet}

`minfi` reads IDATs given a **Basename** (the path up to `_Grn/_Red`). We pair
each Basename with the phenotype row for that GSM. The last three columns, `pc1`–`pc3`,
are the genetic ancestry principal components GEO deposits for each sample; one of the 96
has none. [Genetic ancestry PCs](04b_genetic_pcs.qmd) explains what they are and
why the models use them. Here are the first 8 rows of the sample sheet:

```{r}
#| label: sample-sheet
# The same CSV you downloaded above -- this is the file the whole tutorial reads.
ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)
ss <- ss[order(ss$slide, ss$array_pos), ]   # chip, then position within the chip
cat("Samples:", nrow(ss), " | Chips:", length(unique(ss$slide)), "\n")
knitr::kable(head(ss[, c("sample_id","slide","array_pos","gender","age",
                         "ptsd","race")], 8),
             row.names = FALSE,
             caption = "First 8 rows of the sample sheet")
```

## What the next notebook does {#sec-next}

With the IDATs read and saved as `data/01_RGset.rds`, the [QC notebook](01_qc.qmd)
loads that `RGChannelSet` and runs the quality checks that decide which samples are
trustworthy before we begin normalization.
