---
title: "Setup: getting the data onto your machine"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

This notebook introduces the dataset the tutorial uses, shows how to get the raw
files onto your own machine, and sets out what running everything costs in time,
memory, and disk. It assumes the array background from
[Background](00a_background.qmd).

## Building the software environment {#sec-env}

Everything except the pipeline chapter needs **only R**. If you are new to computing,
start there and ignore conda entirely — you will still be able to work through
preprocessing, QC, cell composition, batch effects, the EWAS itself, and annotation.

::: {.panel-tabset}

### R only (recommended if you are getting started)

You need **R 4.2.x** ([download from CRAN](https://cran.r-project.org/)) and, if you want, [RStudio](https://posit.co/download/rstudio-desktop/). Everything below is typed
into the **Console** pane — the one with the `>` prompt. 

The tutorial ships an installer as
[`install_packages.R`](https://github.com/krferrier/Methylation-EWAS-tutorial/blob/main/tutorial/install_packages.R),
which installs the four CRAN and eighteen Bioconductor packages the chapters use. The
simplest way to run it is to fetch and run it in one line, without downloading anything
yourself:

```r
source("https://raw.githubusercontent.com/krferrier/Methylation-EWAS-tutorial/main/tutorial/install_packages.R")
```

If you cloned the repository (see "First, get the sample sheet" below), you can run the
local copy instead. In RStudio, set the working directory to `tutorial/` first — either
**Session → Set Working Directory → Choose Directory…**, or open `tutorial/` as an
RStudio Project so it is set for you every time:

```r
source("install_packages.R")
```

And if you do prefer a command line, the same script runs there:

```bash
Rscript install_packages.R
```

If you would rather not run a script you have not read — a reasonable instinct — the
whole installation is short enough to paste directly into the Console:

```r
repos <- "https://cloud.r-project.org"
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager", repos = repos)
if (!requireNamespace("remotes",     quietly = TRUE)) install.packages("remotes",     repos = repos)

# matrixStats must be pinned below 1.2.0, and installed BEFORE the rest
if (!requireNamespace("matrixStats", quietly = TRUE) ||
    packageVersion("matrixStats") >= "1.2.0") {
  remotes::install_version("matrixStats", version = "1.0.0",
                           repos = repos, upgrade = "never")
}

BiocManager::install(c(
  "data.table", "ggplot2", "knitr", "DT",
  "minfi", "illuminaio", "wateRmelon", "GEOquery", "Biobase",
  "IlluminaHumanMethylationEPICmanifest",
  "IlluminaHumanMethylationEPICanno.ilm10b4.hg19",
  "FlowSorted.Blood.EPIC", "genefilter",
  "sva", "limma", "bacon",
  "sesame", "sesameData", "GenomicRanges", "rtracklayer",
  "missMethyl", "methylGSA"
), ask = FALSE, update = FALSE, upgrade = "never")
```

`BiocManager` maps R 4.2 to Bioconductor 3.16 automatically — the release that produced
every number in this tutorial. Budget 20-40 minutes the first time — on Windows and
macOS most packages arrive as pre-built binaries, but there are a lot of them; on Linux
they build from source, which is slower.

::: {.callout-note}
## The pin order is not incidental
`matrixStats` goes in first, on its own, and `upgrade = "never"` is on every call.
`wateRmelon` and `sva` both depend on `matrixStats`, so installing them first would pull
the current release and break chapters 01-02 with `useNames = NA is defunct`. The
installer script does the same thing in the same order, and re-checks the pin when it
finishes.
:::

::: {.callout-tip}
## If `matrixStats` fails to build, you have an alternative
On Windows and macOS, CRAN and Bioconductor ship **pre-built binaries** for R 4.2, so
almost nothing here compiles. The one exception is the pinned `matrixStats`: CRAN's
R-4.2 binary is version 1.3.0, and 1.0.0 is only available as source, so that single
package needs a compiler — [Rtools](https://cran.r-project.org/bin/windows/Rtools/) on
Windows, or `xcode-select --install` on macOS.

If you would rather not install a compiler, you can use the 1.3.0 binary and disable
the check that breaks, by putting this at the top of your session — before loading
`minfi`:

```r
install.packages("matrixStats")                      # gets the 1.3.0 binary
options(matrixStats.useNames.NA = "deprecated")      # downgrades the error to a warning
```

That hidden option exists in `matrixStats` 1.3.0 and turns the fatal
`useNames = NA is defunct` back into a warning. It was **removed in 1.5.0**, so this
works only if you stay on 1.3.0. Setting it in `.Rprofile` saves retyping it each
session.
:::

The R version matters more than it looks. R 4.2 is what selects Bioconductor 3.16; on a
newer R you would get a newer Bioconductor, and some results could shift.

The non-R sections: **chapter 07** runs the EWAS through a Snakemake pipeline, and that
genuinely needs conda. Chapter 06 already runs the same association test directly in R,
so nothing analytical is lost — chapter 07 shows how to scale it up. The `comb-p`
region-level section of chapter 08 also needs conda; the rest of that chapter does not.

### conda

If you already use conda, or you want to run chapter 07, three environments are defined in the
repository's `envs/` directory. Chapter 07 needs Snakemake, and the region-level analysis
in chapter 08 needs `comb-p`, which runs on Python 2.7 and cannot share an environment
with Snakemake's Python 3:

```bash
conda env create -f envs/methyl.yml     # chapters 01-06, 08 -- R 4.2.3
conda env create -f envs/smk.yml        # chapter 07         -- Snakemake
conda env create -f envs/combp.yml      # chapter 08 DMRs    -- comb-p
```

| Environment | Chapters | Core tool |
|---|---|---|
| `ewas-methyl` | 01-06, 08 | R 4.2.3, Bioconductor 3.16 |
| `ewas-smk` | 07 | Snakemake 9.26.1 |
| `ewas-combp` | 08 (DMR section) | comb-p on Python 2.7 |

If you only want the R chapters, `envs/methyl.yml` is the only one you need.

Each file lists only the packages the tutorial uses directly and lets conda resolve the
rest. `envs/smk.yml` deliberately holds nothing but the workflow engine: the pipeline
declares its own per-rule R environment and Snakemake builds it on first run, which is
also where METAL is compiled from source.

If you do not have conda, [Miniforge](https://github.com/conda-forge/miniforge) is a
minimal installer that defaults to the conda-forge channel; its README covers macOS,
Windows, and Linux.

:::

The exact software state that produced every number in this tutorial — full
`sessionInfo()` with all chapters' packages loaded, plus troubleshooting for two
version-compatibility problems you may hit — is recorded in
[`SESSIONINFO.md`](https://github.com/krferrier/Methylation-EWAS-tutorial/blob/main/tutorial/SESSIONINFO.md).

## Raw Methylation data files {#sec-idat-files}

Each sample on an Illumina array is stored as **two IDAT files** — one per color
channel:

```
GSM3853168_200932680028_R01C01_Grn.idat.gz   (green channel)
GSM3853168_200932680028_R01C01_Red.idat.gz   (red channel)
```

The filename encodes the physical location of the sample: `200932680028` is the
**Sentrix barcode** (the BeadChip / "slide"), and `R01C01` is the **array
position** (row 1, column 1) on that chip.

For this tutorial, we are going to use a subset of 96 samples from the Grady Trauma Project
(GSE132203), which is available from the NCBI GEO database. GEO stores each sample's IDAT
pair in its own per-sample supplementary FTP directory, e.g.:

```
https://ftp.ncbi.nlm.nih.gov/geo/samples/GSM3853nnn/GSM3853168/suppl/
```

Note the `GSM3853nnn` part: GEO buckets samples into directories of a thousand by
replacing the last three digits of the accession with `nnn`. `GSM3853168` therefore
lives under `GSM3853nnn`. The download scripts below construct this for you.

The 96-sample subset is **192 files, 1.42 GB** — two IDATs per sample. Do not download
these by hand from the GEO website; use one of the three options below.

## First, get the sample sheet {#sec-sample-sheet-get}

All three download options below read `data/sample_sheet.csv`, so you need that file
before you can fetch anything. It is a 96-row table, one row per array, that pairs each
GEO accession with the phenotype data for that sample. The download scripts use two of
its columns: `sample_id` (the `GSMnnnnnnn` accession) and `Basename` (the filename stem
of the array, without the `_Grn.idat.gz` / `_Red.idat.gz` suffix).

The sample sheet ships with this tutorial's repository, so there are two ways to get it.

::: {.panel-tabset}

#### Clone the whole repository

This gets you the sample sheet, the notebooks, and the helper scripts in one step, which
is what you want if you plan to work through the tutorial yourself:

```bash
git clone https://github.com/krferrier/Methylation-EWAS-tutorial.git
```

The sample sheet is then at `Methylation-EWAS-tutorial/tutorial/data/sample_sheet.csv`, and the download commands below
will work as written from that `tutorial/` directory.

#### Download just the one file

If you only want the sample sheet, you can download it directly from the GitHub repository [here](https://github.com/krferrier/Methylation-EWAS-tutorial/blob/e43eea48807773eb6e48a3ebd8ffe5d7e1d084ab/tutorial/data/sample_sheet.csv).

:::

Either way, check that you have 96 samples plus a header row before continuing.

::: {.callout-note}
## The `Basename` column is a relative path
`Basename` reads `idats/GSM3853168_200932680028_R01C01` — a path relative to `data/`,
not an absolute one. That is why the download options below all write into
`data/idats/`, and why `read.metharray()` further down composes the path as
`file.path("data/idats", basename(ss$Basename))`. If you put the IDATs somewhere else,
adjust that path rather than editing the sample sheet.
:::

## Downloading the 96-sample subset {#sec-download}

::: {.callout-tip}
## Which option should I pick?
The 96 sample subset is provided as an archive you'll need to download and unpack in Option 1. Option 2 is how to fetch the 
files from the NCBI GEO ftp site programmatically using R. Option 3 is how to download the files programmatically using the 
command line.

:::

### Option 1 — one archive from Zenodo (simplest)

The subset is deposited as a single archive alongside the other tutorial data, so this is
one download and one extraction, with no per-sample loop and no GEO path construction. You can either 
access and download the archive from Zenodo manually from the
[tutorial data record](https://doi.org/10.5281/zenodo.22135215) — that link always
resolves to the newest version — or 
you can use the provided bash script `get_data.sh` as follows:

```bash
./get_data.sh A_idats
```

If using the bash script, the 192 IDATs will be located in `data/idats/` relative to wherever the script was run. So, make 
sure to run the script from the location you want the dataset to live. If you are downloading the archive manually, I recommend 
following the same filepath structure such that the samples are all located in `data/idats/`, which is where the sample sheet's
`Basename` column expects them.

### Option 2 — from GEO, using R (works on macOS, Windows, and Linux)

This reads the sample sheet, builds each GEO URL, and downloads all 192 files.
It is safe to re-run: files already present are skipped, so an interrupted download
resumes where it stopped.

```r
ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)
dir.create("data/idats", recursive = TRUE, showWarnings = FALSE)

# R's default download timeout is 60 seconds, which is not enough for a 7 MB
# file on a slow connection. Raise it or downloads will fail part-way through.
options(timeout = max(600, getOption("timeout")))

for (k in seq_len(nrow(ss))) {
  gsm    <- ss$sample_id[k]                       # e.g. GSM3853168
  stub   <- basename(ss$Basename[k])              # GSM3853168_200932680028_R01C01
  bucket <- paste0(substr(gsm, 1, nchar(gsm) - 3), "nnn")   # GSM3853nnn

  for (channel in c("Grn", "Red")) {
    fname <- sprintf("%s_%s.idat.gz", stub, channel)
    dest  <- file.path("data/idats", fname)
    if (file.exists(dest)) next                   # already have it
    url <- sprintf("https://ftp.ncbi.nlm.nih.gov/geo/samples/%s/%s/suppl/%s",
                   bucket, gsm, fname)
    # mode = "wb" is required on Windows: without it R corrupts binary files.
    download.file(url, dest, mode = "wb", quiet = TRUE)
  }
  cat(sprintf("[%3d/%d] %s\n", k, nrow(ss), gsm))
}

cat("files downloaded:", length(list.files("data/idats")), "of 192\n")
```

### Option 3 — from GEO, using the command line

::: {.panel-tabset}

#### macOS / Linux

`curl` is preinstalled on macOS. `-C -` resumes a partial file, so re-running is safe.

```bash
mkdir -p data/idats

# Pull the accession (column 1) and Basename (column 11) out of the sample
# sheet, stripping the quotes the CSV uses, then fetch both channels for each.
awk -F, 'NR > 1 { gsub(/"/, "", $1); gsub(/"/, "", $11); print $1, $11 }' \
    data/sample_sheet.csv |
while read -r gsm basename_path; do
  stub="$(basename "$basename_path")"        # GSM3853168_200932680028_R01C01
  bucket="${gsm%???}nnn"                     # GSM3853nnn
  for channel in Grn Red; do
    f="${stub}_${channel}.idat.gz"
    [ -s "data/idats/$f" ] && continue       # already downloaded
    url="https://ftp.ncbi.nlm.nih.gov/geo/samples/${bucket}/${gsm}/suppl/${f}"
    if curl -C - -sfL -o "data/idats/$f" "$url"; then
      echo "ok    $f"
    else
      echo "FAIL  $f"
    fi
  done
done

ls data/idats/*.idat.gz | wc -l      # expect 192
```

#### Windows (PowerShell)

Open **PowerShell** (not Command Prompt) and run:

```powershell
New-Item -ItemType Directory -Force -Path data\idats | Out-Null
$ss = Import-Csv data\sample_sheet.csv

foreach ($row in $ss) {
    $gsm    = $row.sample_id
    $stub   = Split-Path $row.Basename -Leaf
    $bucket = $gsm.Substring(0, $gsm.Length - 3) + "nnn"

    foreach ($channel in @("Grn", "Red")) {
        $fname = "${stub}_${channel}.idat.gz"
        $dest  = "data\idats\$fname"
        if (Test-Path $dest) { continue }
        $url = "https://ftp.ncbi.nlm.nih.gov/geo/samples/$bucket/$gsm/suppl/$fname"
        Invoke-WebRequest -Uri $url -OutFile $dest
    }
    Write-Host "$($gsm) done"
}

(Get-ChildItem data\idats\*.idat.gz).Count    # expect 192
```

:::

::: {.callout-note}
## You do not need to unzip the IDATs
`minfi` reads gzipped IDATs (`.idat.gz`) directly. Leave them compressed.
:::

## Reading the IDATs into R {#sec-read-idats}

The sample sheet's `Basename` column holds the path stem of each array — everything
except the `_Grn.idat.gz` / `_Red.idat.gz` suffix. `read.metharray()` takes those stems
and pairs the two channels for you:

```{r}
#| label: read-idats
#| eval: false
library(minfi)

ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)

RGset <- read.metharray(
  basenames = file.path("data/idats", basename(ss$Basename)),
  force     = TRUE          # see the callout below -- required for this dataset
)

RGset
dim(RGset)                  # 1051943 addresses x 96 arrays
```

The result is an `RGChannelSet`: raw red and green intensities per address, with no
normalization and no probe filtering yet. That object is the input to
[Quality control](01_qc.qmd), and it is what the `B_qc` tier gives you if you skip
this step.

::: {.callout-warning}
## `force = TRUE` is required here, and it is not a formality
Without it this dataset fails with:

```
[read.metharray] Trying to parse IDAT files with different array size
but seemingly all of the same type.
```

Ten of the twelve slides in this series were scanned with 1,052,641 addresses; the
remaining two (`201220980045` and `201228780138`, 16 samples) carry 1,051,943.
`read.metharray()` refuses to parse the different array sizes unless you set `force = TRUE`, which restricts the object to
the addresses common to every array. This is why `dim(RGset)` reports the smaller
number.

This is a scanner-decoding difference, not a data-quality problem, and taking the
intersection is the correct way to handle the situation. A probe missing from some arrays cannot be compared
across all of them anyway. But do not automatically use `force = TRUE` on other
data — if the array *types* genuinely differ (EPIC v1 mixed with v2, say), the same
error is telling you something you should not override. Check what the sizes actually
are before you force it:

```{r}
#| label: idat-sizes
#| eval: false
library(illuminaio)
sizes <- sapply(file.path("data/idats", paste0(basename(ss$Basename), "_Grn.idat.gz")),
                function(f) nrow(readIDAT(f)$Quants))
table(sizes)
```
:::

## Computational cost: time, memory, and disk {#sec-cost}

Everything below was measured on the machine that produced this tutorial — 32 cores,
126 GB RAM, local NVMe — running the final filtered 87-sample cohort and 756,251 CpGs.
Your times will differ, but this should provide a general estimate to guide what sample
size you can analyse with your computing resources. The expensive steps are cell-type
deconvolution, normalization, and SVA; The EWAS itself actually uses less resources depending 
on how it's run.

### Benchmarks per chapter

Each computationally intensive step of the tutorial was benchmarked for runtime and max memory needed. The table below shows one row per step. Memory is the peak **total across every R
process**, which is what has to fit in RAM.

| Chapter | Step                                         | Wall time | Peak memory |
|---|----------------------------------------------|---:|---:|
| [01 QC](01_qc.qmd) | Read 96 IDAT pairs (`read.metharray`)        | 2 min 21 s | 5.8 GB |
| [01 QC](01_qc.qmd) | Detection *p*-values (`detectionP`)          | 48 s | 3.8 GB |
| [02 Normalization](02_normalization.qmd) | Functional normalization (`preprocessFunnorm`) | 2 min 4 s | 12.5 GB |
| [03 Probe filtering](03_probe_filtering.qmd) | Mask + detection-*p* + sex chromosomes       | 26 s | 5.6 GB |
| [04 Cell composition](04_cell_composition.qmd) | Deconvolution (`estimateCellCounts2`, IDOL)  | 1 min 54 s | **15.0 GB** |
| [05 Batch effects](05_batch_effects.qmd) | Stratified `ComBat` on slide                 | 57 s | 4.1 GB |
| [05 Batch effects](05_batch_effects.qmd) | `num.sv` + `sva` at *k* = 6                  | 5 min 1 s | 8.9 GB |
| [06 EWAS](06_ewas.qmd) | EWAS using `lmFit` + `eBayes` + BACON        | 18 s | 4.1 GB |
| [Chapter 07](07_pipeline.qmd) | Non-stratified EWAS on 87 samples, 8 workers | 6 min 12 s | **12.0 GB** |
| [Chapter 07](07_pipeline.qmd) | Stratified EWAS: 2 strata (45 and 42) run simultaneously, 4 workers each | 11 min 37 s | 6.3 GB |
| [08 Annotation](08_annotation.qmd) | GENCODE v41 join over 756,251 CpGs           | 11 s | 1.3 GB |

On average, the whole tutorial should take you about 30 minutes of compute. The 
potential limiting resource is memory. The normalization and cell-type deconvolution takes about 12.5GB and 15GB of 
memory to run, respectively. Those using a laptop with 8GB of RAM may not be able to perform these steps on their 
local computer. Saved checkpoints of these results are available in the Zenodo record so that you can still run the rest of the 
steps in these chapters.

For the Chapter 07 EWAS run using a snakemake pipeline, the first row is the whole cohort in one
EWAS. The second splits it into two — here by sex, giving 45 and 42 samples — and runs
those two in parallel. Note that the split run needs *less* memory, not more: peak memory
is driven by **samples × workers**, and each stratum is running half the samples with half
the workers. The amount of memory required for the EWAS using snakemake can be reduced by
reducing the number of workers. The trade-off for reducing the number of workers is that
the wall time will increase. 

::: {.callout-warning}
## Time versus Memory tradeoffs
You may have noticed that the EWAS method shown in Chapter 6 runs significantly faster than the EWAS performed using 
snakemake in Chapter 07, and at this sample size it also uses less memory — 4.1 GB against 12.0 GB for the same 87 
samples at 8 workers. Those two numbers are not fixed in the same way, though. The pipeline's memory is a function of how 
many workers you give it, so it is a dial you control; the Chapter 06 approach holds the whole matrix in one process, so 
its memory is set by the data and grows with the cohort. 87 samples is quite small, and a full-scale EWAS will have 
several hundreds to thousands of samples. The faster method shown in Chapter 06 will very quickly require more memory resources 
than you have available as the number of samples you try to run at once increase. The Snakemake pipleine in Chapter 07 takes a bit longer,
especially for smaller datasets, because part of what it's doing under the hood is breaking the big dataset into smaller, more 
reasonably sized chunks. The Snakemake EWAS method was designed to minimize memory cost for full-scale analyses; the consequence of this 
is longer runtimes.
::: 

Disk space can also be a constraint. The input files require the largest amount of space, but some of the intermediate files produced 
can also be large. For the 87 sample subset, I would make sure you have at least 6BG of free disk space. 

### Checking what resources your own machine has

Before you plan a run, find out what you are working with. 

#### Cores
We can use R to determine the number of cores available on your machine. This uses base R and should work for Windows, Mac, or Linux OS. 
```{r}
#| label: check-resources
#| eval: false
# --- cores -----------------------------------------------------------------
parallel::detectCores()                      # logical cores, incl. hyperthreading
parallel::detectCores(logical = FALSE)       # physical cores
```

`detectCores()` reports *logical* cores, which on most laptops is twice the physical
count. Set `workers` from the physical number.

#### RAM
You can check how much RAM is availabe to use with Task Manager on Windows, Activity Monitor on Mac, and System Monitor on Linux.

- **Windows**:
  - Press Ctrl + Shift + Esc to open Task Manager.
  - Click on the Performance tab (the speedometer icon on the left or top).
  - Select Memory in the left sidebar.
  - View your In use, Available, and Free RAM amounts at the bottom of the screen.
- **Mac**:
  - Press Cmd + Space to open Spotlight, type Activity Monitor, and press Enter.
  - Click on the Memory tab near the top of the window.
  - Look at the bottom of the window at Memory Pressure and Physical Memory usage graphs to see how much RAM is free or used.
- **Linux**: 
  - Open your applications menu and search for System Monitor (or KSysGuard / Plasma System Monitor on KDE desktops).
  - Click on the Resources tab (or Performance tab depending on your distribution).
  - Look at the Memory section to see a live graph and numbers showing used and free RAM. 


#### Disk Space
You can check free disk space using a graphical user interface (GUI) by opening your system's built-in file manager or storage settings panel on Windows, Mac, and Linux. 

- **Windows**:
  - Open File Explorer (press Windows Key + E).
  - Click on This PC in the left sidebar.
  - Look at the Devices and drives section to see a visual bar and remaining free gigabytes for your drives (like the C: drive).
  - Alternatively, open the Settings app, go to System, and select Storage for a detailed breakdown.
- **Mac**:
  - Click the Apple logo in the top-left corner of your screen.
  - Select System Settings (or System Preferences on older versions).
  - Click General in the sidebar, then select Storage.
  - View a colorful breakdown of your used and available disk space.
- **Linux**:
  - Open your default file manager (such as Nautilus in GNOME or Dolphin in KDE).
  - Click on Other Locations or look at the left sidebar under Devices or Hard Disk to see available storage.

## Optional: the full 795-sample dataset {#sec-full-series}

If you want to try a full-scale model after finishing the tutorial, GEO also serves the
entire series as one tar archive. It is **11.9 GB** compressed and expands to 1,590 IDAT
files, so budget disk space and time accordingly — and expect the EWAS itself to take longer than the 87-sample walkthrough.

::: {.panel-tabset}

#### macOS / Linux command line

```bash
mkdir -p data/idats_full && cd data/idats_full
curl -C - -O "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE132nnn/GSE132203/suppl/GSE132203_RAW.tar"
tar -xvf GSE132203_RAW.tar
ls *.idat.gz | wc -l      # expect 1590
```

#### Windows (PowerShell)

```powershell
New-Item -ItemType Directory -Force -Path data\idats_full | Out-Null
Set-Location data\idats_full
$u = "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE132nnn/GSE132203/suppl/GSE132203_RAW.tar"
Invoke-WebRequest -Uri $u -OutFile GSE132203_RAW.tar
tar -xvf GSE132203_RAW.tar      # tar ships with Windows 10 and later
(Get-ChildItem *.idat.gz).Count
```

#### Web browser
You can also manually download the full sample set by selecting the link for the `GSE132203_RAW.tar` from the Supplementary Files 
section at the bottom of the [GEO Accession page for the Grady Trauma Project](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE132203).

:::

The archive also contains the platform manifest and a `filelist.txt` inventory.

### Building a sample sheet for the full series

The 96-sample sheet was built for you, but for the full series you need to make your own.
The phenotype data lives in the **series matrix file**, which `GEOquery` downloads and
parses for you — you do not need to touch the tar archive to get it.

Two details make this less tidy than it sounds, and both are shown in the code below.
First, GEO packs the phenotype into columns named `characteristics_ch1`,
`characteristics_ch1.1`, and so on, each holding a `"label: value"` *string* rather than a
typed value; you have to split them and recover the label. Second, the Basename is not a
column at all — it has to be recovered from the supplementary-file URL.

```{r}
#| label: full-sample-sheet
#| eval: false
library(GEOquery)

# The series matrix is ~30 MB; the default 60-second timeout is not enough.
options(timeout = max(1200, getOption("timeout")))

# getGPL = FALSE skips the platform manifest, which we do not need here
# and which is large.
gse <- getGEO("GSE132203", GSEMatrix = TRUE, getGPL = FALSE)[[1]]
pd  <- Biobase::pData(gse)
nrow(pd)                     # 795 samples

# Every "characteristics_ch1*" column holds a "label: value" string. Split on the
# first colon, using the label GEO supplies as the column name.
ch <- grep("^characteristics_ch1", names(pd), value = TRUE)
pheno <- lapply(ch, function(k) {
  v     <- as.character(pd[[k]])
  label <- sub(":.*$", "", v[1])
  value <- trimws(sub("^[^:]*:", "", v))
  setNames(data.frame(value, stringsAsFactors = FALSE), make.names(label))
})
pheno <- do.call(cbind, pheno)

# The Grn supplementary URL ends in "<GSM>_<slide>_<pos>_Grn.idat.gz". Strip that
# suffix to recover the Basename stem minfi wants.
stem <- sub("_Grn\\.idat\\.gz$", "", basename(as.character(pd$supplementary_file)))

full <- data.frame(
  sample_id  = pd$geo_accession,
  series     = "GSE132203",
  sentrix_id = pd$title,                       # e.g. 200928190033_R01C01
  slide      = sub("_.*$", "", pd$title),
  array_pos  = sub("^.*_", "", pd$title),
  Basename   = file.path("idats_full", stem),  # matches the download path above
  stringsAsFactors = FALSE
)
full <- cbind(full, pheno)

# Everything out of the series matrix is text, including the numbers.
for (k in c("age", "mergedcapsandpsswinthin30days",
            "childabphyssexemot_ctq_01modandsev")) {
  full[[k]] <- suppressWarnings(as.numeric(full[[k]]))
}

# Derive the two variables the tutorial models. PTSD is the CAPS/PSS indicator;
# 50 samples have it missing and will drop out of any model that uses it.
full$ptsd <- factor(ifelse(full$mergedcapsandpsswinthin30days == 1, "Case", "Control"),
                    levels = c("Control", "Case"))
full$sex  <- ifelse(full$gender == "Female", "F", "M")

write.csv(full, "data/sample_sheet_full.csv", row.names = FALSE)
table(full$ptsd, full$sex, useNA = "ifany")
```

Running that gives 795 rows and 24 columns, with this breakdown:

|         | Female | Male |
|---------|---:|---:|
| Control | 375 | 182 |
| Case    | 160 | 28  |
| missing | 36  | 14  |

Three things to notice before you model any of it. **PTSD status is more imbalanced
than in the teaching subset** — 188 cases against 557 controls, and among males only 28
cases — where the 96-sample subset was deliberately balanced on case/control and sex.
And **50 samples have no PTSD value at all**, so they will be excluded from any model
using that variable, exactly as 9 of the 96 do in this tutorial.


```{r}
#| label: verify-full-sheet
#| eval: false
ss <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)
i  <- match(ss$sample_id, full$sample_id)
sum(!is.na(i))                                   # 96
identical(basename(ss$Basename), basename(full$Basename[i]))   # TRUE
```

Note that `Basename` above points at `idats_full/`, matching the download commands in
this section. If you put the full series somewhere else, change that one line rather
than editing the sheet afterwards.

## The sample sheet {#sec-sample-sheet}

`minfi` reads IDATs given a **Basename** (the path up to `_Grn/_Red`). We pair
each Basename with the phenotype row for that GSM. Here is the first 8 rows of the sample sheet:

```{r}
#| label: sample-sheet
ss <- readRDS("data/sample_sheet.rds")
ss <- ss[order(ss$slide, ss$array_pos), ]
cat("Samples:", nrow(ss), " | Chips:", length(unique(ss$slide)), "\n")
knitr::kable(head(ss[, c("sample_id","slide","array_pos","gender","age",
                         "ptsd","race")], 8),
             row.names = FALSE,
             caption = "First 8 rows of the sample sheet")
```

## What the next notebook does {#sec-next}

With the sample sheet provided, the [QC notebook](01_qc.qmd) reads all 96 IDAT
pairs into a single `RGChannelSet` and runs the quality checks that decide which
samples are trustworthy before we begin normalization.
