---
title: "Quality control: which samples can we trust?"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

The first step in methylation analsysis is quality control. A failed
array, a swapped sample, or a mislabeled sex can cause serious trouble 
downstream if it slips through. This notebook runs the standard `minfi`
[@aryee2014minfi] QC 
process on the 96-sample subset and decides what (if anything) to drop.

We start from the raw two-channel object saved by the [Setup](00_setup.qmd) step.

```{r}
#| label: load
rg  <- readRDS(data_path("01_RGset.rds"))
qcp <- readRDS(data_path("01_qc_pieces.rds"))
ss  <- as.data.frame(pData(rg))
rg
```

## 1. Detection p-values — did the probe beat background?

Every probe is compared to the array's negative-control distribution. A large
**detection p-value** means the signal is indistinguishable from background; the
measurement is noise. We summarize per sample: the mean detection p, and the
fraction of probes failing at p > 0.01.

```{r}
#| label: detp-plot
#| fig-cap: "Per-sample detection p-value. All 96 samples are far below the 0.01 concern line — no failed arrays."
det <- data.frame(sample = names(qcp$det_mean),
                  mean_detP = qcp$det_mean,
                  frac_fail = qcp$det_fail,
                  slide = ss$slide)
ggplot(det, aes(x = reorder(sample, mean_detP), y = mean_detP)) +
  geom_point(color = "#4C72B0", size = 1.6) +
  geom_hline(yintercept = 0.01, linetype = "dashed", color = "#C44E52") +
  labs(x = "sample (ranked)", y = "mean detection p-value",
       title = "Detection p-value per sample") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r}
#| label: detp-table
cat("worst sample mean detP:", signif(max(det$mean_detP), 3), "\n")
cat("worst sample % probes p>0.01:", signif(max(det$frac_fail) * 100, 3), "%\n")
cat("samples failing QC (mean detP > 0.01):", sum(det$mean_detP > 0.01), "\n")
```

::: {.callout-note}
## What a failure would look like
On a real plate you will sometimes see one or two samples an order of magnitude
above the rest, crossing the 0.01 line. Those get dropped here, before
normalization. 
:::

## 2. Signal intensity — is the array bright enough?

`getQC()` reduces each sample to the median methylated (`mMed`) and unmethylated
(`uMed`) log2 intensity. Weak arrays fall toward the bottom-left; `minfi`'s
convention flags samples below a line at `(mMed + uMed)/2 = 10.5`.

```{r}
#| label: intensity-plot
#| fig-cap: "Median methylated vs unmethylated log2 intensity. All samples sit well above the 10.5 low-signal cutoff."
qc <- as.data.frame(qcp$qc); qc$sample <- rownames(qc)
qc$flag <- ((qc$mMed + qc$uMed) / 2) < 10.5
ggplot(qc, aes(mMed, uMed, color = flag)) +
  geom_point(size = 2) +
  geom_abline(intercept = 2 * 10.5, slope = -1, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c(`FALSE` = "#4C72B0", `TRUE` = "#C44E52"),
                     labels = c("pass", "low signal"), name = NULL) +
  labs(x = "median methylated log2 int (mMed)",
       y = "median unmethylated log2 int (uMed)",
       title = "Signal-intensity QC") +
  coord_equal(xlim = c(9, 14), ylim = c(9, 14))
```

```{r}
#| label: intensity-summary
cat("mMed range:", round(min(qc$mMed),2), "-", round(max(qc$mMed),2), "\n")
cat("low-signal flagged samples:", sum(qc$flag), "\n")
```

## 3. Predicted vs reported sex 

Sex is the single most reliable phenotype to predict from methylation: the X and
Y chromosome signal is unambiguous. `getSex()` compares median total intensity on
X vs Y. A mismatch between predicted and reported sex is a strong signal of a
**sample swap or annotation error** — and, because sex is a strong methylation
covariate we must model, catching it here is important.

```{r}
#| label: sex-plot
#| fig-cap: "Predicted sex from X/Y intensity, colored by reported sex. Clean separation, zero mismatches."
sx <- data.frame(xMed = qcp$psex$xMed, yMed = qcp$psex$yMed,
                 predicted = qcp$psex$predictedSex,
                 reported = qcp$rep_sex)
sx$mismatch <- sx$predicted != sx$reported
ggplot(sx, aes(xMed, yMed, color = reported, shape = mismatch)) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(F = "#DD8452", M = "#4C72B0"), name = "reported") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4), guide = "none") +
  labs(x = "median X log2 int", y = "median Y log2 int",
       title = "Predicted vs reported sex")
```

```{r}
#| label: sex-table
tab <- table(reported = sx$reported, predicted = sx$predicted)
knitr::kable(tab, caption = "Reported vs predicted sex")
cat("sex mismatches:", sum(sx$mismatch), "of", nrow(sx), "\n")
```

## 4. SNP-identity fingerprint

EPIC carries ~59 `rs` SNP probes that measure genotype, not methylation. Their
beta values cluster near 0, 0.5, or 1 (the three genotypes). They are useful for
two things: spotting **sample duplicates / swaps** (two arrays from the same
person share a fingerprint) and, in longitudinal designs, confirming that repeat
samples really are the same individual.

```{r}
#| label: snp-heatmap
#| fig-cap: "The ~59 rs SNP probes across samples. Discrete banding near 0 / 0.5 / 1 is the expected genotype signal; no two columns are identical, so no duplicate samples."
snps <- qcp$snps
# cluster samples by SNP fingerprint
d <- dist(t(snps))
ord <- hclust(d)$order
image(t(snps[, ord]), col = colorRampPalette(c("#4C72B0","white","#C44E52"))(64),
      axes = FALSE, main = "rs SNP-probe fingerprints (samples clustered)")
box()
```

```{r}
#| label: snp-dups
# nearest-neighbor distance: a near-zero value would flag a duplicate
dm <- as.matrix(dist(t(snps)))
diag(dm) <- NA
nn <- apply(dm, 1, min, na.rm = TRUE)
cat("SNP probes:", nrow(snps), "\n")
cat("smallest between-sample fingerprint distance:", round(min(nn), 3),
    "(a value near 0 would indicate a duplicate)\n")
```

## Verdict

```{r}
#| label: verdict
drop_detp <- det$sample[det$mean_detP > 0.01]
drop_int  <- qc$sample[qc$flag]
drop_sex  <- sx$reported[sx$mismatch]
keep <- setdiff(colnames(rg), unique(c(drop_detp, drop_int)))
cat("Samples in:", ncol(rg), "\n")
cat("Dropped for detection p:", length(drop_detp), "\n")
cat("Dropped for low intensity:", length(drop_int), "\n")
cat("Sex mismatches (investigate, don't auto-drop):", length(drop_sex), "\n")
cat("Samples passing QC:", length(keep), "\n")
saveRDS(keep, data_path("01_qc_pass_samples.rds"))
```

All 96 samples pass. We carry the full set forward to
[normalization](02_normalization.qmd).

::: {.callout-tip}
## Always do QC. ALWAYS
Even though all samples passed QC thresholds in this subset, doing the QC was 
still important as now we have evidence supporting that these samples are 'clean'.
In a real dataset there are typically several samples that will not pass
QC for one of the 4 checks.
:::

## What the phenotype file is missing

QC checks whether the *arrays* worked. It is also the right moment to check whether
the **phenotype data** you will eventually model is complete, because a covariate you
discover is absent at modeling time is far more expensive than one you plan around
now.

For blood methylation the most consequential gap is usually **smoking**. Smoking
produces some of the largest and most reproducible methylation signals in blood
[@joehanes2016smoking], so an unmeasured smoking variable is unmeasured confounding
in every model you fit. The Grady deposit gives PTSD, age, sex, race, and childhood
trauma — but no smoking variable, which is common in public deposits rather than
unusual.

That absence is worth noting here because it shapes decisions in the next two
chapters: it is one reason to consider excluding literature-derived smoking CpGs at
[probe filtering](03_probe_filtering.qmd), and the reason the surrogate variables
estimated in [batch effects](05_batch_effects.qmd) are doing more work than
technical-noise removal alone. The full set of options, and the one this tutorial
takes, is laid out in [the EWAS chapter](06_ewas.qmd).

::: {.callout-tip}
## Inventory your covariates before you normalize
Write down the model you intend to fit, then check each term against the metadata you
actually have. Note anything missing, and decide *then* how you will handle it.
:::

## References {.unnumbered}

::: {#refs}
:::
