---
title: "Quality control: which samples can we trust?"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

The first step in methylation analysis is quality control. A failed
array, a swapped sample, or sex-mismatch can cause serious trouble
downstream if it slips through. This notebook runs the standard `minfi`
[@aryee2014minfi] QC 
process on the 96-sample subset and walks you through how to decide what (if anything) to drop.

We start from the raw two-channel object saved by the [Setup](00_setup.qmd) step.

```{r}
#| label: load
rg  <- readRDS(data_path("01_RGset.rds"))
qcp <- readRDS(data_path("01_qc_pieces.rds"))
ss  <- as.data.frame(pData(rg))
rg
```

## 1. Detection p-values — did the probe beat background?

Every probe is compared to the array's negative-control distribution. A large
**detection p-value** means the signal is indistinguishable from background; the
measurement is noise. We summarize per sample: the mean detection p, and the
fraction of probes failing at p > 0.01.

```{r}
#| label: detp-plot
#| fig-cap: "Per-sample detection p-value. All 96 samples are far below the 0.01 concern line — no failed arrays."
det <- data.frame(sample = names(qcp$det_mean),
                  mean_detP = qcp$det_mean,
                  frac_fail = qcp$det_fail,
                  slide = ss$slide)
ggplot(det, aes(x = reorder(sample, mean_detP), y = mean_detP)) +
  geom_point(color = ewas_col[["teal"]], size = 1.6) +
  geom_hline(yintercept = 0.01, linetype = "dashed", color = ewas_col[["plum"]]) +
  labs(x = "sample (ranked)", y = "mean detection p-value",
       title = "Detection p-value per sample") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r}
#| label: detp-table
cat("worst sample mean detP:", signif(max(det$mean_detP), 3), "\n")
cat("worst sample % probes p>0.01:", signif(max(det$frac_fail) * 100, 3), "%\n")
cat("samples failing QC (mean detP > 0.01):", sum(det$mean_detP > 0.01), "\n")
```

::: {.callout-note}
## What a failure would look like
On a real plate you will sometimes see one or two samples an order of magnitude
above the rest, crossing the 0.01 line. Those get dropped here, before
normalization. 
:::

## 2. Signal intensity — is the array bright enough?

`getQC()` reduces each sample to the median methylated (`mMed`) and unmethylated
(`uMed`) log2 intensity. Weak arrays fall toward the bottom-left; `minfi`'s
convention flags samples below a line at `(mMed + uMed)/2 = 10.5`.

```{r}
#| label: intensity-plot
#| fig-cap: "Median methylated vs unmethylated log2 intensity. All samples sit well above the 10.5 low-signal cutoff."
qc <- as.data.frame(qcp$qc); qc$sample <- rownames(qc)
qc$flag <- ((qc$mMed + qc$uMed) / 2) < 10.5
ggplot(qc, aes(mMed, uMed, color = flag)) +
  geom_point(size = 2) +
  geom_abline(intercept = 2 * 10.5, slope = -1, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c(`FALSE` = ewas_col[["teal"]], `TRUE` = ewas_col[["plum"]]),
                     labels = c("pass", "low signal"), name = NULL) +
  labs(x = "median methylated log2 int (mMed)",
       y = "median unmethylated log2 int (uMed)",
       title = "Signal-intensity QC") +
  coord_equal(xlim = c(9, 14), ylim = c(9, 14))
```

```{r}
#| label: intensity-summary
cat("mMed range:", round(min(qc$mMed),2), "-", round(max(qc$mMed),2), "\n")
cat("low-signal flagged samples:", sum(qc$flag), "\n")
```

## 3. Predicted vs reported sex 

Sex is the single most reliable phenotype to predict from methylation: the X and
Y chromosome signal is unambiguous. `getSex()` compares median total intensity on
X vs Y. A mismatch between predicted and reported sex is a strong signal of a
**sample swap or annotation error** — and, because sex is a strong methylation
covariate we must model, catching it here is important.

```{r}
#| label: sex-plot
#| fig-cap: "Predicted sex from X/Y intensity, colored by reported sex. Clean separation, zero mismatches."
sx <- data.frame(xMed = qcp$psex$xMed, yMed = qcp$psex$yMed,
                 predicted = qcp$psex$predictedSex,
                 reported = qcp$rep_sex)
sx$mismatch <- sx$predicted != sx$reported
ggplot(sx, aes(xMed, yMed, color = reported, shape = mismatch)) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(F = ewas_col[["sand"]], M = ewas_col[["teal"]]), name = "reported") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4), guide = "none") +
  labs(x = "median X log2 int", y = "median Y log2 int",
       title = "Predicted vs reported sex")
```

```{r}
#| label: sex-table
tab <- table(reported = sx$reported, predicted = sx$predicted)
knitr::kable(tab, caption = "Reported vs predicted sex")
cat("sex mismatches:", sum(sx$mismatch), "of", nrow(sx), "\n")
```

## 4. SNP-identity fingerprint

EPIC carries ~59 `rs` SNP probes that measure genotype, not methylation. Their
beta values cluster near 0, 0.5, or 1 (the three genotypes). They are useful for
two things: spotting **sample duplicates / swaps** (two arrays from the same
person share a fingerprint) and, in longitudinal designs, confirming that repeat
samples really are the same individual.

```{r}
#| label: snp-heatmap
#| fig-cap: "The ~59 rs SNP probes across samples. Discrete banding near 0 / 0.5 / 1 is the expected genotype signal; no two columns are identical, so no duplicate samples."
snps <- qcp$snps
# cluster samples by SNP fingerprint
d <- dist(t(snps))
ord <- hclust(d)$order
image(t(snps[, ord]), col = colorRampPalette(c(ewas_col[["teal"]],"white",ewas_col[["plum"]]))(64),
      axes = FALSE, main = "rs SNP-probe fingerprints (samples clustered)")
box()
```

```{r}
#| label: snp-dups
# nearest-neighbor distance: a near-zero value would flag a duplicate
dm <- as.matrix(dist(t(snps)))
diag(dm) <- NA
nn <- apply(dm, 1, min, na.rm = TRUE)
cat("SNP probes:", nrow(snps), "\n")
cat("smallest between-sample fingerprint distance:", round(min(nn), 3),
    "(a value near 0 would indicate a duplicate)\n")
```

## Verdict

```{r}
#| label: verdict
drop_detp <- det$sample[det$mean_detP > 0.01]
drop_int  <- qc$sample[qc$flag]
drop_sex  <- sx$reported[sx$mismatch]
keep <- setdiff(colnames(rg), unique(c(drop_detp, drop_int)))
cat("Samples in:", ncol(rg), "\n")
cat("Dropped for detection p:", length(drop_detp), "\n")
cat("Dropped for low intensity:", length(drop_int), "\n")
cat("Sex mismatches (investigate, don't auto-drop):", length(drop_sex), "\n")
cat("Samples passing QC:", length(keep), "\n")
saveRDS(keep, data_path("01_qc_pass_samples.rds"))
```

All 96 samples pass. We carry the full set forward to
[normalization](02_normalization.qmd).

Note that "96 samples pass QC" is not the same as "96 samples are analyzed". Nine of
them have no recorded PTSD status and will drop out when the exposure enters the
model — see [phenotype completeness](#what-the-phenotype-file-is-missing) below.

::: {.callout-tip}
## Always do QC. ALWAYS
Even though all samples passed QC thresholds in this subset, doing the QC was 
still important as now we have evidence supporting that these samples are 'clean'.
In a real dataset there are typically several samples that will not pass
QC for one of the 4 checks.
:::

## What the phenotype file is missing

QC checks whether the *arrays* worked. It is also the right moment to check whether
the **phenotype data** you will eventually model is complete, because a covariate you
discover is absent at modeling time is far more expensive than one you plan around
now.

There are two distinct kinds of gap, and they have different consequences. A variable
that is **absent for every sample** is unmeasured confounding — you cannot adjust for
it. A variable that is **present but incomplete** costs you samples, because any model
containing that term silently drops the rows where it is missing. Check for both.

### Missing values in the variables we intend to model

Tabulate missingness for every term in the planned model, before you normalize:

```{r}
#| label: pheno-missing
model_vars <- c("ptsd", "sex", "age", "childhood_abuse",
                "mergedcapsandpsswinthin30days")
miss <- data.frame(
  variable = model_vars,
  n_missing = sapply(model_vars, function(v)
    sum(is.na(ss[[v]]) | ss[[v]] == "")),
  row.names = NULL)
knitr::kable(miss, caption = "Missing values per candidate model variable, across all 96 QC-passing arrays.")
```

PTSD status — the exposure — is missing for
`r sum(is.na(ss$ptsd) | ss$ptsd == "")` of the 96 samples. The continuous CAPS/PSS
severity score is missing for the same samples, so there is no alternative outcome
to fall back on for them.

::: {.callout-important}
## These nine samples are dropped later, and it is not a QC failure

Every one of the 96 arrays passed QC. But an EWAS model contains the exposure, so
`lm` / `limma` will drop any sample with no PTSD status — the analysis in
[chapter 06](06_ewas.qmd) and the pipeline run in
[chapter 07](07_pipeline.qmd) therefore fit **n = 87**, not 96.

This tutorial keeps all 96 samples through normalization
([chapter 02](02_normalization.qmd)), probe filtering
([chapter 03](03_probe_filtering.qmd)), cell-composition estimation
([chapter 04](04_cell_composition.qmd)), and batch diagnosis
([chapter 05](05_batch_effects.qmd)) on purpose. None of those steps uses PTSD
status: deconvolution and PCA describe the arrays, not the association model, so
there is no reason to discard usable arrays before the step that actually needs the
exposure. The count changes exactly once, at the point the exposure enters.

The distinction worth internalising: **samples dropped for quality** are samples you
do not trust, and **samples dropped for missing phenotype** are samples you trust
but cannot model. Report them separately. Collapsing them into one "n = 87 after QC"
sentence — the most common way this gets written up — hides the fact that nine
usable arrays were lost to an incomplete metadata table, which is a very different
problem from nine failed arrays.
:::

Because the nine are a loss from missingness rather than quality, it is worth a
moment to ask whether they are **missing at random** or missing in a way that
correlates with something. If the samples without PTSD status were, say, all male or
all from one chip, dropping them would distort the design:

```{r}
#| label: missing-pattern
no_ptsd <- is.na(ss$ptsd) | ss$ptsd == ""
cat("samples with no PTSD status:", sum(no_ptsd), "\n")
cat("sex:", paste(names(table(ss$sex[no_ptsd])),
                 table(ss$sex[no_ptsd]), collapse = " / "), "\n")
cat("age range:", paste(range(ss$age[no_ptsd]), collapse = "-"), "\n")
cat("distinct chips:", length(unique(ss$slide[no_ptsd])),
    "of", length(unique(ss$slide)), "\n")
```

Four female and five male, ages spanning the full cohort range, spread across eight
of the twelve chips — the missingness looks incidental rather than structured. That
is reassuring, but note that it is a *check you can only do for observed variables*;
it cannot rule out the samples differing on something unrecorded.

### Variables absent for every sample

For blood methylation the most consequential gap is usually **smoking**. Smoking
produces some of the largest and most reproducible methylation signals in blood
[@joehanes2016smoking], so an unmeasured smoking variable is unmeasured confounding
in every model you fit. The Grady deposit gives PTSD, age, sex, race, and childhood
trauma — but no smoking variable, which is common in public deposits rather than
unusual.

That absence is worth noting here because it shapes decisions in the next two
chapters. The full set of options, and the one this tutorial
takes, is laid out in [the EWAS chapter](06_ewas.qmd).

::: {.callout-tip}
## Inventory your covariates before you normalize
Write down the model you intend to fit, then check each term against the metadata you
actually have. For each term ask two questions: *is it present at all?* and *how many
samples is it missing for?* The first tells you what you cannot adjust for; the second
tells you your real analysis n. Note both, and decide *then* how you will handle them —
including whether a term is worth the samples it costs.
:::

## References {.unnumbered}

::: {#refs}
:::
