---
title: "Normalization: making arrays comparable"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

The next step in methylation data pre-processing is normalization. There is variation 
in intensities from the different probe chemistries, as well as from technical variation 
in the assay, which needs to be accounted for before the probe intensities can be compared 
across samples in downstream analyses. Normalization is the method used to do this. This 
notebook walks through the standard normalization strategies used for methylation data.

Since all of the samples from our small example subset passed QC, we will start normaliztion 
from the raw two-channel object produced in [QC](01_qc.qmd). If you did identify samples that 
failed to meet QC thresholds, those samples should be removed prior to normalization.

::: {.callout-note}
## What this chapter needs
This chapter starts from `data/01_RGset.rds`, written by [chapter 00](00_setup.qmd). If you
did not run that chapter, fetch the published checkpoint instead:

```{.bash filename="Terminal"}
./get_data.sh B_qc
```
:::

```{r}
#| label: load
library(minfi)                                        # preprocessRaw/Noob/Funnorm(), getBeta()
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19) # EPIC probe annotation: design type, coordinates
library(ggplot2)                                       # every figure in this chapter

# The RGChannelSet written by the Setup chapter: raw red and green intensities
# for 1,051,943 addresses across 96 arrays, with no normalization applied. This
# is the same object the QC chapter worked from -- all 96 samples passed, so
# nothing has been dropped from it.
rg <- readRDS("data/01_RGset.rds")

rg
```

## 1. Two sources of unwanted variation

**Within-array, technical.** Every Infinium measurement is a fluorescence
intensity, and fluorescence has a background floor (autofluorescence, non-specific
hybridization, optical noise) and a **dye bias**: the Cy3 (green) and Cy5 (red)
channels are not excited or detected with equal efficiency, so the same amount of
methylated and unmethylated DNA does not produce the same raw signal. Left
uncorrected, this shifts β-values systematically.

**Within-array, by probe design.** The EPIC array (like the 450K before it) mixes two 
assay chemistries [@pidsley2016epic]:

- **Type I** probes use two bead types per CpG (one for methylated, one for
  unmethylated), both read in the *same* color channel. They tolerate a wider
  dynamic range but use up two bead addresses per site.
- **Type II** probes use a single bead per CpG, with methylated and unmethylated
  states read in *different* channels. They are twice as space-efficient — which
  is why **84%** of EPIC probes are Type II (723,722 of 865,859; the comparable
  450K figure is 72%) — but their β-value distribution is
  **compressed toward the center**: they under-call fully methylated and fully
  unmethylated sites relative to Type I.

We can see this compression directly in the subset. Below are the median β and the
between-probe spread (SD of per-probe mean β) for each design, on raw data:

```{r}
#| label: probe-design
# Which design each CpG uses comes from the EPIC annotation package, not from the
# data: the `Type` column is "I" or "II" for each of the 865,859 annotated probes.
# These are the probes that carry hg19 coordinates, which is also what the genome
# mapping inside preprocessFunnorm needs -- so sampling from this set guarantees
# the same probes exist in all three normalized matrices built below.
ann    <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
typeI  <- rownames(ann)[ann$Type == "I"]
typeII <- rownames(ann)[ann$Type == "II"]

# preprocessRaw() only re-arranges the two raw channels into methylated and
# unmethylated signal -- no correction of any kind is applied, which is exactly
# what we want as the "before" picture. It takes a few seconds.
mset_raw     <- preprocessRaw(rg)
beta_raw_all <- getBeta(mset_raw)

# The figures and tables in this chapter describe the shape of the β
# distribution, and 20,000 probes describe that shape as well as 865,859 do at a
# thirtieth of the memory (a full matrix is ~665 MB; three of them will not fit
# comfortably alongside normalization). The seed is fixed so the sample -- and
# therefore every number in the tables below -- is the same on every render.
set.seed(1)
probe_sample <- sample(intersect(rownames(ann), rownames(beta_raw_all)), 20000)

beta_raw <- beta_raw_all[probe_sample, ]

# Drop the full-size objects as soon as the subset exists; the peak memory of
# this chapter is set by the normalization step below, not by this one.
rm(beta_raw_all, mset_raw)
invisible(gc())

dim(beta_raw)
```

```{r}
#| label: type-bias-table
# One summary row per probe design. `spread_sd` is the SD of the per-probe mean β
# across samples, i.e. how much the probes of a design differ from one another --
# not within-probe noise, which is why rowMeans() comes first.
type_summ <- function(beta, typeI, typeII) {
  bi  <- beta[rownames(beta) %in% typeI, , drop = FALSE]
  bii <- beta[rownames(beta) %in% typeII, , drop = FALSE]
  data.frame(
    design      = c("Type I", "Type II"),
    n_probes    = c(nrow(bi), nrow(bii)),
    median_beta = round(c(median(bi, na.rm = TRUE), median(bii, na.rm = TRUE)), 3),
    spread_sd   = round(c(sd(rowMeans(bi, na.rm = TRUE)),
                          sd(rowMeans(bii, na.rm = TRUE))), 3)
  )
}
knitr::kable(type_summ(beta_raw, typeI, typeII),
             caption = "Raw β by probe design (20k-probe sample). Type II sits closer to 0.5 with a narrower spread — the design-bias normalization must correct.")
```

::: {.callout-important}
## Why this matters for an EWAS
If Type I and Type II probes are on different scales, a differential-methylation
test will partly reflect *which chemistry* a probe uses rather than biology, and
effect sizes are not comparable across probe types.
:::

## 2. Normalization methods

Methods fall into two groups by *what* they try to make comparable.

### Within-array methods (fix probe-design bias per sample)

- **noob** (normal-exponential out-of-band background correction) models the
  background as a normal-exponential mixture using the out-of-band Type I signals,
  then applies a dye-bias correction [@triche2013noob]. It is a background/dye
  step, not a design-bias step, but it is the standard first move and is what most
  pipelines build on.
- **BMIQ** (beta-mixture quantile) fits a three-state beta mixture
  (unmethylated / hemimethylated / methylated) separately to Type I and Type II
  probes, then transforms the Type II distribution onto the Type I scale
  [@teschendorff2013bmiq]. It directly targets the design bias.
- **SWAN** (subset-quantile within-array normalization) matches Type I and Type II
  intensity distributions using a subset of probes with matched underlying biology
  [@maksimovic2012swan].

### Between-array methods (make samples comparable to each other)

- **Quantile normalization** forces every sample's intensity distribution to a
  common reference. Powerful, but it assumes the *global* methylation distribution
  is the same across samples — a poor assumption when groups differ biologically
  (e.g. tumor vs normal), where it can erase real signal.
- **Functional normalization (funnorm)** is a between-array method designed for
  exactly that case [@fortin2014funnorm]. Instead of forcing distributions to
  match, it regresses out variation explained by the array's **internal control
  probes** — bisulfite conversion, staining, extension, hybridization controls —
  which capture technical variation *without* being tied to biological signal. It
  is the recommended default when you expect large biological differences between
  groups, and it wraps noob as its first step.

::: {.callout-note}
## Choosing a method
There is no single winner. A defensible default for blood EWAS is **noob +
functional normalization** (what `preprocessFunnorm` does in one call), because it
corrects background/dye and removes technical variation without assuming samples
share a methylation distribution. If your study is a within-tissue comparison with
small expected differences, quantile-based methods (or `preprocessQuantile`) are
also reasonable.
:::

## 3. Running it on the subset

We compute three versions and compare: **raw** (`preprocessRaw`, no correction),
**noob** (`preprocessNoob`), and **functional** (`preprocessFunnorm`). The
functional result is a `GenomicRatioSet` — CpGs mapped to the genome with β and
M-values ready for analysis.

The code to produce these subsets is shown below. It may take several minutes to 
complete. 

```{r}
#| label: normalize-noob
# noob on its own: background correction from the out-of-band Type I signals plus
# a dye-bias correction, one array at a time. Roughly a minute and several GB for
# 96 arrays. Subset to the same 20,000 probes immediately and drop the full
# MethylSet, so the funnorm step below starts from a clean slate.
mset_noob <- preprocessNoob(rg)
beta_noob <- getBeta(mset_noob)[probe_sample, ]

rm(mset_noob)
invisible(gc())
```

```{r}
#| label: normalize-funnorm
# This is the expensive step in the chapter: about 2 minutes and a peak of
# ~12.5 GB at 96 arrays, because funnorm runs noob first and then fits the
# control-probe PCA across the full 865,859-probe matrix. If your machine cannot
# hold that, take the resume path in the callout below.
grs_fun <- preprocessFunnorm(rg)

# `grs_fun` is the handoff object for the rest of the tutorial -- keep it whole.
# Only the β matrix gets subset, for the figures in this chapter.
beta_fun <- getBeta(grs_fun)[probe_sample, ]

# A design comparison is only meaningful if all three matrices describe the same
# probes; fail loudly here rather than silently comparing different probe sets.
stopifnot(identical(rownames(beta_raw), rownames(beta_noob)),
          identical(rownames(beta_raw), rownames(beta_fun)))

grs_fun
```

::: {.callout-important}
## If `preprocessFunnorm` stops with `pthread_create() is 22`

On some installs — most often a conda environment — the chunk above dies partway
through normalization with:

```
Error in preprocessCore::normalize.quantiles.use.target(matrix(crtColumn.reduced),  : 
  ERROR; return code from pthread_create() is 22
```

Nothing is wrong with your data, and nothing is wrong with the chapter. The
quantile-normalization step inside `preprocessFunnorm` calls `preprocessCore`,
whose pre-built binary is compiled with threading enabled, and on these
installs the threads it asks for cannot be created. Setting
`OPENBLAS_NUM_THREADS=1` or `OMP_NUM_THREADS=1` does **not** help — the
threading has to come out of the package itself.

**The fix that works in any R installation** is to rebuild `preprocessCore`
from source with threading disabled:

```{.r filename="RStudio Console"}
BiocManager::install(
  "preprocessCore",
  configure.args = c(preprocessCore = "--disable-threading"),
  force  = TRUE,
  update = TRUE,
  type   = "source"
)
```

`BiocManager` will then ask about your other packages:

```
Update all/some/none? [a/s/n]:
```

Answer **`n`**. That prompt is about every *other* out-of-date package, not
about `preprocessCore` — the one you asked for is rebuilt either way. Answering
`a` recompiles a large part of the Bioconductor stack from source, which takes
a long time and can move packages this tutorial deliberately holds back (the
`r-lattice<0.23` pin in `envs/methyl.yml` exists for exactly that reason).

Restart R, then confirm you picked up the rebuilt copy before re-running the
chunk:

```{.r filename="RStudio Console"}
find.package("preprocessCore")   # should NOT be your conda env's library
```

**If you are in a conda environment** there is a second route: the threaded
OpenBLAS build is what `preprocessCore` is picking its threads up from, so
downgrading it also clears the error:

```{.bash filename="Terminal"}
conda activate ewas-methyl
conda install "openblas=0.3.3"
```

Be aware of what that transaction does, though. `openblas=0.3.3` is the old
standalone package, so conda removes `libopenblas` and satisfies BLAS and
LAPACK from `blis` and Netlib reference LAPACK instead. Reference LAPACK is
markedly slower than OpenBLAS, and every linear-algebra-heavy step after this
chapter — the control-probe PCA here, `sva` in chapter 05, `lmFit` in chapter
06 — pays for it. The source rebuild above leaves your numeric libraries alone,
so prefer it unless you have a reason not to.
:::

::: {.callout-note collapse="true"}
## If normalization does not fit on your machine

The two chunks above need the 460 MB `RGChannelSet` in memory and peak around
12.5 GB. If that is not practical, the published checkpoint holds exactly the
pieces the figures below use — the three 20,000-probe β matrices and the two
design lists — so you can fetch it and carry on:

```{.bash filename="Terminal"}
./get_data.sh C_normalized
```

```{.r filename="RStudio Console"}
np <- readRDS("data/02_norm_pieces.rds")
beta_raw  <- np$beta_raw
beta_noob <- np$beta_noob
beta_fun  <- np$beta_fun
typeI     <- np$typeI
typeII    <- np$typeII
```

Every figure and table in this chapter will then run. What you give up is
`grs_fun` itself, which [probe filtering](03_probe_filtering.qmd) reads — for
that chapter, fetch `./get_data.sh D_filtered` instead. You also give up
watching the normalization run, so take this route because you have to, not to
save two minutes.
:::

Then we can compare the distribution of signals across the three versions of the data and the probe types. 

```{r}
#| label: density-compare
#| fig-cap: "β-value density by probe design, before and after normalization (20k-probe sample). Raw Type II (dashed) is visibly pulled toward the center; noob and funnorm restore a more bimodal shape and bring the two designs closer together."
#| fig-width: 9
#| fig-height: 3.4
# Reshape one β matrix into the long form ggplot needs. 8,000 probes x 96 samples
# is already ~768,000 density points per method, which is plenty -- the extra
# thinning keeps the three panels quick to draw.
mk_long <- function(beta, label, typeII) {
  b <- beta[sample(nrow(beta), min(8000, nrow(beta))), ]
  design <- ifelse(rownames(b) %in% typeII, "Type II", "Type I")
  # as.vector() unstacks a matrix column by column, so the probe labels repeat
  # once per sample -- rep(..., times = ncol(b)) keeps them aligned to `beta`.
  data.frame(beta = as.vector(b),
             design = rep(design, times = ncol(b)),
             method = label)
}
long <- rbind(
  mk_long(beta_raw,  "raw",     typeII),
  mk_long(beta_noob, "noob",    typeII),
  mk_long(beta_fun,  "funnorm", typeII)
)
# Panels in processing order, not alphabetical: raw is the "before" picture.
long$method <- factor(long$method, levels = c("raw", "noob", "funnorm"))
ggplot(long, aes(x = beta, color = design, linetype = design)) +
  geom_density(na.rm = TRUE, linewidth = 0.7) +
  facet_wrap(~method) +
  scale_color_manual(values = c("Type I"  = "#1A6B75",     # teal
                                "Type II" = "#8C3A4A")) +  # plum
  labs(x = "β-value", y = "density",
       title = "Probe-design bias, before and after normalization") +
  theme(legend.position = "bottom")
```

```{r}
#| label: post-norm-table
# The same two-row summary as above, stacked across the three methods, so the
# Type I / Type II gap can be read down the column.
tab <- rbind(
  cbind(method = "raw",     type_summ(beta_raw,  typeI, typeII)),
  cbind(method = "noob",    type_summ(beta_noob, typeI, typeII)),
  cbind(method = "funnorm", type_summ(beta_fun,  typeI, typeII))
)
knitr::kable(tab, row.names = FALSE,
             caption = "Type I vs Type II β summaries across methods. The Type II spread widens toward the Type I spread after normalization. The medians do not converge, and are not expected to — see the text.")
```

Read the `spread_sd` column, not the medians. Type II's spread rises from **0.286
(raw)** to **0.313** after functional normalization, closing about a sixth of the
gap to the Type I spread of 0.385. That expansion of the compressed Type II range
is what design-bias correction does, and it is the same thing the density curves
above show.

The medians move the other way: Type II sits at **0.745** raw and **0.790** after
correction, so the median gap between the two designs widens slightly rather than
closing. That is not a failure of the correction, and it is worth understanding why.
Type I probes are enriched in CpG islands, which are predominantly unmethylated,
while Type II probes cover proportionally more open sea. The two rows are therefore
summarising different populations of CpGs, and the distance between their medians is
a fact about where each design sits in the genome rather than a measure of design
bias. A statistic that mixes the two cannot tell you whether normalization worked.

Functional
normalization additionally removes between-array technical variance that these
density plots do not show; its payoff appears later, in the [batch-effects](05_batch_effects.qmd)
notebook, as reduced technical structure in the principal components.

## 4. β-values vs M-values

Normalization outputs β-values (proportion methylated, bounded 0–1), which are
interpretable but **heteroscedastic** — their variance is small near 0 and 1 and
large near 0.5, which violates the constant-variance assumption of linear models.
For statistical testing we use the **M-value**, the log₂ ratio of methylated to
unmethylated signal [@du2010comparison]:

$$
M = \log_2\!\left(\frac{\beta}{1-\beta}\right)
$$

M-values are roughly homoscedastic and better behaved for the linear models used
in an EWAS. The convention is to **test on M-values and report/interpret effect
sizes on β-values** (a β difference of 0.05 is a "5% methylation change," which is
biologically legible; an M-value difference is not). It is becoming more common to 
report/interpret from M-values instead of beta values, so here we only show analysis
and reporting from M-values, but to report effect sizes from Beta values you would
follow the same steps, just with the Beta value matrix instead of the M-value matrix.

```{r}
#| label: mvalue-demo
#| fig-cap: "The β→M transform. β is bounded and heteroscedastic; M is unbounded and closer to constant-variance."
#| fig-width: 7
#| fig-height: 3
# The transform itself, drawn over the open interval: β = 0 and β = 1 map to
# -Inf and +Inf, which is the practical reason M-values need a bounded β.
b <- seq(0.001, 0.999, length.out = 500)
m <- log2(b / (1 - b))
dfm <- data.frame(beta = b, M = m)
ggplot(dfm, aes(beta, M)) +
  geom_line(color = "#1A6B75", linewidth = 0.9) +        # teal
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0.5, color = "gray70") +
  labs(x = "β-value", y = "M-value", title = "β → M-value transform")
```

## 5. Normalizing from a processed matrix

Many GEO series do not deposit raw IDATs, but do provide a processed matrix: 
a table of β-values (or sometimes M-values) with CpGs in rows and samples in columns, often as a
`GSExxxxx_series_matrix.txt.gz` or a supplementary file. You cannot run
`minfi`'s IDAT-based normalization on these, because the raw two-channel
intensities are gone. Here is how to work with them.

**Step 1 — get the matrix into R.** For a series matrix:

```{.r filename="RStudio Console"}
library(GEOquery)
gse  <- getGEO("GSExxxxxx", GSEMatrix = TRUE)      # returns a list of ExpressionSets
eset <- gse[[1]]
beta <- Biobase::exprs(eset)                        # CpG x sample β matrix
pheno <- Biobase::pData(eset)                       # sample metadata
```

For a supplementary flat file (`getGEOSuppFiles` downloads it), read it with
`data.table::fread()` and coerce to a matrix with CpG IDs as rownames.

**Step 2 — find out what was already done.** Read the series' processing description (`experimentData()` /
`pData(eset)$data_processing`) to learn whether the matrix is already normalized,
and how. Common cases:

- *Already noob + BMIQ (or funnorm) normalized* — the most common. Do **not**
  re-normalize; treat the matrix as analysis-ready β and proceed to probe filtering.
- *Raw/unnormalized β* — you can apply a **β-space** normalization such as BMIQ
  (`wateRmelon::BMIQ()`), which only needs β and probe design, not intensities.
- *Already batch-corrected (e.g. ComBat) or cell-composition-adjusted* — be
  cautious: further correction can over-adjust, and you have lost the ability to
  model those effects yourself. Note it as a limitation.

**Step 3 — QC on what you have.** Without intensities you lose detection
p-values and the intensity/sex-signal checks from the [QC](01_qc.qmd) notebook.
You can still:

- check for **missing values** and impute or drop probes with excessive `NA`;
- run **PCA/MDS** to spot outlier samples and technical structure;
- **predict sex** from the X/Y-chromosome β distributions (or `wateRmelon`'s
  `estimateSex`) and compare to the reported sex, catching label swaps;
- verify the probe count and platform match what you expect (EPIC ≈ 865k).

**Step 4 — harmonize probe IDs.** Processed matrices vary: some are already
masked/filtered, some use different ID conventions, some are 450K and some EPIC.
Before combining or comparing series, intersect on common `cg`/`ch` probe IDs and
apply the same probe-filtering ([next notebook](03_probe_filtering.qmd)) you would
apply to IDAT-derived data.

::: {.callout-tip}
## Practical rule
From IDATs you control normalization; from a processed matrix you **inherit** it.
The job shifts from "normalize" to "audit what was done and decide whether it is
adequate," then continue from probe filtering. Always record, in your methods, the
processing state of every series you use — reviewers will ask.
:::

We save the functional-normalized `GenomicRatioSet` for the downstream steps:

```{r}
#| label: save-norm
# 1. The handoff object: 865,859 CpGs x 96 samples, genome-mapped, noob +
#    functional normalized. This is what probe filtering (chapter 03) reads, and
#    it is the file deposited as Zenodo tier C_normalized. ~1 GB on disk.
saveRDS(grs_fun, "data/02_funnorm_grs.rds")

# 2. The pieces the figures above are drawn from: the three 20,000-probe β
#    matrices and the two design lists. Saving them means the resume callout in
#    section 3 restores exactly the objects this chapter computed, and that a
#    reader who cannot run normalization still gets every figure. Small next to
#    the GenomicRatioSet (~46 MB), and an RDS rather than a flat file because it
#    is three matrices and two character vectors, not one rectangle.
norm_pieces <- list(
  beta_raw  = beta_raw,    # 20,000 x 96, no correction
  beta_noob = beta_noob,   # 20,000 x 96, noob only
  beta_fun  = beta_fun,    # 20,000 x 96, noob + functional normalization
  typeI     = typeI,       # Type I probe IDs, whole array
  typeII    = typeII       # Type II probe IDs, whole array
)
saveRDS(norm_pieces, "data/02_norm_pieces.rds")
```

---

**Next:** [Probe filtering](03_probe_filtering.qmd) — removing probes that are
unreliable by design (cross-reactive, SNP-affected, sex-chromosome), using Zhou's
curated masks with ancestry awareness.

## References {.unnumbered}

::: {#refs}
:::
