---
title: "Normalization: making arrays comparable"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

The next step in methylation data pre-processing is normalization. There is variation 
in intensities from the different probe chemistries, as well as from technical variation 
in the assay, which needs to be accounted for before the probe intensities can be compared 
across samples in downstream analyses. Normalization is the method used to do this. This 
notebook walks through the standard normalization strategies used for methylation data.

Since all of the samples from our small example subset passed QC, we will start normaliztion 
from the raw two-channel object produced in [QC](01_qc.qmd). If you did identify samples that 
failed to meet QC thresholds, those samples should be removed prior to normalization.

```{r}
#| label: load
rg <- readRDS(data_path("01_RGset.rds"))
np <- readRDS(data_path("02_norm_pieces.rds"))
rg
```

## 1. Two sources of unwanted variation

**Within-array, technical.** Every Infinium measurement is a fluorescence
intensity, and fluorescence has a background floor (autofluorescence, non-specific
hybridization, optical noise) and a **dye bias**: the Cy3 (green) and Cy5 (red)
channels are not excited or detected with equal efficiency, so the same amount of
methylated and unmethylated DNA does not produce the same raw signal. Left
uncorrected, this shifts β-values systematically.

**Within-array, by probe design.** The EPIC array (like the 450K before it) mixes two 
assay chemistries [@pidsley2016epic]:

- **Type I** probes use two bead types per CpG (one for methylated, one for
  unmethylated), both read in the *same* color channel. They tolerate a wider
  dynamic range but use up two bead addresses per site.
- **Type II** probes use a single bead per CpG, with methylated and unmethylated
  states read in *different* channels. They are twice as space-efficient — which
  is why ~72% of EPIC probes are Type II — but their β-value distribution is
  **compressed toward the center**: they under-call fully methylated and fully
  unmethylated sites relative to Type I.

We can see this compression directly in the subset. Below are the median β and the
between-probe spread (SD of per-probe mean β) for each design, on raw data:

```{r}
#| label: type-bias-table
type_summ <- function(beta, typeI, typeII) {
  bi  <- beta[rownames(beta) %in% typeI, , drop = FALSE]
  bii <- beta[rownames(beta) %in% typeII, , drop = FALSE]
  data.frame(
    design      = c("Type I", "Type II"),
    n_probes    = c(nrow(bi), nrow(bii)),
    median_beta = round(c(median(bi, na.rm = TRUE), median(bii, na.rm = TRUE)), 3),
    spread_sd   = round(c(sd(rowMeans(bi, na.rm = TRUE)),
                          sd(rowMeans(bii, na.rm = TRUE))), 3)
  )
}
knitr::kable(type_summ(np$beta_raw, np$typeI, np$typeII),
             caption = "Raw β by probe design (20k-probe sample). Type II sits closer to 0.5 with a narrower spread — the design-bias normalization must correct.")
```

::: {.callout-important}
## Why this matters for an EWAS
If Type I and Type II probes are on different scales, a differential-methylation
test will partly reflect *which chemistry* a probe uses rather than biology, and
effect sizes are not comparable across probe types.
:::

## 2. Normalization methods

Methods fall into two groups by *what* they try to make comparable.

### Within-array methods (fix probe-design bias per sample)

- **noob** (normal-exponential out-of-band background correction) models the
  background as a normal-exponential mixture using the out-of-band Type I signals,
  then applies a dye-bias correction [@triche2013noob]. It is a background/dye
  step, not a design-bias step, but it is the standard first move and is what most
  pipelines build on.
- **BMIQ** (beta-mixture quantile) fits a three-state beta mixture
  (unmethylated / hemimethylated / methylated) separately to Type I and Type II
  probes, then transforms the Type II distribution onto the Type I scale
  [@teschendorff2013bmiq]. It directly targets the design bias.
- **SWAN** (subset-quantile within-array normalization) matches Type I and Type II
  intensity distributions using a subset of probes with matched underlying biology
  [@maksimovic2012swan].

### Between-array methods (make samples comparable to each other)

- **Quantile normalization** forces every sample's intensity distribution to a
  common reference. Powerful, but it assumes the *global* methylation distribution
  is the same across samples — a poor assumption when groups differ biologically
  (e.g. tumor vs normal), where it can erase real signal.
- **Functional normalization (funnorm)** is a between-array method designed for
  exactly that case [@fortin2014funnorm]. Instead of forcing distributions to
  match, it regresses out variation explained by the array's **internal control
  probes** — bisulfite conversion, staining, extension, hybridization controls —
  which capture technical variation *without* being tied to biological signal. It
  is the recommended default when you expect large biological differences between
  groups, and it wraps noob as its first step.

::: {.callout-note}
## Choosing a method
There is no single winner. A defensible default for blood EWAS is **noob +
functional normalization** (what `preprocessFunnorm` does in one call), because it
corrects background/dye and removes technical variation without assuming samples
share a methylation distribution. If your study is a within-tissue comparison with
small expected differences, quantile-based methods (or `preprocessQuantile`) are
also reasonable.
:::

## 3. Running it on the subset

We compute three versions and compare: **raw** (`preprocessRaw`, no correction),
**noob** (`preprocessNoob`), and **functional** (`preprocessFunnorm`). The
functional result is a `GenomicRatioSet` — CpGs mapped to the genome with β and
M-values ready for analysis.

The code to produce these subsets is shown below. It may take several minutes to 
complete. 

```r
mset_raw  <- preprocessRaw(rg);   beta_raw  <- getBeta(mset_raw)
mset_noob <- preprocessNoob(rg);  beta_noob <- getBeta(mset_noob)
grs_fun   <- preprocessFunnorm(rg); beta_fun <- getBeta(grs_fun)  # GenomicRatioSet
```

Then we can compare the distribution of signals across the three versions of the data and the probe types. 

```{r}
#| label: density-compare
#| fig-cap: "β-value density by probe design, before and after normalization (20k-probe sample). Raw Type II (dashed) is visibly pulled toward the center; noob and funnorm restore a more bimodal shape and bring the two designs closer together."
#| fig-width: 9
#| fig-height: 3.4
library(ggplot2)
mk_long <- function(beta, label, typeII) {
  b <- beta[sample(nrow(beta), min(8000, nrow(beta))), ]
  v <- as.vector(b)
  design <- ifelse(rownames(b) %in% typeII, "Type II", "Type I")
  data.frame(beta = as.vector(b),
             design = rep(design, times = ncol(b)),
             method = label)
}
long <- rbind(
  mk_long(np$beta_raw,  "raw",     np$typeII),
  mk_long(np$beta_noob, "noob",    np$typeII),
  mk_long(np$beta_fun,  "funnorm", np$typeII)
)
long$method <- factor(long$method, levels = c("raw", "noob", "funnorm"))
ggplot(long, aes(x = beta, color = design, linetype = design)) +
  geom_density(na.rm = TRUE, linewidth = 0.7) +
  facet_wrap(~method) +
  scale_color_manual(values = c("Type I" = "#4C72B0", "Type II" = "#C44E52")) +
  labs(x = "β-value", y = "density",
       title = "Probe-design bias, before and after normalization") +
  theme(legend.position = "bottom")
```

```{r}
#| label: post-norm-table
tab <- rbind(
  cbind(method = "raw",     type_summ(np$beta_raw,  np$typeI, np$typeII)),
  cbind(method = "noob",    type_summ(np$beta_noob, np$typeI, np$typeII)),
  cbind(method = "funnorm", type_summ(np$beta_fun,  np$typeI, np$typeII))
)
knitr::kable(tab, row.names = FALSE,
             caption = "Type I vs Type II β summaries across methods. The gap in median β between the two designs narrows after normalization.")
```

The Type II median moves from **0.75 (raw)** toward the Type I scale after
correction, and the design-specific density curves in the figure come into closer
agreement — exactly what design-bias correction should do. Functional
normalization additionally removes between-array technical variance that these
density plots do not show; its payoff appears later, in the [batch-effects](05_batch_effects.qmd)
notebook, as reduced technical structure in the principal components.

## 4. β-values vs M-values

Normalization outputs β-values (proportion methylated, bounded 0–1), which are
interpretable but **heteroscedastic** — their variance is small near 0 and 1 and
large near 0.5, which violates the constant-variance assumption of linear models.
For statistical testing we use the **M-value**, the log₂ ratio of methylated to
unmethylated signal [@du2010comparison]:

$$
M = \log_2\!\left(\frac{\beta}{1-\beta}\right)
$$

M-values are roughly homoscedastic and better behaved for the linear models used
in an EWAS. The convention is to **test on M-values and report/interpret effect
sizes on β-values** (a β difference of 0.05 is a "5% methylation change," which is
biologically legible; an M-value difference is not). It is becoming more common to 
report/interpret from M-values instead of beta values, so here we only show analysis
and reporting from M-values, but to report effect sizes from Beta values you would
follow the same steps, just with the Beta value matrix instead of the M-value matrix.

```{r}
#| label: mvalue-demo
#| fig-cap: "The β→M transform. β is bounded and heteroscedastic; M is unbounded and closer to constant-variance."
#| fig-width: 7
#| fig-height: 3
b <- seq(0.001, 0.999, length.out = 500)
m <- log2(b / (1 - b))
dfm <- data.frame(beta = b, M = m)
ggplot(dfm, aes(beta, M)) + geom_line(color = "#4C72B0", linewidth = 0.9) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0.5, color = "gray70") +
  labs(x = "β-value", y = "M-value", title = "β → M-value transform")
```

## 5. Normalizing from a processed matrix

Many GEO series do not deposit raw IDATs, but do provide a processed matrix: 
a table of β-values (or sometimes M-values) with CpGs in rows and samples in columns, often as a
`GSExxxxx_series_matrix.txt.gz` or a supplementary file. You cannot run
`minfi`'s IDAT-based normalization on these, because the raw two-channel
intensities are gone. Here is how to work with them.

**Step 1 — get the matrix into R.** For a series matrix:

```r
library(GEOquery)
gse  <- getGEO("GSExxxxxx", GSEMatrix = TRUE)      # returns a list of ExpressionSets
eset <- gse[[1]]
beta <- Biobase::exprs(eset)                        # CpG x sample β matrix
pheno <- Biobase::pData(eset)                       # sample metadata
```

For a supplementary flat file (`getGEOSuppFiles` downloads it), read it with
`data.table::fread()` and coerce to a matrix with CpG IDs as rownames.

**Step 2 — find out what was already done.** Read the series' processing description (`experimentData()` /
`pData(eset)$data_processing`) to learn whether the matrix is already normalized,
and how. Common cases:

- *Already noob + BMIQ (or funnorm) normalized* — the most common. Do **not**
  re-normalize; treat the matrix as analysis-ready β and proceed to probe filtering.
- *Raw/unnormalized β* — you can apply a **β-space** normalization such as BMIQ
  (`wateRmelon::BMIQ()`), which only needs β and probe design, not intensities.
- *Already batch-corrected (e.g. ComBat) or cell-composition-adjusted* — be
  cautious: further correction can over-adjust, and you have lost the ability to
  model those effects yourself. Note it as a limitation.

**Step 3 — QC on what you have.** Without intensities you lose detection
p-values and the intensity/sex-signal checks from the [QC](01_qc.qmd) notebook.
You can still:

- check for **missing values** and impute or drop probes with excessive `NA`;
- run **PCA/MDS** to spot outlier samples and technical structure;
- **predict sex** from the X/Y-chromosome β distributions (or `wateRmelon`'s
  `estimateSex`) and compare to the reported sex, catching label swaps;
- verify the probe count and platform match what you expect (EPIC ≈ 865k).

**Step 4 — harmonize probe IDs.** Processed matrices vary: some are already
masked/filtered, some use different ID conventions, some are 450K and some EPIC.
Before combining or comparing series, intersect on common `cg`/`ch` probe IDs and
apply the same probe-filtering ([next notebook](03_probe_filtering.qmd)) you would
apply to IDAT-derived data.

::: {.callout-tip}
## Practical rule
From IDATs you control normalization; from a processed matrix you **inherit** it.
The job shifts from "normalize" to "audit what was done and decide whether it is
adequate," then continue from probe filtering. Always record, in your methods, the
processing state of every series you use — reviewers will ask.
:::

We save the functional-normalized `GenomicRatioSet` for the downstream steps:

```r
saveRDS(grs_fun, data_path("02_funnorm_grs.rds"))
```

---

**Next:** [Probe filtering](03_probe_filtering.qmd) — removing probes that are
unreliable by design (cross-reactive, SNP-affected, sex-chromosome), using Zhou's
curated masks with ancestry awareness.

## References {.unnumbered}

::: {#refs}
:::
