---
title: "Probe filtering"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

Not every one of the ~865,000 EPIC probes should be used in downstream analyses. 
Some probes need to be removed because they were designed for assessing quality 
control, as we did in [QC](01_qc.qmd). Other probes need to be removed because 
they have been found to be problematic in some way: cross-hybridization to more than 
one place in the genome; proximity to a common SNP (so what looks like a methylation 
difference is really a genotype difference); mapping to repetitive regions. These are 
properties of the **probe**, so these problematic probes are filtered out from everyone. 
This notebook explains the categories and applies Wanding Zhou's curated
masks [@zhou2017comprehensive] for Illumina methylation arrays. 

```{r}
#| label: load
grs   <- readRDS(data_path("03_grs_filtered.rds"))   # already-filtered result, for display
funl  <- readRDS(data_path("03_filter_funnel.rds"))
mp    <- readRDS(data_path("03_mask_pieces.rds"))
```

## 1. Why some probes are unreliable by design

There are three main types of problematic probes and each has been characterized 
in dedicated methods papers:

**Cross-reactive / non-specific probes.** A 50-mer probe sequence is not
guaranteed to be unique in the genome. Some probes hybridize to multiple loci, so
their signal is a blend from several genomic locations. Chen et al.
[@chen2013crossreactive] cataloged cross-reactive probes on the 450K; McCartney
et al. [@mccartney2016identification] extended this to the EPIC array. Zhou's
`MASK_mapping` flag captures probes that do not map uniquely.

**SNP-affected probes.** If a common SNP falls in the probe body — especially at
the interrogated CpG or the single-base extension position — then the assay may 
reflect **genotype**, and not methylation. Two individuals with different alleles 
will show a large "methylation difference" that is really a difference in DNA 
sequence. This is a dangerous and often overlooked category for an EWAS, because 
it produces reproducible, highly significant, and completely spurious associations 
whenever the SNP allele frequency differs between your comparison groups. Zhou provides 
SNP masks recommendations built from 1000 Genomes allele frequencies. If you are studying a 
population that is not well represented by one of the 1000 Genomes Project, it is 
recommended to find or curate a list of population-specific SNPs and exclude any CpGs 
within the last 5bp of the CpG probe. If there is no appropriate 1000 Genomes population and 
you don't have population/study-specific allele frequencies, you can use the general population SNP
mask and note this as a limitation of the study. If you are performing a study on a multi-ancestry 
cohort or don't know the ancestry of your cohort, then the general population masking is appropriate.

**Repetitive / structurally masked probes.** Probes overlapping RepeatMasker
regions, copy-number-variable segments, or requiring an extension-base substitution
have also been flagged historically — as `MASK_rmsk15`, `MASK_sub25_copy` and
`MASK_typeINextBaseSwitch` in the legacy manifest files from Zhou's annotation database. The current v8.1
mask retains only the extension-base category (as `M_1baseSwitchSNP_*`), and the
repeat and copy-number terms are now handled by the `MASK_mapping` criteria (based on a
mapping quality score).

The latest CpG masking recommendations from Zhou (`EPIC.hg38.mask.cm`, v8.1) include
24 separate mask terms. One of those terms is `M_general`, which
combines the recommended masking for probes with problematic mapping (the probes
that should be masked independent of population) with CpGs within 5bp of common SNPs
(MAF ≥ 5%) for the general population. If you do have a specific population, the
remaining terms give population-specific masking for the 5 super-populations from
the 1000 Genomes. The population-specific masking is a bit more stringent, as it
uses MAF ≥ 1% — which is what the `_1pt` suffix on those term names records.

The terms are compositional: rather than a single pre-baked per-population column,
you select the mapping terms plus the SNP terms for each ancestry you need and take
their union. The full set of 24 terms for EPIC is:

- **mapping / specificity** — `M_mapping`, `M_nonuniq`
- **SNP at the interrogated CpG**, per population — `M_SNP_{AFR,EUR,EAS,SAS,AMR}_1pt`,
  plus `M_SNPcommon_1pt` and `M_SNPcommon_5pt` for the general population
- **SNP at the extension base** — `M_2extBase_SNP_*` (same population suffixes)
- **Type I extension-base switch** — `M_1baseSwitchSNP_*`
- **the recommended general default** — `M_general`

If you need more specific sub-population masking, there is an archived version of the
recommended population masking (`EPIC.hg38.manifest.pop.tsv.gz`) that includes the 5
super-populations and 26 sub-populations. If using the archived population masking
file, the `MASK_general_<POP>` column includes the recommended problematic probes as
well as the population-specific probes. The `MASK_snp5_<POP>` columns only flag the
population-specific masking.

::: {.callout-note}
## Why use Zhou's masks specifically
The Zhou lab re-mapped every Infinium probe against the genome from scratch and
integrated SNP information across 1000 Genomes populations
[@zhou2017comprehensive]. The masks are the annotation layer underneath the
`sesame` package [@zhou2018sesame] and are the current community standard.
:::

### Getting the mask: YAME

The v8.1 mask ships as `EPIC.hg38.mask.cm` — a bit-packed **CX-format** file rather
than a table. It stores one bit per probe per mask term over the array's canonical
probe ordering, which is why the whole thing is only ~360 kB where the equivalent
TSV is 2.8 MB. Because the file carries no probe names (row *i* means whatever row
*i* of the ordering means), you fetch the ordering alongside it.

Read it with [YAME](https://github.com/zhou-lab/YAME), the Zhou lab's encoder for
these formats:

```bash
# install: conda install -c zhou-lab -c conda-forge yame
# (bioconda's linux build is pinned at 1.8, which predates `yame fetch` --
#  use the zhou-lab channel or build from source)

yame fetch EPIC/EPIC.hg38.mask.cm EPIC/EPIC.ordering.tsv.gz

yame info    EPIC.hg38.mask.cm   # 24 mask terms over 866,553 probes
yame summary EPIC.hg38.mask.cm   # how many probes each term flags
yame unpack -a EPIC.hg38.mask.cm > mask_bits.tsv   # all 24 terms, one column each
```

`yame summary` is the quickest way to see what a term includes you before you commit
to it. Turning the bits into a probe list is then a row-wise OR across the columns
you chose, indexed against the ordering:

```r
library(data.table)
terms  <- names(fread("mask.cm.idx", header = FALSE))  # term order = column order
probes <- fread("EPIC.ordering.tsv.gz")$Probe_ID
bits   <- fread("mask_bits.tsv", header = FALSE); setnames(bits, terms)

drop <- probes[Reduce(`|`, lapply(mask_cols, function(m) bits[[m]] == 1L))]
```

::: {.callout-note}
## The archived manifest is still there when you need it
`EPIC.hg38.manifest.pop.tsv.gz` remains available and is the right choice when you
need **sub-population** resolution: it carries `MASK_general_<POP>` and
`MASK_snp5_<POP>` columns for the 5 super-populations *and* 26 sub-populations
(GWD, YRI, ASW, CEU, FIN, …), where v8.1's `.cm` stops at the 5 super-populations.
Note also that the two generations do not mask the same categories — see the
comparison at the end of this section.
:::

Broken down, the probe removal includes: 

```{r}
#| label: funnel-table
funnel <- data.frame(
  step = c("Start (normalized probes)",
           "− Zhou v8.1 mask (8 codes, AFR + EUR)",
           "− Detection-p failures (>10% of samples)",
           "− Sex chromosomes (autosomal EWAS)",
           "= Retained (EWAS-ready)"),
  probes_dropped = c(NA, funl$drop_mask, funl$drop_detp, funl$drop_sex, NA),
  probes_remaining = c(funl$n0, NA, NA, NA, funl$retained)
)
knitr::kable(funnel, caption = "Probe-filtering funnel on the 96-sample subset (AFR + EUR aware).")
cat(sprintf("Retained %d of %d probes (%.1f%%).\n",
            funl$retained, funl$n0, 100 * funl$retained / funl$n0))
```

## 2. The categories, quantified on EPIC

The SNP mask is **not the same for every study population**, because a "common
SNP" depends on which population you ask about. The number of SNP-affected probes varies substantially by ancestry:

```{r}
#| label: pop-snp-table
snp <- data.frame(
  population  = c("AFR (African)", "AMR (Admixed American)", "SAS (South Asian)",
                  "EUR (European)", "EAS (East Asian)"),
  snp_masked  = funl$snp_by_pop[c("AFR", "AMR", "SAS", "EUR", "EAS")]
)
knitr::kable(snp, row.names = FALSE,
             caption = "Probes flagged by M_SNP_<POP>_1pt (common SNP at the interrogated CpG, MAF >= 1%) by 1000 Genomes super-population. African-ancestry samples have roughly twice as many SNP-affected probes as European.")
```

African-ancestry populations carry the most common genetic variation, so more
probes overlap a common SNP. Concretely, in these masks
**`r format(funl$snp_by_pop[["AFR"]], big.mark = ",")`** probes are SNP-flagged for
AFR versus **`r format(funl$snp_by_pop[["EUR"]], big.mark = ",")`** for EUR:

```{r}
#| label: afr-eur-overlap
afr <- mp$per_code_sets$M_SNP_AFR_1pt
eur <- mp$per_code_sets$M_SNP_EUR_1pt
cat("masked in AFR but NOT EUR:", length(setdiff(afr, eur)), "\n")
cat("masked in EUR but NOT AFR:", length(setdiff(eur, afr)), "\n")
cat("masked in both:           ", length(intersect(afr, eur)), "\n")
```

While there is some overlap between the AFR and EUR recommended SNP masks, there are distinct maskings for each
population.

## 3. Ancestry matters

The Grady Trauma Project subset is **~95% African American**. If we filtered with 
a general or only European-population mask (a common default), we would leave in 
probes that are SNP-affected in African-ancestry individuals. As African American 
individuals tend to have an admixture of both European and African Ancestry, it is 
important to include masking recommendations for both.

```{r}
#| label: ancestry-contrast
#| fig-cap: "Probes retained under the AFR+EUR union mask used here versus a EUR-only mask. Filtering this African-American cohort on European SNP flags alone would keep tens of thousands of probes that are SNP-affected in African-ancestry individuals."
#| fig-width: 6.5
#| fig-height: 3
library(ggplot2)
contrast <- data.frame(
  mask = c("AFR + EUR union\n(used here)", "EUR only\n(wrong here)"),
  retained = c(funl$retained, funl$retained_eur)
)
ggplot(contrast, aes(mask, retained, fill = mask)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = format(retained, big.mark = ",")), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("#4C72B0", "#C44E52"), guide = "none") +
  labs(x = NULL, y = "probes retained",
       title = "Ancestry-aware probe filtering") +
  expand_limits(y = max(contrast$retained) * 1.06)
```

```{r}
#| label: ancestry-diff
cat("Probes kept by a EUR-only mask but dropped by the AFR + EUR union:",
    funl$afr_not_eur_kept, "\n")
```

The union is what the code below applies, and it is the operational meaning of
"include masking recommendations for both".

So, if we are using the latest recommended masking file (v8.1), the columns we would want
to include in the masking would be `M_mapping`, `M_nonuniq`, `M_SNP_AFR_1pt`,`M_1baseSwitchSNP_AFR_1pt`,
`M_2extBase_SNP_AFR_1pt`, `M_SNP_EUR_1pt`, `M_1baseSwitchSNP_EUR_1pt`, and `M_2extBase_SNP_EUR_1pt`. If 
we were using the archived manifest population file, we would use the `MASK_general_AFR` and `MASK_snp5_EUR`.


### What changing mask generation costs you

It is worth seeing the size of the version effect directly, because it runs the
opposite way to intuition. The eight v8.1 codes cover **two** ancestries; the legacy
`MASK_general_AFR` covers one. Yet the legacy mask is the more aggressive:

```{r}
#| label: mask-generation-contrast
gen <- data.frame(
  mask = c("Legacy MASK_general_AFR", "v8.1, 8 codes (AFR + EUR)"),
  probes_masked   = c(124132, funl$union8),
  probes_retained = c(724282, funl$retained)
)
knitr::kable(gen, row.names = FALSE,
             caption = "Same cohort, same detection-p and sex filters; only the mask generation differs.")
```

The difference is the repeat and copy-number terms described in §2: the legacy
`MASK_general_<POP>` bundles them in, and v8.1 has no equivalent. So moving to v8.1
retains **`r format(funl$retained - 724282, big.mark = ",")`** more probes here — more
sites tested, but also the repeat-overlapping probes no longer excluded. Neither
choice is automatically right; what matters is that you state which mask generation
and which terms you used, since the two are not interchangeable.

::: {.callout-important}
## The teaching point
Probe filtering is not a fixed recipe — the SNP component depends on your study
population. For a cohort of known, relatively homogeneous ancestry, use that
population's mask. For an admixed or mixed-ancestry cohort, the conservative choice
is to take the **union** of the relevant population masks (drop a probe if it is
SNP-affected in *any* represented ancestry). For a multi-ancestry cohort or cohort 
of unknown ancestry, the general population mask is the most appropriate.
:::

::: {.callout-note}
## Smoking-associated probes
Everything above masks probes that are unreliable *by design*. There is a separate,
optional reason to drop probes: smoking is one of the largest environmental drivers
of blood methylation, and if your dataset has no smoking variable — as the Grady
deposit does not — one of the available strategies is to exclude CpGs known from the
literature to be smoking-associated, at this filtering stage rather than after the
fact. That is a decision about confounding, not about probe quality, so it does not
belong in the funnel above. This tutorial does exclude smoking-associated CpGs, but
at a later step as these probes are needed to create a proxy smoking variable. See
[the smoking section in the EWAS chapter](06_ewas.qmd) for the full set of options.
:::

## 4. Detection-p and sex-chromosome filtering

Two study-specific filters complete the funnel:

- **Detection-p failures.** From [QC](01_qc.qmd) we have the per-probe detection
  p-values. A probe that fails (p > 0.01) in more than ~10% of samples is
  unreliable *in this dataset* and is dropped. Here that removed
  `r funl$drop_detp` probes.
- **Sex chromosomes.** For an autosomal EWAS, X- and Y-chromosome probes are also removed 
  (`r funl$drop_sex` probes): X-inactivation and sex differences would
  otherwise dominate, and they need sex-stratified handling if analyzed. They are
  *removed from the autosomal analysis*, not deleted from the study — a sex-specific
  analysis would use them deliberately.

The exact filtering code (run once, checkpointed):

```r
library(data.table)

# --- Zhou v8.1 mask, read from the bit-packed .cm (see "Getting the mask") ---
terms  <- fread("EPIC.hg38.mask.cm.idx", header = FALSE)$V1  # term = column order
probes <- fread("EPIC.ordering.tsv.gz")$Probe_ID
bits   <- fread("mask_bits.tsv", header = FALSE)             # yame unpack -a
setnames(bits, terms)

# mapping/specificity terms, plus SNP terms for BOTH represented ancestries
mask_cols <- c("M_mapping", "M_nonuniq",
               "M_SNP_AFR_1pt", "M_1baseSwitchSNP_AFR_1pt", "M_2extBase_SNP_AFR_1pt",
               "M_SNP_EUR_1pt", "M_1baseSwitchSNP_EUR_1pt", "M_2extBase_SNP_EUR_1pt")
drop8 <- probes[Reduce(`|`, lapply(mask_cols, function(m) bits[[m]] == 1L))]

drop_mask <- intersect(rownames(grs), drop8)
# Detection-p failures in >10% of samples
frac_fail <- rowMeans(detP[rownames(grs), ] > 0.01)
drop_detp <- rownames(grs)[frac_fail > 0.10]
# Sex chromosomes
drop_sex  <- rownames(grs)[getAnnotation(grs)$chr %in% c("chrX", "chrY")]
# Apply
keep <- setdiff(rownames(grs), Reduce(union, list(drop_mask, drop_detp, drop_sex)))
grs_filt <- grs[keep, ]
```

## 5. When you don't have IDATs: filtering a processed matrix

Probe filtering is one of the few preprocessing steps that works **identically**
whether your data came from IDATs or from a processed β-matrix — because the masks
are keyed on **probe IDs**, not intensities. The workflow:

1. **Load the matrix** (see the [normalization](02_normalization.qmd) non-IDAT
   section) so you have a β-matrix with `cg`/`ch` probe IDs as rownames.
2. **Apply the same masks by ID.** Read Zhou's mask/manifest for the correct array
   (450K vs EPIC vs EPICv2 — they have different probe sets), pick the mask for
   your study population, and subset:

   ```r
   # v8.1 (recommended): the union built above works unchanged on a matrix,
   # because the mask is keyed on probe IDs
   beta_filt <- beta[!rownames(beta) %in% drop8, ]

   # archived manifest, if you need sub-population resolution instead
   library(data.table)
   pop  <- fread("EPIC.hg38.manifest.pop.tsv.gz")
   drop <- pop$probeID[pop$MASK_general_ASW]           # e.g. African-ancestry SW US
   beta_filt <- beta[!rownames(beta) %in% drop, ]
   ```
3. **You lose the detection-p filter** unless the submitter deposited detection
   p-values alongside the matrix (some do, as a separate supplementary file — use
   it if present). Without it, rely on the mask plus removing probes with excessive
   `NA` values.
4. **Sex chromosomes** filter the same way, by intersecting with the manifest's
   chromosome annotation. It is possible sex chromosome CpGs will have already been 
   removed from a pre-processed dataset. 

::: {.callout-tip}
## Match the mask to the array *and* the genome build
Zhou publishes masks per platform (450K / EPIC / EPICv2) and per genome build
(hg19 / hg38). Use the manifest that matches the array the data were generated on;
the probe IDs mostly overlap between 450K and EPIC but the mask *contents* differ.
If you are combining a 450K series with an EPIC series, filter each on its own
manifest, then intersect on the shared probe IDs.
:::

We save the filtered, EWAS-ready object for cell-composition estimation next:

```r
saveRDS(grs_filt, data_path("03_grs_filtered.rds"))
```

The result: **`r format(funl$retained, big.mark=",")`** autosomal probes on
96 samples, filtered under the Zhou v8.1 mask for mapping quality, SNP masking
appropriate to an admixed African-American cohort (AFR and EUR terms together),
detection reliability, and sex chromosomes.

Nearly every probe that survives here is tested in the EWAS. Two small deductions
happen later, both in [chapter 05](05_batch_effects.qmd):

- the **20 CpGs used to build the smoking proxy** are excluded, because regressing
  those probes on a score derived from themselves is circular; and
- **two probes** (`cg17759086`, `cg01801182`) have β pinned at 0 or 1 in the
  87-sample analysis subset, so their M-values are ±∞ and they cannot be fitted.

So `r format(funl$retained, big.mark=",")` probes pass filtering and
**756,251** are actually tested.

---

**Next:** [Cell composition](04_cell_composition.qmd) — whole blood is a mixture of
cell types with very different methylomes; we estimate the proportions and validate
against the study's deposited values.

## References {.unnumbered}

::: {#refs}
:::
