---
title: "Probe filtering"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

Not every one of the ~865,000 EPIC probes should be used in downstream analyses. 
Some probes need to be removed because they were designed for assessing quality 
control, as we did in [QC](01_qc.qmd). Other probes need to be removed because 
they have been found to be problematic in some way: cross-hybridization to more than 
one place in the genome; proximity to a common SNP (so what looks like a methylation 
difference is really a genotype difference); mapping to repetitive regions. These are 
properties of the **probe**, so these problematic probes are filtered out from everyone. 
This notebook explains the categories and applies Wanding Zhou's curated
masks [@zhou2017comprehensive] for Illumina methylation arrays. 

::: {.callout-note}
## What this chapter needs
This chapter starts from `data/02_funnorm_grs.rds`, written at the end of
[normalization](02_normalization.qmd). If you did not run that chapter, fetch the
published checkpoint instead — note it is a 1.25 GB download:

```{.bash filename="Terminal"}
./get_data.sh C_normalized
```
:::

```{r}
#| label: load
library(minfi)       # getAnnotation() for probe coordinates, detectionP() for the p-value matrix
library(data.table)  # fread() for the mask tables

# The functional-normalized GenomicRatioSet written at the end of
# [normalization](02_normalization.qmd): 96 samples, beta and M-values for every
# probe that survived normalization. Nothing is filtered yet -- that is this
# chapter's job.
grs <- readRDS("data/02_funnorm_grs.rds")
grs
```

## 1. Why some probes are unreliable by design {#sec-why}

There are three main types of problematic probes and each has been characterized 
in dedicated methods papers:

**Cross-reactive / non-specific probes.** A 50-mer probe sequence is not
guaranteed to be unique in the genome. Some probes hybridize to multiple loci, so
their signal is a blend from several genomic locations. Chen et al.
[@chen2013crossreactive] cataloged cross-reactive probes on the 450K; McCartney
et al. [@mccartney2016identification] extended this to the EPIC array. Zhou's
`MASK_mapping` flag captures probes that do not map uniquely.

**SNP-affected probes.** If a common SNP falls in the probe body — especially at
the interrogated CpG or the single-base extension position — then the assay may 
reflect **genotype**, and not methylation. Two individuals with different alleles 
will show a large "methylation difference" that is really a difference in DNA 
sequence. This is a dangerous and often overlooked category for an EWAS, because 
it produces reproducible, highly significant, and completely spurious associations 
whenever the SNP allele frequency differs between your comparison groups. Zhou provides 
SNP masks recommendations built from 1000 Genomes allele frequencies. If you are studying a 
population that is not well represented by one of the 1000 Genomes Project, it is 
recommended to find or curate a list of population-specific SNPs and exclude any CpGs 
within the last 5bp of the CpG probe. If there is no appropriate 1000 Genomes population and 
you don't have population/study-specific allele frequencies, you can use the general population SNP
mask and note this as a limitation of the study. If you are performing a study on a multi-ancestry 
cohort or don't know the ancestry of your cohort, then the general population masking is appropriate.

**Repetitive / structurally masked probes.** Probes overlapping RepeatMasker
regions, copy-number-variable segments, or requiring an extension-base substitution
have also been flagged historically — as `MASK_rmsk15`, `MASK_sub25_copy` and
`MASK_typeINextBaseSwitch` in the legacy manifest files from Zhou's annotation database. The current v8.1
mask retains only the extension-base category (as `M_1baseSwitchSNP_*`), and the
repeat and copy-number terms are now handled by the `MASK_mapping` criteria (based on a
mapping quality score).

The latest CpG masking recommendations from Zhou (`EPIC.hg38.mask.cm`, v8.1) include
24 separate mask terms. One of those terms is `M_general`, which
combines the recommended masking for probes with problematic mapping (the probes
that should be masked independent of population) with CpGs within 5bp of common SNPs
(MAF ≥ 5%) for the general population. If you do have a specific population, the
remaining terms give population-specific masking for the 5 super-populations from
the 1000 Genomes. The population-specific masking is a bit more stringent, as it
uses MAF ≥ 1% — which is what the `_1pt` suffix on those term names records.

The terms are compositional: rather than a single pre-baked per-population column,
you select the mapping terms plus the SNP terms for each ancestry you need and take
their union. The full set of 24 terms for EPIC is:

- **mapping / specificity** — `M_mapping`, `M_nonuniq`
- **SNP at the interrogated CpG**, per population — `M_SNP_{AFR,EUR,EAS,SAS,AMR}_1pt`,
  plus `M_SNPcommon_1pt` and `M_SNPcommon_5pt` for the general population
- **SNP at the extension base** — `M_2extBase_SNP_*` (same population suffixes)
- **Type I extension-base switch** — `M_1baseSwitchSNP_*`
- **the recommended general default** — `M_general`

::: {.callout-note}
## Why use Zhou's masks specifically
The Zhou lab re-mapped every Infinium probe against the genome from scratch and
integrated SNP information across 1000 Genomes populations
[@zhou2017comprehensive]. The masks are the annotation layer underneath the
`sesame` package [@zhou2018sesame] and are the current community standard.
:::

### Getting the mask

The mask comes in two forms, and **this chapter uses both** — they are not
alternatives you pick between. The ready-made TSV carries the eight codes the
filtering is built on, and that is all you need to reproduce the probe set. The full
release carries all 24 codes, three of which the per-population table in
[§2](#sec-ancestry) needs and the TSV does not have. Start with the TSV; read the
second part when you want a term outside the eight.

#### The ready-made TSV — the eight codes the filtering uses {#the-tsv}

The eight mask codes this chapter uses are shipped with the tutorial data as a plain
tab-delimited table, so you can filter probes with `fread()` and nothing else. No new
software to install:

::: {.callout-note}
## Where the mask table comes from
`data/EPIC.hg38.mask.v81.8code.tsv.gz` is tracked in the repository, so a fresh clone
already has it — 2.9 MB, no download needed. It is also inside the `D_filtered` data
tier, which is where it lived before; if you have already fetched that tier you have
two copies of the same file and can ignore this note.

It is a derived table, not a Zhou lab release: [the full release](#route-2) below shows the
`yame` commands that produce it from the published `.cm` mask, so you can rebuild it
for another array or another mask version.
:::

```{r}
#| label: mask-read
# One row per probe on the array, one 0/1 column per mask code, plus the
# row-wise OR of the eight in `mask_v81_8code`.
mask <- fread("data/EPIC.hg38.mask.v81.8code.tsv.gz")

dim(mask)                            # 866,553 probes x 10 columns
sum(mask$mask_v81_8code)             # 90,832 probes flagged by the union

# the probes to drop
drop8 <- mask[mask_v81_8code == 1L, Probe_ID]
```

The table has one row per probe on the array and one 0/1 column per mask code, plus a
`mask_v81_8code` column that is the row-wise OR of the eight — which is exactly the
filter this chapter applies. Because the individual code columns are kept, you can
also build a different combination: to mask on African-ancestry SNP flags only, OR
the three `*_AFR_1pt` columns and ignore the rest.

This file was generated from the official `.cm` release with YAME (below) and verified
to reproduce all eight per-code counts and the 90,832-probe union exactly. It is a
convenience layer, not a separate source of truth.

#### The full release with YAME — all 24 codes {#route-2}

Use this if you want a mask term the TSV does not carry (it ships 8 of the 24), a
different array, or a newer mask version than the one pinned here.

The v8.1 mask ships as `EPIC.hg38.mask.cm` — a bit-packed **CX-format** file rather
than a table. It stores one bit per probe per mask term over the array's canonical
probe ordering, which is why the whole thing is only ~360 kB where the equivalent
TSV is 2.8 MB. Because the file carries no probe names (row *i* means whatever row
*i* of the ordering means), you fetch the ordering alongside it.

Read it with [YAME](https://github.com/zhou-lab/YAME), the Zhou lab's encoder for
these formats:

```bash
# install: conda install -c zhou-lab -c conda-forge yame
# (bioconda's linux build is pinned at 1.8, which predates `yame fetch` --
#  use the zhou-lab channel or build from source)

yame fetch EPIC/EPIC.hg38.mask.cm EPIC/EPIC.ordering.tsv.gz

yame info    EPIC.hg38.mask.cm   # 24 mask terms over 866,553 probes
yame summary EPIC.hg38.mask.cm   # how many probes each term flags
yame unpack -a EPIC.hg38.mask.cm > mask_bits.tsv   # all 24 terms, one column each
```

`yame summary` is the quickest way to see what a term includes you before you commit
to it. Turning the bits into a probe list is then a row-wise OR across the columns
you chose, indexed against the ordering:

You do not have to run any of that to follow along: the three files those commands
produce or fetch are committed under `data/`, which is what the code below reads.

```{.r filename="RStudio Console"}
library(data.table)
# `yame unpack` writes the term names to a sidecar .idx, one per line, in the
# same order as the columns of mask_bits.tsv -- so read the column, not the header.
terms  <- fread("data/EPIC.hg38.mask.cm.idx", header = FALSE)$V1
probes <- fread("data/EPIC.ordering.tsv.gz")$Probe_ID     # canonical probe ordering
bits   <- fread("data/mask_bits.tsv.gz", header = FALSE); setnames(bits, terms)

# Choose the terms to mask on. These eight are the ones this chapter uses, and
# the same eight the ready-made TSV carries.
mask_cols <- c("M_mapping", "M_nonuniq",
               "M_SNP_AFR_1pt", "M_1baseSwitchSNP_AFR_1pt", "M_2extBase_SNP_AFR_1pt",
               "M_SNP_EUR_1pt", "M_1baseSwitchSNP_EUR_1pt", "M_2extBase_SNP_EUR_1pt")

drop <- probes[Reduce(`|`, lapply(mask_cols, function(m) bits[[m]] == 1L))]
```

This is where the five-population table in [§2](#sec-ancestry) comes from, and it is
the concrete reason the chapter needs both forms of the mask. The TSV carries
`M_SNP_AFR_1pt` and `M_SNP_EUR_1pt`, so the AFR-versus-EUR comparisons in that section
come straight from it — but it does not carry the AMR, SAS or EAS equivalents, so three
of the five rows can only come from the full release. Because the unpacked bits are committed, the count runs here rather
than being quoted at you:

```{r}
#| label: snp-by-pop
# Both inputs are the committed output of the `yame unpack -a` call above, so this
# runs without the yame binary. The bit matrix is 866,553 x 24 of 0/1, which gzips
# from 40 MB down to 0.5 MB -- small enough to keep in the repository.
terms  <- fread("data/EPIC.hg38.mask.cm.idx", header = FALSE)$V1
bits   <- fread("data/mask_bits.tsv.gz", header = FALSE); setnames(bits, terms)

# M_SNP_<POP>_1pt flags a common SNP (MAF >= 1%) at the interrogated CpG in that
# 1000 Genomes super-population. Counted over the whole array, not over this
# study's probes, because it is a property of the platform.
pops <- c("AFR", "AMR", "SAS", "EUR", "EAS")
snp_by_pop <- vapply(pops,
                     function(p) sum(bits[[paste0("M_SNP_", p, "_1pt")]] == 1L),
                     integer(1))

# Five rows of two columns, committed to the repository so the §2 table is
# readable straight from git without unpacking the mask first.
write.table(data.frame(population = pops, n_probes = snp_by_pop),
            "data/03_snp_by_pop.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

::: {.callout-note}
## The archived manifest is still there when you need it
`EPIC.hg38.manifest.pop.tsv.gz` remains available and is the right choice when you
need **sub-population** resolution: it covers 26 sub-populations (GWD, YRI, ASW, CEU,
FIN, …) as well as the 5 super-populations, where v8.1's `.cm` stops at the 5. The two
generations do not mask the same categories either — [§4](#sec-apply) quantifies the
difference and shows what it costs this cohort.
:::

## 2. Ancestry matters {#sec-ancestry}

The SNP mask is **not the same for every study population**, because a "common
SNP" depends on which population you ask about. The number of SNP-affected probes varies substantially by ancestry:

```{r}
#| label: pop-snp-table
# The five-row table written by the `snp-by-pop` chunk above. Read here rather
# than recounted, so this section and the funnel quote one source.
sp <- fread("data/03_snp_by_pop.tsv")
snp_by_pop <- setNames(sp$n_probes, sp$population)

snp <- data.frame(
  population  = c("AFR (African)", "AMR (Admixed American)", "SAS (South Asian)",
                  "EUR (European)", "EAS (East Asian)"),
  snp_masked  = snp_by_pop[c("AFR", "AMR", "SAS", "EUR", "EAS")]
)
knitr::kable(snp, row.names = FALSE,
             caption = "Probes flagged by M_SNP_<POP>_1pt (common SNP at the interrogated CpG, MAF >= 1%) by 1000 Genomes super-population. African-ancestry samples have roughly twice as many SNP-affected probes as European.")
```

African-ancestry populations carry the most common genetic variation, so more
probes overlap a common SNP. Concretely, in these masks
**`r format(snp_by_pop[["AFR"]], big.mark = ",")`** probes are SNP-flagged for
AFR versus **`r format(snp_by_pop[["EUR"]], big.mark = ",")`** for EUR:

```{r}
#| label: afr-eur-overlap
# One probe list per mask code, built straight from the shipped table's columns.
# The filtering section below reuses these, and chapter-level saves carry them.
mask_cols <- c("M_mapping", "M_nonuniq",
               "M_SNP_AFR_1pt", "M_1baseSwitchSNP_AFR_1pt", "M_2extBase_SNP_AFR_1pt",
               "M_SNP_EUR_1pt", "M_1baseSwitchSNP_EUR_1pt", "M_2extBase_SNP_EUR_1pt")
per_code_sets <- lapply(mask_cols, function(m) mask$Probe_ID[mask[[m]] == 1L])
names(per_code_sets) <- mask_cols
mp <- list(per_code_sets = per_code_sets)

# The comparison is between two mask terms over the same probe universe.
afr <- per_code_sets$M_SNP_AFR_1pt
eur <- per_code_sets$M_SNP_EUR_1pt
cat("masked in AFR but NOT EUR:", length(setdiff(afr, eur)), "\n")
cat("masked in EUR but NOT AFR:", length(setdiff(eur, afr)), "\n")
cat("masked in both:           ", length(intersect(afr, eur)), "\n")
```

While there is some overlap between the AFR and EUR recommended SNP masks, there are distinct maskings for each
population.

The Grady Trauma Project subset is **~95% African American**. If we filtered with 
a general or only European-population mask (a common default), we would leave in 
probes that are SNP-affected in African-ancestry individuals. As African American 
individuals tend to have an admixture of both European and African Ancestry, it is 
important to include masking recommendations for both.

::: {.callout-important}
## The teaching point
Probe filtering is not a fixed recipe — the SNP component depends on your study
population. For a cohort of known, relatively homogeneous ancestry, use that
population's mask. For an admixed or mixed-ancestry cohort, the conservative choice
is to take the **union** of the relevant population masks (drop a probe if it is
SNP-affected in *any* represented ancestry). For a multi-ancestry cohort or cohort 
of unknown ancestry, the general population mask is the most appropriate.
:::

## 3. Detection-p and sex-chromosome filtering {#sec-detp-sex}

Two study-specific filters complete the funnel:

- **Detection-p failures.** From [QC](01_qc.qmd) we have the per-probe detection
  p-values. A probe that fails (p > 0.01) in more than ~10% of samples is
  unreliable *in this dataset* and is dropped.
- **Sex chromosomes.** For an autosomal EWAS, X- and Y-chromosome probes are also
  removed: X-inactivation and sex differences would
  otherwise dominate, and they need sex-stratified handling if analyzed. They are
  *removed from the autosomal analysis*, not deleted from the study — a sex-specific
  analysis would use them deliberately.

The funnel in [§4](#sec-apply) reports how many probes each of these removed.

## 4. Applying the filters {#sec-apply}

Everything above is a decision about *which* probes to drop and why. This section
carries it out: the mask from [§1](#sec-why) and [§2](#sec-ancestry), the two
study-specific filters from [§3](#sec-detp-sex), and the funnel that accounts for
all of them.

::: {.callout-note}
## What the detection-p filter needs
This one step needs the raw two-channel object, `data/01_RGset.rds`, written by
[chapter 00](00_setup.qmd) — the QC chapter kept only per-sample summaries, and the
filter is per probe. If you did not run chapter 00, fetch the published checkpoint:

```{.bash filename="Terminal"}
./get_data.sh B_qc
```
:::

```{r}
#| label: probe-detp
# The detection-p filter is per PROBE, so it needs the full probe x sample
# p-value matrix. The QC chapter kept only per-sample summaries, so the matrix is
# recomputed from the raw two-channel object here: about 40 seconds and ~4 GB for
# 96 arrays, which is the most expensive step in this chapter.
rg   <- readRDS("data/01_RGset.rds")
detP <- detectionP(rg)

# Fraction of samples in which each probe failed the p > 0.01 cut, over the
# probes that survived normalization -- those are the only ones the funnel can
# drop. A probe unreliable in more than 10% of samples goes.
frac_fail <- rowMeans(detP[rownames(grs), ] > 0.01)
drop_detp <- rownames(grs)[frac_fail > 0.10]

# Nothing below needs the raw object or the p-value matrix, and together they are
# the memory high-water mark of the chapter, so release them now.
rm(rg, detP); invisible(gc())
length(drop_detp)
```

```{r}
#| label: probe-filters
# The eight-code union from the TSV above, restricted to the probes actually present
# after normalization.
drop_mask <- intersect(rownames(grs), drop8)

# Sex-chromosome probes, from the manifest coordinates minfi attached to the
# object -- removed for an autosomal EWAS, as set out in §3.
drop_sex <- rownames(grs)[getAnnotation(grs)$chr %in% c("chrX", "chrY")]

# A probe is dropped if ANY of the three lists flags it, so the three counts
# below overlap and do not sum to the total dropped.
keep <- setdiff(rownames(grs), Reduce(union, list(drop_mask, drop_detp, drop_sex)))


# Every count the funnel, the tables and the prose below quote, in one place.
funl <- list(
  n0               = nrow(grs),                        # probes entering the funnel
  drop_mask        = length(drop_mask),
  drop_detp        = length(drop_detp),
  drop_sex         = length(drop_sex),
  retained         = length(keep),                     # EWAS-ready probes
  union8           = sum(mask$mask_v81_8code),         # array-wide, before any study filter
  snp_by_pop       = setNames(sp$n_probes, sp$population)
)
```

Broken down, the probe removal includes: 

```{r}
#| label: funnel-table
funnel <- data.frame(
  step = c("Start (normalized probes)",
           "− Zhou v8.1 mask (8 codes, AFR + EUR)",
           "− Detection-p failures (>10% of samples)",
           "− Sex chromosomes (autosomal EWAS)",
           "= Retained (EWAS-ready)"),
  probes_dropped = c(NA, funl$drop_mask, funl$drop_detp, funl$drop_sex, NA),
  probes_remaining = c(funl$n0, NA, NA, NA, funl$retained)
)
knitr::kable(funnel, caption = "Probe-filtering funnel on the 96-sample subset (AFR + EUR aware).")
cat(sprintf("Retained %d of %d probes (%.1f%%).\n",
            funl$retained, funl$n0, 100 * funl$retained / funl$n0))
```


```{r}
#| label: apply-filter
# `keep` is the set union of the three drop lists, taken out of the normalized
# probe set: the eight-code mask (`drop_mask`), the detection-p failures
# (`drop_detp`) and the sex chromosomes (`drop_sex`), all built above. Order does
# not matter -- a probe flagged by two filters is dropped once.
grs_filt <- grs[keep, ]

# Subsetting a GenomicRatioSet carries the beta/M matrices, the genome
# coordinates and the sample metadata along with it, so this object is
# analysis-ready as it stands.
grs_filt
```

### Sub-population masks

If you need more specific sub-population masking, there is an archived version of the
recommended population masking (`EPIC.hg38.manifest.pop.tsv.gz`) that includes the 5
super-populations and 26 sub-populations. If using the archived population masking
file, the `MASK_general_<POP>` column includes the recommended problematic probes as
well as the population-specific probes. The `MASK_snp5_<POP>` columns only flag the
population-specific masking.

The two generations do not mask the same things, and the legacy pair masks more. The
reason is not the one the column names suggest, so it is worth deriving rather than
asserting.

The legacy equivalent of the eight v8.1 codes is `MASK_general_AFR` — problematic probes
*plus* AFR population SNPs — together with `MASK_snp5_EUR`, the EUR SNP column on its own
so the general terms are not counted twice. The archived manifest is 14 MB, so the
tutorial ships only its flagged probe IDs:

```{.r filename="RStudio Console"}
# How data/03_legacy_pop_mask.tsv.gz was built from the archived manifest.
library(data.table)
pop  <- fread("EPIC.hg38.manifest.pop.tsv.gz",
              select = c("probeID", "MASK_general_AFR", "MASK_snp5_AFR", "MASK_snp5_EUR"))
long <- rbindlist(lapply(c("MASK_general_AFR", "MASK_snp5_AFR", "MASK_snp5_EUR"),
  function(cl) data.table(probeID = pop$probeID[pop[[cl]]], mask = cl)))
fwrite(long[order(mask, probeID)], "data/03_legacy_pop_mask.tsv.gz", sep = "\t")
```

```{r}
#| label: mask-generation-contrast
# Long form: one row per (probe, legacy mask column) pair, flagged probes only --
# which is how an 866k-probe annotation fits in 0.8 MB.
leg <- fread("data/03_legacy_pop_mask.tsv.gz")
L   <- split(leg$probeID, leg$mask)

# The legacy mask applied to this study, with detection-p and sex held fixed so the
# mask generation is the only thing that differs from the funnel in §4.
legacy_union <- union(L$MASK_general_AFR, L$MASK_snp5_EUR)
leg_masked   <- intersect(rownames(grs), legacy_union)
leg_keep     <- setdiff(rownames(grs), Reduce(union, list(leg_masked, drop_detp, drop_sex)))

gen <- data.frame(
  mask = c("Legacy MASK_general_AFR + MASK_snp5_EUR", "v8.1, 8 codes (AFR + EUR)"),
  probes_masked   = c(length(leg_masked), funl$drop_mask),
  probes_retained = c(length(leg_keep),   funl$retained)
)
knitr::kable(gen, row.names = FALSE, format.args = list(big.mark = ","),
             caption = "Same cohort, same detection-p and sex filters; only the mask generation differs.")
```

Where the gap comes from is worth one more table, because two plausible explanations are
both wrong. Split each generation into its population-SNP terms and everything else:

```{r}
#| label: mask-generation-decomp
# Both generations mask SNPs at >1% allele frequency, so these rows are comparable:
# the `5` in the legacy `snp5` name is a 5 bp window around the target, not a 5%
# frequency cut-off [@zhou2017comprehensive]. Everything is restricted to the probes
# the v8.1 annotation covers; the archived manifest covers 635 fewer.
v81_snp  <- Reduce(union, mp$per_code_sets[c("M_SNP_AFR_1pt", "M_1baseSwitchSNP_AFR_1pt",
              "M_2extBase_SNP_AFR_1pt", "M_SNP_EUR_1pt", "M_1baseSwitchSNP_EUR_1pt",
              "M_2extBase_SNP_EUR_1pt")])
v81_rest <- union(mp$per_code_sets$M_mapping, mp$per_code_sets$M_nonuniq)
leg_snp  <- intersect(union(L$MASK_snp5_AFR, L$MASK_snp5_EUR), mask$Probe_ID)
leg_rest <- intersect(setdiff(L$MASK_general_AFR, L$MASK_snp5_AFR), mask$Probe_ID)

knitr::kable(data.frame(
  component = c("population SNP terms (>1% allele frequency)", "everything else"),
  legacy    = c(length(leg_snp),  length(leg_rest)),
  v8.1      = c(length(v81_snp),  length(v81_rest))),
  row.names = FALSE, format.args = list(big.mark = ","),
  caption = "Array-wide, not restricted to this study's probes.")

cat("legacy 'everything else' probes also flagged by v8.1 mapping/nonuniq:",
    sprintf("%.0f%%", 100 * length(intersect(leg_rest, v81_rest)) / length(leg_rest)), "\n")
cat("...flagged by none of the eight v8.1 codes:",
    format(length(setdiff(leg_rest, Reduce(union, mp$per_code_sets))), big.mark = ","), "\n")
```

So it is **not** the allele-frequency threshold — both generations use >1% — and the SNP
components come out within about four thousand probes of each other. What differs there is
*where* a SNP counts: the legacy columns flag any qualifying SNP within 5 bp of the target,
while v8.1 flags specific positions (the interrogated CpG, a 1-base colour-channel switch,
the two extension bases).

Nor is it that v8.1 abandoned the repeat and copy-number criteria. Almost the whole gap is
the non-SNP content `MASK_general_<POP>` bundles in — mapping quality, repeat overlap and
copy number of the 3′ subsequence. Around 40% of those probes are still flagged by v8.1's
mapping terms, and the remainder are no longer masked on that basis, which is the original
paper's own conclusion: 95.1% of probes overlapping repeats map uniquely to the genome
[@zhou2017comprehensive]. v8.1 does publish an `M_general` term, but it is not one of the
eight codes this chapter uses.

Moving to v8.1 therefore retains
**`r format(funl$retained - length(leg_keep), big.mark = ",")`** more probes in this
cohort — more sites tested, at the cost of the repeat-overlapping probes the legacy mask
excluded. Neither choice is automatically right; what matters is that you state which mask
generation and which terms you used, since the two are not interchangeable.

::: {.callout-note}
## Smoking-associated probes
Everything above masks probes that are unreliable *by design*. There is a separate,
optional reason to drop probes: smoking is one of the largest environmental drivers
of blood methylation, and if your dataset has no smoking variable — as the Grady
deposit does not — one of the available strategies is to exclude CpGs known from the
literature to be smoking-associated, at this filtering stage rather than after the
fact. That is a decision about confounding, not about probe quality, so it does not
belong in the funnel above. This tutorial does exclude smoking-associated CpGs, but
at a later step as these probes are needed to create a proxy smoking variable. See
[the smoking section in the EWAS chapter](06_ewas.qmd) for the full set of options.
:::

## 5. When you don't have IDATs: filtering a processed matrix {#sec-no-idats}

Probe filtering is one of the few preprocessing steps that works **identically**
whether your data came from IDATs or from a processed β-matrix — because the masks
are keyed on **probe IDs**, not intensities. The workflow:

1. **Load the matrix** (see the [normalization](02_normalization.qmd) non-IDAT
   section) so you have a β-matrix with `cg`/`ch` probe IDs as rownames.
2. **Apply the same masks by ID.** Read Zhou's mask/manifest for the correct array
   (450K vs EPIC vs EPICv2 — they have different probe sets), pick the mask for
   your study population, and subset. If you came straight to this section, run the
   `mask-read` chunk in [§1](#sec-why) first — that is where `drop8` is built:

   ```{.r filename="RStudio Console"}
   # `drop8` is the eight-code union built by the `mask-read` chunk in §1. It works
   # unchanged on a matrix, because the mask is keyed on probe IDs and nothing else.
   beta_filt <- beta[!rownames(beta) %in% drop8, ]

   # archived manifest, if you need sub-population resolution instead
   library(data.table)
   pop  <- fread("EPIC.hg38.manifest.pop.tsv.gz")
   drop <- pop$probeID[pop$MASK_general_ASW]           # e.g. African-ancestry SW US
   beta_filt <- beta[!rownames(beta) %in% drop, ]
   ```
3. **You lose the detection-p filter** unless the submitter deposited detection
   p-values alongside the matrix (some do, as a separate supplementary file — use
   it if present). Without it, rely on the mask plus removing probes with excessive
   `NA` values.
4. **Sex chromosomes** filter the same way, by intersecting with the manifest's
   chromosome annotation. It is possible sex chromosome CpGs will have already been 
   removed from a pre-processed dataset. 

::: {.callout-tip}
## Match the mask to the array *and* the genome build
Zhou publishes masks per platform (450K / EPIC / EPICv2) and per genome build
(hg19 / hg38). Use the manifest that matches the array the data were generated on;
the probe IDs mostly overlap between 450K and EPIC but the mask *contents* differ.
If you are combining a 450K series with an EPIC series, filter each on its own
manifest, then intersect on the shared probe IDs.
:::

We save the filtered, EWAS-ready object for cell-composition estimation next:

```{r}
#| label: save-filtered
# 1. The EWAS-ready GenomicRatioSet. An S4 object with four parallel matrices and
#    a GRanges, so RDS rather than a flat file. Read by cell-composition
#    estimation ([chapter 04](04_cell_composition.qmd)) and the batch-effect
#    diagnostics ([chapter 05](05_batch_effects.qmd)).
saveRDS(grs_filt, "data/03_grs_filtered.rds")

# 2. The funnel counts, so a later chapter or a methods section can quote them
#    without re-running the filter.
saveRDS(funl, "data/03_filter_funnel.rds")

# 3. The per-code probe sets built in §2. A list of eight character vectors of
#    unequal length, which a flat file would not hold.
saveRDS(mp, "data/03_mask_pieces.rds")

# 4. The funnel as a flat table as well: five rows, so it is committed to the
#    repository and readable straight from GitHub by anyone who cannot hold the
#    GenomicRatioSet in memory.
write.table(funnel, "data/03_filter_funnel.tsv",
            sep = "\t", row.names = FALSE, quote = FALSE, na = "")
```

::: {.callout-note collapse="true"}
## If the detection-p step will not fit on your machine
Recomputing `detectionP()` holds the raw two-channel object and a probe × sample
p-value matrix in memory at once — roughly 4 GB on top of the normalized object. If
that is more than you have, fetch the filtered checkpoint this chapter writes and
start from it instead:

```{.bash filename="Terminal"}
./get_data.sh D_filtered
```

```{.r filename="RStudio Console"}
grs_filt <- readRDS("data/03_grs_filtered.rds")
```

What you give up is the funnel: every count above is computed from the three probe
lists, not read from a file, so none of it appears if you skip the chapter's code.
:::

The result: **`r format(funl$retained, big.mark=",")`** autosomal probes on
96 samples, filtered under the Zhou v8.1 mask for mapping quality, SNP masking
appropriate to an admixed African-American cohort (AFR and EUR terms together),
detection reliability, and sex chromosomes.

Nearly every probe that survives here is tested in the EWAS. Two small deductions
happen later, both in [chapter 05](05_batch_effects.qmd):

- the **20 CpGs used to build the smoking proxy** are excluded, because regressing
  those probes on a score derived from themselves is circular; and
- **two probes** (`cg17759086`, `cg01801182`) have β pinned at 0 or 1 in the
  87-sample analysis subset, so their M-values are ±∞ and they cannot be fitted.

So `r format(funl$retained, big.mark=",")` probes pass filtering and
**756,251** are actually tested.

---

**Next:** [Cell composition](04_cell_composition.qmd) — whole blood is a mixture of
cell types with very different methylomes; we estimate the proportions and validate
against the study's deposited values.

## References {.unnumbered}

::: {#refs}
:::
