---
title: "Genetic ancestry principal components"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

## What genetic PCs capture {#sec-what}

**Genetic ancestry** describes which populations a person's genome descends from. It is
a property of DNA, not a social label. Because populations that were separated for a
long time drifted apart in allele frequencies, an individual's genotypes at many
thousands of common variants carry a record of that descent. In a cohort drawn from
several continents the largest differences in allele frequency fall between continental
groups, so the leading axes of genetic variation broadly separate African, European,
East Asian, South Asian and American ancestries. Within one continental group, the same
approach picks up finer regional structure; within Europe, for example, the first two
PCs reproduce the map of the continent [@novembre2008geography]. **Admixture** is the case where a person's
genome descends from more than one population, as it does for most African Americans,
who carry both African and European ancestry in proportions that vary from person to
person [@bryc2015ancestry].

**Principal component analysis** (PCA) summarises a table of many correlated variables
with a few new variables, the principal components (PCs) [@jolliffe2016pca]. Each PC is a weighted sum of
the original variables, chosen so that the first PC captures as much of the variance in
the data as possible, the second captures the most of what remains while being
uncorrelated with the first, and so on. Applied to a genotype matrix (individuals ×
variants, coded 0/1/2 copies of one allele), each individual gets a score on each PC.
Two individuals with similar allele frequencies across the genome get similar scores,
so they sit close together when the scores are plotted; individuals from different
populations sit far apart, and admixed individuals fall between the groups their
ancestry comes from, roughly in proportion to how much of each they carry. The first
few PCs of genome-wide genotypes therefore summarise genetic ancestry in a handful of
continuous numbers [@patterson2006eigen]. Adjusting for those components is the standard correction
for population stratification in genetic association studies
[@price2006pca; @patterson2006eigen], and it carries over to EWAS. DNA methylation
differs with ancestry too. Barfield et al. found widespread associations between
methylation and race in 388 individuals and showed that, left unadjusted, this
population structure can confound a methylation study [@barfield2014stratification].
Part of the reason is genetic: sequence variants near a CpG can shift its methylation,
and allele frequencies differ between populations. Chapter 03's masks do not deal with
this. They remove probes whose *measurement* a SNP can corrupt, but a CpG whose
methylation genuinely differs with ancestry passes every mask. If ancestry is also
related to the exposure, it is a confounder, and even if it is not, it is structured
variation the model can explain.

From the [next chapter](05_batch_effects.qmd) on, every model in this tutorial carries
three genetic ancestry principal components (PCs) as covariates: in ComBat and surrogate
variable analysis in chapter 05, in the EWAS design in chapter 06, and in the pipeline's
phenotype file in chapter 07. They are added to account for underlying population
substructure: if methylation at a CpG differs with ancestry, and ancestry is also
unevenly distributed between cases and controls, the CpG would look associated with the
exposure when it is not. Carrying the PCs as covariates lets the model attribute that
variation to ancestry rather than to the exposure. This short chapter explains what the
PCs are, how they relate to self-reported race, and what to do when a dataset does not
include them.

::: {.callout-note}
## What this chapter needs
Everything up to the EPISTRUCTURE example needs only `data/sample_sheet.csv`, which is in
the repository; the three PCs are its last three columns, `pc1`, `pc2` and `pc3`. The
[EPISTRUCTURE example](#sec-epistructure) at the end also reads `data/02_funnorm_grs.rds`,
written at the end of [normalization](02_normalization.qmd), and
`data/04_cellcounts.rds`, written by [cell composition](04_cell_composition.qmd) and
committed to the repository.
:::

```{r}
#| label: load
library(ggplot2)   # the scatter plot below

ss   <- read.csv("data/sample_sheet.csv", stringsAsFactors = FALSE)
gpcs <- c("pc1", "pc2", "pc3")

# The 86 samples an association model can use: PTSD status, sex, age and all three
# PCs recorded. These are the same 86 samples chapter 05 onwards models; one of the
# 96 arrays has no PCs, and nine have no PTSD status.
keep <- !(is.na(ss$ptsd) | ss$ptsd == "") & !is.na(ss$sex) & !is.na(ss$age) &
        complete.cases(ss[, gpcs])
pcs  <- ss[keep, ]
cat("samples with PCs:", sum(complete.cases(ss[, gpcs])), "of", nrow(ss),
    "| modeled samples:", nrow(pcs), "\n")
```

## Genetic PCs and self-reported race {#sec-race}

Genetic PCs and self-reported race are different kinds of variable, and neither replaces
the other. Race is a social category: the labels available, and which one a person is
assigned or chooses, are set by the society they live in, not by their genome
[@nasem2023descriptors; @borrell2021race]. Genetic ancestry is a property of the genome,
and the PCs describe it continuously [@lewis2022ancestry]. The two are correlated in
many cohorts, because people who share a racial label often share ancestry, but the
correlation is not fixed, and it is often weak within a group: individuals recorded
under one label can differ widely in ancestry, most obviously in admixed populations
[@bryc2015ancestry]. Race also carries information genetic PCs cannot: it is the
variable through which the effects of systemic racism, socioeconomic position, and
other social and environmental exposures act on health, and those exposures shape DNA methylation
[@borrell2021race]. So a race covariate and genetic PCs adjust for different things, and
the 2023 National Academies report on population descriptors recommends using genetic
ancestry, not race, when the aim is to account for genetic population structure, and
naming which descriptor was used and why [@nasem2023descriptors].

For this tutorial's purpose, adjusting for population substructure, genetic PCs are the
appropriate variable: they measure the structure directly and continuously. Whether to
also include self-reported race depends on the question. In a study of a socially
patterned exposure, race may be an important covariate or even the exposure itself; in a
study of a molecular trait, genetic PCs alone may be the better-specified choice. The
Grady cohort also shows why race cannot stand in for the PCs here: almost every modeled
sample carries the same label, so a categorical race variable doesn't provide meaningful
differentiation for a statistical model, while the genetic PCs still vary from sample to
sample:

```{r}
#| label: genetic-pcs
table(race = ifelse(is.na(pcs$race) | pcs$race == "", "not recorded", pcs$race))

# Their spread within the modeled set: continuous, and centred near zero because
# they were computed on the whole cohort.
round(sapply(pcs[, gpcs], summary), 4)
```

```{r}
#| label: genetic-pc-scatter
#| fig-width: 6.4
#| fig-height: 4.6
#| fig-cap: "The first two genetic ancestry PCs for the 86 modeled samples, coloured by self-reported race. Samples recorded with the same race label still spread along both components: this is the genetic variation a categorical race variable cannot represent, and the reason the PCs are used to adjust for population substructure."
ggplot(pcs, aes(pc1, pc2, colour = ifelse(is.na(race) | race == "", "not recorded", race))) +
  geom_point(size = 2.2, alpha = 0.8) +
  scale_colour_manual(values = c(AA = "#1A6B75", Other = "#8C3A4A",
                                 `not recorded` = "#8A8279"),
                      name = "Self-reported race") +
  labs(x = "Genetic PC1", y = "Genetic PC2")
```

The heatmap in [§2 of Batch effects](05_batch_effects.qmd#sec-pca) includes the three
components, so you can see how much of the methylation variance they account for
before any model is fit.

## How many components {#sec-how-many}

How many components a cohort needs depends on how structured it is. The usual way to
decide is a scree plot: plot the variance each component explains against its rank and
keep the components before the curve flattens, the "elbow" [@cattell1966scree].
EIGENSOFT's `smartpca` also tests each component formally against a Tracy–Widom
distribution [@patterson2006eigen]. Both need the full set of eigenvalues, which a
deposit that ships only the first three PCs, as Grady's does, cannot give you. Three is
a common choice, and it is what the PGC-PTSD meta-analysis used for genotyped cohorts
[@katrinli2024ptsd]. In practice, most PCs beyond the first two or three within a
continental group have little association with methylation and adjusting for them
mainly costs degrees of freedom.

::: {.callout-tip collapse="true"}
## If your GEO deposit has no genetic PCs

Most GEO methylation deposits do not include genetic PCs, so you will often have to
make your own. In order of preference:

1. **Compute them from genotypes.** If genotype data exist for the same individuals,
   run a PCA on them. For a public methylation deposit this usually means the genotypes
   are under controlled access (for example on dbGaP), which requires an approved data
   access request from an institutional signing official, so it is rarely an option a
   student can arrange alone; ask whether your PI already has access. Remove rare
   variants and prune SNPs in linkage disequilibrium first, so that the components
   describe genome-wide ancestry rather than a few long haplotypes.
   PLINK 2 [@chang2015plink2] and EIGENSOFT's `smartpca`
   [@patterson2006eigen; @price2006pca] are the standard tools:

   ```{.bash filename="Generic example, do not run"}
   # `mydata` is a placeholder for a PLINK fileset (.bed/.bim/.fam) of your cohort.
   plink2 --bfile mydata --autosome --maf 0.05 \
          --indep-pairwise 200 50 0.2 --out pruned
   plink2 --bfile mydata --extract pruned.prune.in --pca 10 --out mydata_pcs
   # mydata_pcs.eigenvec: one row per individual, PC1-PC10 as columns
   ```

2. **Estimate ancestry from the methylation data.** Without genotypes, PCs can be
   computed from the methylation array itself. Barfield et al. showed that PCs built
   from CpGs chosen because they proxy nearby SNPs adjust for population stratification
   nearly as well as genotype PCs [@barfield2014stratification]. EPISTRUCTURE, part of
   the GLINT toolset, implements a related approach [@rahmani2017epistructure;
   @rahmani2017glint]; the [example below](#sec-epistructure) runs it on this cohort. The
   PGC-PTSD meta-analysis used Barfield's method for its cohorts without genotypes and
   kept PCs 2 and 3, the components most correlated with self-reported ancestry
   [@katrinli2024ptsd]. Because methylation-based components can also pick up cell
   composition or batch, check what they correlate with before you use them. Comparing
   them with self-reported race and with chapter 04's cell proportions is our
   suggestion for a minimum check.

3. **Fall back on self-reported race or ethnicity.** If neither is possible, a
   categorical self-reported race or ethnicity covariate captures some of the
   population structure where the labels and ancestry are correlated, but it is a
   different kind of variable ([above](#sec-race)): it cannot represent admixture, and
   in a cohort where most samples share one label it adjusts for very little. Report
   it as a limitation and say which descriptor you used.
:::

## Worked example: EPISTRUCTURE on the Grady data {#sec-epistructure}

Because Grady deposits genetic PCs, it lets us check how well a methylation-only method
recovers them. EPISTRUCTURE [@rahmani2017epistructure] works from a fixed list of 4,913
CpGs whose methylation was found to be strongly explained by nearby SNPs. It removes the
effect of cell composition from those sites and takes principal components of what is
left. GLINT, the toolset that implements it, is written for Python 2.7
[@rahmani2017glint], so the steps are written out in R below. Run on this dataset,
GLINT's own `--epi` gives the same three components as this code: they correlate at
r = −1 (the sign of a PC is arbitrary) and differ in scale only because scikit-learn
standardizes with *n* rather than *n* − 1.

Two things about the input matter. First, the reference CpGs are exactly the kind of
SNP-influenced sites [probe filtering](03_probe_filtering.qmd) removes, so EPISTRUCTURE
has to run on the normalized but **unfiltered** object; GLINT's documentation makes the
same recommendation. Second, the list was selected in European individuals, and this
cohort is mostly African American, so how well it works here is a real question rather
than a foregone conclusion.

```{r}
#| label: epistructure
library(minfi)   # getBeta()

# GLINT's reference list, read from the GLINT repository at a fixed commit so it
# keeps returning this exact file.
ref <- readLines(paste0("https://raw.githubusercontent.com/cozygene/glint/",
                        "31a138ce380faeaf0e698733e781771df33374fe/",
                        "parsers/assets/epistructure_reference_sites.txt"))

# Normalized, unfiltered beta values for the reference sites that are on the EPIC
# array (the list was built on the 450K array, so a few are missing).
grs  <- readRDS("data/02_funnorm_grs.rds")
beta <- getBeta(grs[rownames(grs) %in% ref, ])
rm(grs)   # 1.25 GB; only the reference sites are needed from here on
cat("reference CpGs:", length(ref), "| on this array:", nrow(beta), "\n")

# Covariates to remove before the PCA, in the same sample order as `beta`.
cells <- readRDS("data/04_cellcounts.rds")[colnames(beta), ]
sheet <- ss[match(colnames(beta), ss$sample_id), ]

# EPISTRUCTURE: regress each reference site on the covariates, keep the residuals,
# standardize each site, and take principal components across samples.
epistructure <- function(beta, covariates, n_pcs = 3) {
  res <- t(lm.fit(covariates, t(beta))$residuals)
  prcomp(t(res), center = TRUE, scale. = TRUE)$x[, seq_len(n_pcs)]
}

# GLINT's documented usage regresses out cell composition only. On these arrays the
# components that gives also track the chip each sample was run on, so the version
# used below also removes sex and chip.
epi_cells <- epistructure(beta, cbind(1, cells))
epi       <- epistructure(beta, model.matrix(~ cells + sex + factor(slide), sheet))

# How much each version's first component tracks chip, and how closely it follows
# the deposited pc1 (95 samples; one has no deposited PCs).
has_pcs <- complete.cases(sheet[, gpcs])
chip_r2 <- function(e) summary(lm(e[, 1] ~ factor(sheet$slide)))$r.squared
data.frame(covariates_removed = c("cell types", "cell types, sex, chip"),
           chip_R2_of_EPI1    = round(c(chip_r2(epi_cells), chip_r2(epi)), 2),
           abs_r_with_pc1     = round(abs(c(cor(epi_cells[has_pcs, 1], sheet$pc1[has_pcs]),
                                            cor(epi[has_pcs, 1], sheet$pc1[has_pcs]))), 2))
```

```{r}
#| label: epistructure-compare
# Correlation of each deposited genetic PC with each EPISTRUCTURE component, and
# how much of each deposited PC the three components explain together.
epi_cor <- cor(sheet[has_pcs, gpcs], epi[has_pcs, ])
colnames(epi_cor) <- paste0("EPI", 1:3)
epi_r2 <- sapply(gpcs, function(p) summary(lm(sheet[has_pcs, p] ~ epi[has_pcs, ]))$r.squared)
round(cbind(epi_cor, R2_from_EPI1_3 = epi_r2), 2)
```

```{r}
#| label: fig-epistructure
#| fig-width: 6.4
#| fig-height: 4.6
#| fig-cap: "The first EPISTRUCTURE component against the deposited genetic PC1, for the 95 samples with deposited PCs, coloured by self-reported race. The sign of a principal component is arbitrary, so a negative correlation is as good as a positive one."
cmp <- data.frame(pc1 = sheet$pc1[has_pcs], EPI1 = epi[has_pcs, 1],
                  race = ifelse(is.na(sheet$race[has_pcs]) | sheet$race[has_pcs] == "",
                                "not recorded", sheet$race[has_pcs]))
ggplot(cmp, aes(pc1, EPI1, colour = race)) +
  geom_point(size = 2.2, alpha = 0.8) +
  scale_colour_manual(values = c(AA = "#1A6B75", Other = "#8C3A4A",
                                 `not recorded` = "#8A8279"),
                      name = "Self-reported race") +
  labs(x = "Deposited genetic PC1", y = "EPISTRUCTURE component 1")
```

EPISTRUCTURE recovers the main ancestry axis reasonably well: its first component
correlates with the deposited PC1 at |r| = `r sprintf("%.2f", abs(epi_cor["pc1", "EPI1"]))`,
and the three components together explain
`r sprintf("%.0f%%", 100 * epi_r2["pc1"])` of PC1's variance. EPISTRUCTURE's components explain almost none of PC2
(`r sprintf("%.0f%%", 100 * epi_r2["pc2"])`) or PC3
(`r sprintf("%.0f%%", 100 * epi_r2["pc3"])`) in this cohort. The table printed by the `epistructure` chunk also shows why it is worth checking what
methylation-based components correlate with before using them: with only cell types
removed, the first component partly tracks the chip.

So EPISTRUCTURE is a reasonable substitute for genetic PCs if they are not provided. The rest of this
tutorial uses the deposited genetic PCs, not these.

::: {.callout-note}
## If you do use EPISTRUCTURE components as covariates
EPISTRUCTURE doesn't require removing the reference sites from the association test. A
conservative option is to exclude the ones still left after probe filtering, or to note
that their association estimates were adjusted for a covariate built partly from their
own methylation. Either way, report which option you chose.
:::

---

**Next:** [Batch effects](05_batch_effects.qmd) — technical variation from chips,
array positions, and plates, and how to detect it before deciding whether to
correct.

## References {.unnumbered}

::: {#refs}
:::
