---
title: "Batch effects"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

Every methylation array is run somewhere on a physical chip, at some position on
that chip, on some plate, on some day. These **technical** factors leave systematic
fingerprints in the data, known as "batch effects". If a batch factor happens to line up
with your exposure of interest, it manufactures false positives. Even when it
doesn't, it inflates noise and costs power. This notebook shows how to **detect and
quantify** batch structure with PCA, how to tell a batch effect apart from a
biological one, and — importantly — how to decide *whether and how* to correct.


::: {.callout-note}
## What this chapter needs
This chapter starts from `data/03_grs_filtered.rds`, written by [chapter 03](03_probe_filtering.qmd). If you
did not run that chapter, fetch the published checkpoint instead:

```{.bash filename="Terminal"}
./get_data.sh D_filtered
```
:::

```{r}
#| label: load
library(minfi)   # getBeta() on the filtered GenomicRatioSet
library(sva)     # ComBat(), num.sv(), sva()
library(limma)   # lmFit()/eBayes(), used in the k-selection sweep
library(ggplot2) # every figure in this chapter

# num.sv()'s permutation test and sva()'s iterative reweighting are both
# stochastic, so the seed is what makes the surrogate variables -- and every
# number derived from them below -- reproducible from one render to the next.
set.seed(42)

# The probe-filtered, funnorm-normalized object written by chapter 03:
# 756,273 EWAS-ready CpGs x 96 QC-passing arrays.
grs <- readRDS("data/03_grs_filtered.rds")

# The reference-based cell proportions written by chapter 04 (96 x 6). This one
# is small enough to live in the repository, so it needs no download.
props <- readRDS("data/04_cc_full.rds")$prop

# The sample sheet rides along on the object, so the phenotype table comes back
# out of it rather than being read from the CSV again.
md <- as.data.frame(pData(grs))

# Everything below indexes samples three different ways -- array columns, rows
# of `md`, rows of `props` -- so align them once, here, and assert it. A silent
# mis-alignment would produce plausible-looking nonsense in every figure.
props <- props[md$sample_id, , drop = FALSE]
stopifnot(identical(colnames(grs), md$sample_id),
          identical(rownames(props), md$sample_id))

# The 87 samples an association model can actually use: PTSD status is NA for
# nine arrays (see chapter 01), and a model containing the exposure drops them.
# The PCA diagnostic in section 2 deliberately keeps all 96.
keep  <- !(is.na(md$ptsd) | md$ptsd == "") & !is.na(md$sex) &
         !is.na(md$age) & complete.cases(props)
mdk   <- md[keep, ]
propk <- props[keep, , drop = FALSE]

# Control is the reference level: the PTSD coefficient is then "Case relative to
# Control", which is the direction the whole tutorial reports.
mdk$ptsd <- factor(mdk$ptsd, levels = c("Control", "Case"))

cat("arrays:", ncol(grs), "| probes:", nrow(grs),
    "| arrays with modelable phenotype:", sum(keep), "\n")
```

::: {.callout-note}
## Where 96 becomes 87

Two sample counts appear in this chapter and it is worth being clear about which is
which.

The **PCA diagnostics in §2** run on all **96** QC-passing arrays. They describe
technical structure in the arrays and take no model, so the exposure never enters and
there is no reason to exclude anything.

**Batch effect adjustment runs on 87** — the stratified ComBat,
and SVA. For each of these analyses, it's necessary to provide which variables should
be protected which includes the phenotype of interest. There are nine samples with no recorded PTSD status,
therefore these are excluded.

The **Smoking proxy** is also run on 87 samples, but for the reason of capturing the variance of the sample set that the EWAS will actually be performed on.
:::

## 1. The sources of technical variation on an array {#sec-sources}

The usual suspects, in rough order of how often they matter:

- **Chip / BeadChip (Sentrix ID).** Eight samples share one EPIC chip. Samples on
  the same chip are processed together and tend to be more similar to each other
  than to samples on other chips — the strongest and most common array batch factor.
- **Array position (Sentrix position).** Where a sample sits on the chip. The EPIC v1
  BeadChip is an **8 × 1** grid, so the positions are `R01C01` through `R08C01` and
  there are exactly **eight** of them. Position effects are subtler than chip effects but real and
  reproducible — edge rows in particular can differ systematically.
- **Plate.** Chips are processed in plates. A full plate holds 96 arrays — twelve
  chips of eight — and captures reagent lots, enzyme batches, and the handling
  session, all *upstream* of any individual chip.
- **Well.** The sample's position in the 96-well DNA plate, labeled `A01`–`H12`.
  Well captures pipetting variation, bisulfite-conversion efficiency, reagent
  gradients across the plate, and plate-edge effects (evaporation and thermal
  gradients hit the outer wells hardest). Well is *related to but distinct from*
  Sentrix position: under a standard plate map, well is essentially the
  **(chip, position) pair taken jointly**, so within a single plate, adjusting for
  chip and position largely covers well. Across plates they separate — two samples
  can both sit at `R03C01` on plates processed weeks apart.
- **Scan date / scanner.** When and on which machine the chip was imaged.

These are recorded in the IDAT filenames and sample sheet
(see [Setup](00_setup.qmd)). Here we have chip (`slide`) and position (`array_pos`);
the Grady deposit doesn't give explicit plate, well, or scan date.


## 2. Detecting structure with PCA {#sec-pca}

Principal component analysis is a commonly used method to assess for technical variation.
Each principal component is an axis of coordinated methylation variation
across samples; we then ask **what each PC is associated with** by regressing it on
every candidate variable — technical and biological alike — and reading off the
$R^2$. A PC that has a high $R^2$ with a technical factor is a batch effect.

```{r}
#| label: mvalues
# PCA, ComBat and SVA all work on M-values rather than betas: log2(beta/(1-beta))
# is unbounded and roughly homoscedastic, so a linear model is not fighting the
# floor and ceiling of the 0-1 beta scale.
#
# Cost: getBeta() materializes a 756,273 x 96 numeric matrix (~580 MB) and M is a
# second one, so this chunk peaks near 1.2 GB. `beta` is dropped as soon as the
# M-values exist; only the handful of panel probes needed in section 5 are read
# back off `grs` later.
beta <- getBeta(grs)
M    <- log2(beta / (1 - beta))
M[!is.finite(M)] <- NA          # beta exactly 0 or 1 gives -Inf/+Inf
rm(beta); invisible(gc())
cat("M-value matrix:", paste(dim(M), collapse = " x "),
    "| probes with a non-finite value:", sum(rowSums(is.na(M)) > 0), "\n")
```

```{r}
#| label: pca-compute
# Rank probes by variance and keep the top 20,000. PCA on all 756,273 probes
# would take minutes and give the same picture: the leading components are driven
# by the most variable probes, and this diagnostic only needs to identify what
# those components track. The SVA in section 6 uses every probe.
v    <- apply(M, 1, var, na.rm = TRUE)
top  <- names(sort(v, decreasing = TRUE))[1:20000]
Mtop <- M[top, ]
Mtop <- Mtop[complete.cases(Mtop), ]   # prcomp() cannot take an NA

# center but do not scale: every row is already on the same M-value scale, and
# scaling would give the least variable of the 20,000 probes equal say.
pr  <- prcomp(t(Mtop), center = TRUE, scale. = FALSE)
pcs <- pr$x[, 1:10]
pve <- (pr$sdev^2 / sum(pr$sdev^2))[1:10]   # fraction of variance per PC
cat("PCA input:", nrow(Mtop), "probes | PC1-PC3 variance:",
    paste(sprintf("%.1f%%", 100 * pve[1:3]), collapse = " "), "\n")
```

```{r}
#| label: pc-r2
# One row per candidate explanatory variable, technical and biological together
# -- the point of the diagnostic is that they compete on the same axes.
# Position is collapsed to its row (R01..R08 -> "R01"), because on an 8 x 1 EPIC
# chip the column is constant and adds nothing.
slide  <- factor(as.character(md$slide))
posrow <- substr(md$array_pos, 1, 3)
vars <- list(chip_slide = slide, array_position = factor(posrow),
             PTSD = factor(md$ptsd), sex = factor(md$sex), age = md$age,
             Neu = props[, "Neu"], CD4T = props[, "CD4T"], CD8T = props[, "CD8T"])

# R^2 of each PC regressed on each variable: a factor enters as its full set of
# dummies, so a many-level factor like chip is compared against the others with
# that caveat in mind (see take-away 2 below).
pc_r2 <- t(sapply(vars, function(x)
  sapply(1:10, function(j) summary(lm(pcs[, j] ~ x))$r.squared)))
colnames(pc_r2) <- paste0("PC", 1:10)
round(pc_r2, 2)
```

```{r}
#| label: pc-heatmap
#| fig-cap: "Association (R²) between the top 10 principal components and each candidate variable. Computed on the 20,000 most variable probes — a subset used here only to keep this diagnostic fast; the SVA below runs on the full filtered probe set. Cell-composition variables (Neu, CD8T) dominate PC1; age and CD4T explain PC2; sex and chip are entangled on PC5; chip smears across many PCs. PTSD — the exposure of interest — is essentially absent, the honest reality of a subtle epigenetic effect."
#| fig-width: 9.6
#| fig-height: 4.6
# Long form for geom_tile, with display labels instead of the variable names
# used in the matrix.
nice <- c(chip_slide = "Chip (slide)", array_position = "Array position",
          PTSD = "PTSD", sex = "Sex", age = "Age",
          Neu = "Neutrophil", CD4T = "CD4 T", CD8T = "CD8 T")
hm <- as.data.frame(as.table(pc_r2))
names(hm) <- c("variable", "PC", "r2")
# rev() so the first variable in `nice` ends up at the top of the y axis
hm$variable <- factor(nice[as.character(hm$variable)], levels = rev(unname(nice)))
hm$PC <- factor(hm$PC, levels = paste0("PC", 1:10))

p_hm <- ggplot(hm, aes(PC, variable, fill = r2)) +
  geom_tile(color = "white", linewidth = 0.7) +
  # white numbers on the dark tiles, dark on the pale ones
  geom_text(aes(label = sub("^0", "", sprintf("%.2f", r2)), color = r2 > 0.45),
            size = 2.9, show.legend = FALSE) +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "#3A362F")) +   # gray_dark
  scale_fill_gradient(low = "#F4F2EE", high = "#0F3D43", limits = c(0, 1),   # teal_dark
                      name = expression(R^2)) +
  # each PC's share of variance printed under its label, so the heatmap says how
  # much of the data each column actually represents
  scale_x_discrete(position = "top", labels = function(x)
    sprintf("%s\n%.1f%%", x, 100 * pve[match(x, paste0("PC", 1:10))])) +
  labs(x = NULL, y = NULL,
       title = "What the leading principal components are made of",
       subtitle = "Cell composition owns PC1; chip and sex are entangled on PC5; PTSD is nowhere",
       caption = "Percentages under each label are the variance that component explains.") +
  theme(panel.grid = element_blank(), panel.border = element_blank(),
        axis.text.x.top = element_text(color = "#6E675B", lineheight = 1.15))  # gray
p_hm

# The repository tracks a copy of this figure; rewrite it here so the file on
# disk is always the one the code above just produced.
ggsave("data/05_pc_heatmap.png", plot = p_hm, width = 9.6, height = 4.6, dpi = 200)
```

Key take-aways from the heatmap:

1. **The largest axis of variation is biology you already modeled — cell
   composition.** PC1 (`r sprintf("%.0f%%", 100*pve[1])` of variance) is almost entirely neutrophil fraction
   ($R^2$ = `r sprintf("%.2f", pc_r2["Neu","PC1"])`) and CD8T
   ($R^2$ = `r sprintf("%.2f", pc_r2["CD8T","PC1"])`). This is why
   [cell composition](04_cell_composition.qmd) came first: if you skip it, your
   biggest "batch effect" is actually unmodeled cell mixture.
2. **Chip is a real technical effect,** associating with several PCs (peaking at
   $R^2$ = `r sprintf("%.2f", max(pc_r2["chip_slide",]))`). Some of this magnitude is
   inflated because `slide` has many levels (11 degrees of freedom soak up variance
   mechanically), so we interpret chip effects with the adjusted $R^2$ and with the
   confounding check below, not the raw number alone.
3. **PTSD doesn't explain much.** The exposure explains almost none of the leading variance.
   This is not a cause for alarm. Real epigenetic
   effects of a psychiatric phenotype are small and spread across many probes, making them 
   practically invisible at the whole-genome PCA level. The association test in
   [analysis](06_ewas.qmd) is designed to find these signals.

```{r}
#| label: pc-neu-scatter
#| fig-cap: "Samples in PC1–PC2 space, colored by estimated neutrophil fraction. The leading component is a smooth gradient in cell composition — a textbook illustration that PC1 of a blood methylome is usually immune-cell mixture, not the phenotype under study."
#| fig-width: 7.4
#| fig-height: 5
sc <- data.frame(PC1 = pcs[, 1], PC2 = pcs[, 2], Neu = 100 * props[, "Neu"])

p_sc <- ggplot(sc, aes(PC1, PC2, color = Neu)) +
  geom_point(size = 2.7, alpha = 0.92) +
  # a single-hue ramp: neutrophil fraction is a quantity, not a category, so pale
  # to teal reads as "more" without implying groups
  scale_color_gradient(low = "#DCE9EA", high = "#0F3D43",              # teal_dark
                        name = "Neutrophil\nfraction",
                        labels = function(x) paste0(x, "%")) +
  # variance explained goes in the axis titles, so the scatter is readable alone
  labs(x = sprintf("PC1 (%.1f%% of variance)", 100 * pve[1]),
       y = sprintf("PC2 (%.1f%%)", 100 * pve[2]),
       title = "PC1 of a blood methylome is immune-cell mixture",
       subtitle = "Each point is one array, colored by its estimated neutrophil fraction")
p_sc

ggsave("data/05_pc_scatter_neu.png", plot = p_sc, width = 7.4, height = 5, dpi = 200)
```

## 3. The crucial step: is the batch factor confounded? {#sec-confounding}

Detecting a batch effect is not enough. Before choosing a correction you must ask
whether the technical factor is **confounded** with something you care about,
because that determines which corrections are even *possible*. The diagnostic is a
simple cross-tabulation (Cramér's V for two categorical variables):

```{r}
#| label: confound-check
# Cramer's V rescales a chi-squared statistic to [0, 1] so that two tables of
# different sizes are comparable: 0 is independence, 1 means one variable is a
# relabelling of the other. chisq.test() warns about small expected counts in a
# table this sparse, which is expected here and not informative, so it is silenced.
cramV <- function(a, b){
  t <- table(a, b); if (any(dim(t) < 2)) return(NA)
  chi <- suppressWarnings(chisq.test(t)$statistic)
  as.numeric(sqrt((chi / sum(t)) / (min(dim(t)) - 1)))
}
# All 96 arrays: this asks how the design was built, not how a model behaves.
conf <- data.frame(
  pair = c("chip × PTSD", "chip × sex", "position × PTSD", "position × sex"),
  cramers_V = round(c(
    cramV(md$slide, md$ptsd), cramV(md$slide, md$sex),
    cramV(substr(md$array_pos,1,3), md$ptsd),
    cramV(substr(md$array_pos,1,3), md$sex)), 2))
knitr::kable(conf, caption = "Confounding between technical factors and biological variables (Cramér's V; 0 = independent, 1 = perfectly confounded).")
```

This results of this can greatly impact how you proceed with correcting for batch effects.
In this case:

- **chip × PTSD = `r round(cramV(md$slide, md$ptsd),2)`** — low. The exposure is
  *well mixed* across chips (all 12 chips in the subset carry both cases and
  controls, by design of the [subset selection](00_setup.qmd)). Chip effects can
  therefore be adjusted without harming the PTSD contrast.
- **chip × sex = `r round(cramV(md$slide, md$sex),2)`** — nearly total. **In the
  Grady Trauma Project, samples were plated by sex**: 135 of the 137 chips in the
  full study are single-sex, and our subset faithfully reproduces this (10 of 12
  chips are single-sex). Chip and sex are essentially the same variable.

::: {.callout-important}
## Why the chip × sex confound matters
Sex is a strong source of variation in methylation data, so as a rule sex should be accounted for
in the model design. Here chip is collinear with sex. If you put both `chip`
and `sex` in the design as fixed effects, the model cannot separate them and the
sex coefficient and the chip coefficients become unidentifiable.[@leek2010sva_confound].
:::

## 4. Choosing a correction strategy {#sec-strategy}

### The correction methods, compared

**ComBat** [@johnson2007combat] is the most widely used named-batch corrector. For a
batch factor you *recorded* (plate, chip, scan date), it fits a location-and-scale
model per batch and uses empirical Bayes to shrink each batch's mean and variance
toward the global estimate — borrowing strength across probes so small batches are
corrected stably. You protect the biology by passing a model matrix of covariates to
keep (`mod`), and ComBat removes only the batch-associated shift:

```r
library(sva)
mod <- model.matrix(~ ptsd + sex + age + Neu + CD4T + CD8T + Bcell + Mono, data = md)
M_combat <- ComBat(dat = M, batch = md$plate, mod = mod)   # M = M-value matrix
```

ComBat is fast, matrix-native, and the default in many pipelines. But it applies a
**global, parametric adjustment**: every sample in a batch is shifted by the same
estimated location/scale, under the assumption that the batch effect is a single
additive+multiplicative offset shared by all probes in that batch. When that
assumption doesn't hold, the global shift can *introduce*
distortion, and because it also alters the variance structure it can inflate false
positives if the design is unbalanced. This is the core reason some analysts prefer
to leave the data untouched and instead **model** the batch (as a covariate or via
surrogate variables), rather than overwrite the matrix.

::: {.callout-note}
## An alternative philosophy: distribution-aware correction (BeCLEAR)
Not everyone accepts the global-shift assumption. Methods such as **BeCLEAR** take a
different tack: rather than shifting every sample by one batch-level offset, they
detect batch effects by comparing **distributions** — using the
Kolmogorov–Smirnov test and **median differences** between batches at each probe —
and correct only where a distributional difference is actually present. The appeal is
that it does not force a uniform parametric adjustment on probes that don't need one.

However, BeCLEAR was **not designed to scale** to large EPIC datasets — the per-probe,
per-batch distributional testing is computationally heavy — and it is far less widely
validated and supported than ComBat or SVA. For a teaching pipeline, and for most
production EWAS, we stay with the standard tools; but it is worth knowing that the
"global adjustment" of ComBat is a *choice*, not a law, and that distribution-aware
alternatives exist for cases where it matters.
:::


**SVA and RUV** — latent-variable correction:

- **SVA** [@leek2012sva] discovers **unnamed** structure and returns surrogate
  variables you add as covariates. Use it for unknown or unmeasured technical or
  even biological variation (such as smoking) which can cause systemic methylation
  changes that confound the association testing.
- **RUV** [@gagnon2012ruv] uses negative-control probes (or replicate samples) to
  estimate unwanted variation. A good choice when reliable control probes exist.

Latent variable methods are not mutually exclusive with ComBat: a common recipe is ComBat for a clean,
recorded, un-confounded plate effect, then SVA for whatever residual structure
remains. That is close to what we do here — with one modification forced by the
chip × sex confound, described next.


### The design we actually use: ComBat *within* sex strata

The confounding between Chip and sex adds a layer of complexity to adjusting for
batch effects in the Grady Trauma Project. While this particular design flaw may not
be common, it highlights the importance of looking critically at the relationship between
technical and biological variables relevant to your analysis. Another dataset may not have
confounding between chip and sex, but may have confounding by something else, or may not have
any confounding at all! What matters is that you check for it so that you can address it appropriately.

Here, chip is a known variable, so ideally we would use ComBat to adjust for this. However, because of the collinearity with sex,
correcting chip across the whole cohort would partly remove biological signal from sex. Passing sex in ComBat's
`mod` is supposed to 'protect' it, but this only works when the batch and the covariate are not confounded.

There are a number of ways this could be handled, but the way we chose and recommend is to **stratify the correction**.
Within the female samples alone, chip is no longer confounded with sex — there is only one sex present — so chip is a clean
batch factor, and the same holds within the males. So we run ComBat on `slide`
separately in each sex stratum and then recombine:

```r
## ComBat on chip, separately within each sex stratum, then recombine.
## Within a stratum, chip is no longer confounded with sex.
##
## This is the call; it is *run* in section 6, not here, because `smoke` -- the
## methylation-derived smoking covariate ComBat has to protect -- is not built
## until section 5. Nothing about the design changes in between.
M_adj <- M
for (s in c("F", "M")) {
  i <- which(md$sex == s)
  mod_s <- model.matrix(~ ptsd + age + smoke + Neu + NK + CD4T + CD8T + Bcell + Mono,
                        data = md[i, ])
  M_adj[, i] <- sva::ComBat(dat = M[, i], batch = factor(as.character(md$slide[i])),
                            mod = mod_s, mean.only = TRUE, par.prior = TRUE)
}
```

Two choices in that call are worth explaining:

- **`mean.only = TRUE`** corrects each chip's mean but not its variance. ComBat's
  default is to adjust both, and this is normally fine. In this specific case,
  there are two chips which have males and females present. When we split by sex to
  do the stratified ComBat adjustment, we end up with a problematic splitting of the
  sexes on one of the chips.

  ```{r}
  #| label: stratum-chip-sizes
  ## The 87 modeled samples, split the way ComBat will split them. `slide` is the
  ## Sentrix barcode of the chip, so it is the batch label itself; splitting it by
  ## sex shows the batch sizes each stratum's ComBat call will actually see.
  for (s in c("F", "M")) {
    tb <- table(as.character(mdk$slide)[mdk$sex == s])
    cat(sprintf("%s stratum: n = %2d across %d chips | chip sizes: %s\n",
                s, sum(tb), length(tb), paste(sort(as.vector(tb)), collapse = ", ")))
  }
  ## A chip contributing one sample to a stratum is a batch of size one, which is
  ## the situation that forces the mean.only choice discussed above.
  cat("\nchips contributing a single sample to a stratum:\n")
  for (s in c("F", "M")) {
    tb <- table(as.character(mdk$slide)[mdk$sex == s])
    if (any(tb == 1)) cat(sprintf("  %s: %s\n", s, paste(names(tb)[tb == 1], collapse = ", ")))
  }
  ```

  Chip `201114400024` carries 6 females and 1 male. In the **male** stratum that
  chip is a batch of size **one**, and a batch of one has no within-batch variance
  to estimate — the scale parameter for that chip is unidentifiable. `sva::ComBat`
  checks for exactly this and switches to `mean.only = T` from its default of `mean.only = F`.
  So, if we did not explicitly specify `mean.only = T`, the female arm would end up being
  corrected differently from the male arm.

  Note where that singleton came from. Across all 96 arrays this chip carries 6
  females and **2** males; one of those two males is among the nine samples with no
  PTSD status ([chapter 01](01_qc.qmd)), so it leaves the modeled set and the chip
  drops to a single male. The incomplete phenotype did not just cost nine
  samples — it changed which batch corrections are estimable. Missing metadata
  propagates into methods choices in ways that are easy to miss if you only look at
  the final n.

- **`par.prior = TRUE`** uses the parametric empirical-Bayes prior. It is the
  stable choice at this sample size; the non-parametric alternative needs more
  samples per batch than we have.

The correction works: chip explains R² ≈ 0.20 (female) and 0.14 (male) of the
leading residual structure *before* using ComBat and is reduced after, and PTSD retains R² ≈ 0.01 in both strata —
the batch signal comes down without affecting the exposure.

### Array position

Chip position (`array_pos`) is the second measured technical factor, and it is *not* confounded
with anything biological here. While in theory we could use ComBat to adjust for this as well in
the chip-adjusted-recombined dataset, in practice this is not the best choice. The array_pos only has 8 levels,
so it's cost in degrees of freedom (7) in modeling array_pos as a covariate is acceptable for the sample size we are testing in.



## 5. Constructing a smoking proxy when smoking is not recorded {#sec-smoking}

Smoking is the single largest environmental driver of the blood methylome
[@joehanes2016smoking; @zeilinger2013smoking], and it is comorbid with PTSD. But, GSE132203's public phenotype table has no
smoking variable — a very common situation with GEO datasets. Fortunately, this
is something we can address with the methylation data itself.

We can take CpGs that the literature has established as smoking-associated,
and summarize their methylation into a single continuous score. We use a 20-CpG panel
of loci replicated across the large smoking EWAS
[@joehanes2016smoking; @zeilinger2013smoking; @shenker2013epic; @elliott2014smoking],
anchored on `cg05575921` in *AHRR* — the most reproducible smoking-associated CpG in
blood:

```{r}
#| label: smoke-panel
# The panel, grouped by locus so the provenance of each CpG stays visible. These
# are the loci that replicate across the smoking EWAS cited above; grouping also
# makes it obvious that the score is not driven by a single gene.
panel <- list(
  AHRR      = c("cg05575921","cg21161138","cg23576855","cg25648203","cg26703534","cg11902777"),
  F2RL3     = c("cg03636183","cg21911711"),
  GPR15     = c("cg19859270"),
  `2q37.1`  = c("cg05951221","cg21566642","cg01940273","cg03329539","cg06126421"),
  GFI1      = c("cg09935388","cg12876356","cg18146737","cg06338710"),
  MYO1G     = c("cg12803068","cg22132788","cg04180046","cg19089201"),
  PRSS23    = c("cg14391737"),
  RARA      = c("cg17739917")
)
# Not every published CpG survives chapter 03's probe filtering, so the panel is
# intersected with the probes actually present. `panel_probes` is what the score
# is built from -- and, below, what has to come out of the tested set.
pv <- unlist(panel, use.names = FALSE)
panel_probes <- pv[pv %in% rownames(M)]

cat("Panel CpGs (n =", length(panel_probes), "):\n")
cat(strwrap(paste(panel_probes, collapse = ", "), width = 78), sep = "\n")
cat("published CpGs dropped by probe filtering:",
    paste(setdiff(pv, panel_probes), collapse = ", "), "\n")
```

```{r}
#| label: smoke-proxy
## PC1 of the replicated smoking panel, oriented so that higher = heavier exposure.
## Built on the 87 modeled arrays, so the score describes the variance of the set
## the EWAS is actually fit on.
##
## The panel betas are read back off `grs` rather than taken from `M`: 20 rows of
## a 756,273-row object is a cheap subset, and the score is defined on the beta
## scale.
Bp <- getBeta(grs[panel_probes, keep])

## Each probe is z-scored across samples first, so that a probe with a wide beta
## range does not dominate PC1 purely because of its scale. scale. = FALSE in
## prcomp() then, because the scaling has already happened.
Z  <- t(scale(t(Bp)))
Z  <- Z[rowSums(is.na(Z)) == 0, , drop = FALSE]
pc <- prcomp(t(Z), center = TRUE, scale. = FALSE)
smoke     <- pc$x[, 1]
panel_pve <- pc$sdev^2 / sum(pc$sdev^2)

## AHRR cg05575921 is hypomethylated in smokers, so flip the sign if needed
## to make the score increase with exposure rather than decrease.
if (cor(smoke, Bp["cg05575921", ], use = "complete.obs") > 0) smoke <- -smoke
## Standardized, so the proxy enters every model below in units of one SD.
smoke <- as.numeric(scale(smoke))

cat("panel PC1 variance explained:", sprintf("%.1f%%", 100 * panel_pve[1]), "\n")
cat("correlation of the score with the AHRR anchor:",
    round(cor(smoke, Bp["cg05575921", ], use = "complete.obs"), 3),
    "(negative = higher score means less AHRR methylation)\n")
```

PC1 captures `r sprintf("%.0f%%", 100*max(panel_pve))` of the variance across the
panel, consistent with the panel behaving as one coherent axis rather than 20
independent probes.


```{r}
#| label: tested-matrix
# `Mk` is the matrix everything downstream is estimated on: the 87 modeled arrays,
# with the proxy's own panel probes removed and any probe that is non-finite in
# *any* of those 87 samples dropped -- ComBat and sva() both refuse an Inf.
excl <- rownames(M) %in% panel_probes
Mk   <- M[!excl, keep, drop = FALSE]
Mk   <- Mk[rowSums(!is.finite(Mk)) == 0, , drop = FALSE]

excluded_n    <- sum(excl)
tested_probes <- nrow(Mk)
cat("excluded panel probes:", excluded_n,
    "| non-finite probes dropped:", nrow(M) - excluded_n - tested_probes,
    "| probes carried into modeling:", format(tested_probes, big.mark = ","), "\n")

# The full 96-sample matrix is no longer needed; `top` keeps the variable-probe
# names for the before/after diagnostic in section 6.
rm(M); invisible(gc())
```

### The panel probes must be excluded from the EWAS
A proxy built from methylation is a linear combination of methylation. If those same
20 CpGs stay in the tested probe set, you are regressing each of them partly on
itself — guaranteed significance that means nothing. We therefore drop all
`r excluded_n` panel probes before testing. Together with two probes whose
M-values are non-finite in this 87-sample subset (`cg17759086`, `cg01801182` —
β pinned at 0 or 1, so log₂(β/(1−β)) is ±∞), that is why the EWAS reports
**`r format(tested_probes, big.mark = ",")`** CpGs rather than the 756,273 that
survived filtering in [chapter 03](03_probe_filtering.qmd).

The same logic applies to any methylation-derived covariate — epigenetic age, cell
proportions estimated from the same array, an EpiSmokEr score
[@bollepalli2019epismoker]. Know which probes went into it, and take them out.


::: {.callout-note}
## A proxy is not the phenotype
This score is a *measurement of methylation that correlates with smoking*, not a
measurement of smoking. There are two characteristics of this proxy score worth stating. First, it is a continuous
variable vs a categorical. A score doesn't distinguish "current smoker" from "recently quit heavy smoker". It is possible
to convert this continuous scale into categories, and some tools that do return 'smoker' vs 'non-smoker', if this would
be preferred to a continuous scale. Second, some
smoking-associated CpGs act partly through shifts in cell composition
[@bauer2015gpr15], so the proxy and the cell-proportion covariates are not fully
independent. If real smoking data exists (and is reliable), use it; a proxy is what you do when it
doesn't.
:::

## 6. Surrogate Variable Analysis, done correctly {#sec-sva}

After doing ComBat for chip, adding array_pos as a covariate, and creating the smoking proxy,
we then use SVA to create surrogate variables for any additional unknown/unmeasured variation.
For example, we know that plate and well can cause technical variation, but these were not
reported in the Grady Trauma Project phenotypes, so we can use SVA to hopefully capture any
remaining technical variation from these types of batch effects if they are present. SVA is run on
M-values with a **full model** (everything you want to keep) and a
**null model** (everything except the exposure). We 'protect' PTSD, sex, age, the
smoking proxy, array position, and the six estimated cell-type proportions:

```{r}
#| label: combat-run
# The stratified ComBat of section 4, run now that `smoke` exists. About 57
# seconds for the two strata on this matrix.
#
# droplevels() matters twice over: a stratum carries only some of the 12 chips, and
# ComBat cannot estimate a location for a batch level with no samples in it.
sexk    <- factor(mdk$sex)
slidek  <- droplevels(factor(as.character(mdk$slide)))
cellk   <- as.data.frame(propk)[, c("Neu", "NK", "CD4T", "CD8T", "Bcell", "Mono")]

Mcb <- Mk
for (s in levels(sexk)) {
  i  <- which(sexk == s)
  bs <- droplevels(slidek[i])
  # sex is constant inside a stratum, so it is not -- and cannot be -- in cb_mod.
  # Everything else we intend to keep is passed here so ComBat does not remove it
  # along with the chip shift.
  d  <- cbind(cellk[i, , drop = FALSE],
              ptsd = mdk$ptsd[i], age = as.numeric(mdk$age)[i], smoke = smoke[i])
  cb_mod <- model.matrix(~ ptsd + age + smoke + Neu + NK + CD4T + CD8T + Bcell + Mono,
                         data = d)
  cat(sprintf("%s stratum: n = %d | chips = %d | smallest chip = %d\n",
              s, length(i), nlevels(bs), min(table(bs))))
  Mcb[, i] <- ComBat(dat = Mk[, i, drop = FALSE], batch = bs, mod = cb_mod,
                     par.prior = TRUE, mean.only = TRUE, prior.plots = FALSE)
}
```

```{r}
#| label: combat-effect
# Did the correction do what it was supposed to, and did sex survive it?
#
# Per-probe R^2 against a factor, for a whole matrix in one call: lm.fit() on the
# transposed matrix fits all probes simultaneously, which is the difference
# between seconds and an hour of looping lm() over 20,000 probes.
mean_r2 <- function(X, f) {
  Y   <- t(X)
  rss <- colSums(resid(lm.fit(model.matrix(~ f), Y))^2)
  tss <- colSums(sweep(Y, 2, colMeans(Y))^2)
  mean(1 - rss / tss)
}

# Judge the correction on the same 20,000 variable probes the PCA used -- that is
# where batch structure is visible in the first place.
diag_probes <- intersect(top, rownames(Mk))
report <- function(label, f)
  cat(sprintf("mean per-probe R2 vs %-22s %.4f -> %.4f\n", label,
              mean_r2(Mk[diag_probes, ], f), mean_r2(Mcb[diag_probes, ], f)))

report("chip (should fall)", slidek)
report("sex (must survive)", sexk)          # ComBat was never told about sex
report("PTSD (must survive)", mdk$ptsd)     # the exposure must be untouched
```

```{r}
#| label: sva-fit
# Array position enters as a fixed covariate rather than being ComBat-ed: all
# eight levels are present, unconfounded, and cost only 7 degrees of freedom.
pos <- factor(as.character(mdk$array_pos))
mdm <- cbind(cellk, ptsd = mdk$ptsd, sex = sexk, age = as.numeric(mdk$age),
             smoke = smoke, pos = pos)

mod  <- model.matrix(~ ptsd + sex + age + smoke + pos + Neu + NK + CD4T + CD8T + Bcell + Mono, data = mdm)
mod0 <- model.matrix(~        sex + age + smoke + pos + Neu + NK + CD4T + CD8T + Bcell + Mono, data = mdm)
stopifnot(qr(mod)$rank == ncol(mod))       # a rank-deficient full model makes sva() meaningless

# num.sv()'s Buja-Eyuboglu test on the uncorrected matrix: how many dimensions of
# structure are *detectable* before any correction. This and the sva() call below
# are the expensive pair in the chapter -- together roughly 5 minutes, peaking
# near 8.9 GB, because both work on the full 756,251 x 87 matrix.
n_sv_be <- num.sv(Mk, mod, method = "be")

# We fit 6, not n_sv_be -- see the k-selection sweep below for why. The sweep
# compared these candidates; 6 is the one adopted.
k_candidates <- c(6, 8, 10, 15)
k_selected   <- 6
svobj <- sva(Mcb, mod, mod0, n.sv = k_selected)
SV <- svobj$sv                            # add these as covariates in the EWAS
colnames(SV) <- paste0("SV", seq_len(ncol(SV)))
cat("num.sv (Buja-Eyuboglu):", n_sv_be, "| surrogate variables carried forward:",
    ncol(SV), "\n")
```

Note what is in the **null** model as well as the full one. Anything you put in
`mod0` is variation SVA is told *not* to bother capturing, because you are already
modeling it explicitly.

Note that SVA runs on the **full filtered probe set**, not the 20,000-probe subset
used for the PCA diagnostic above. The PCA subset exists only to make the
variable-association heatmap legible; the surrogate variables that go into the model
are estimated from every probe that survived filtering.

### How many surrogate variables? (choosing *k*)

`num.sv`'s Buja–Eyuboglu permutation test suggests **`r n_sv_be`** SVs here.
We fit **`r k_selected`**. `num.sv` answers *how many dimensions of residual structure are statistically
detectable* — and with 756,251 probes, plenty are detectable, but are also tiny.
What we need to know is *how many are worth a degree of freedom* in modeling, and at n = 87,
degrees of freedom are a constraint. So, we swept k over
`r paste(k_candidates, collapse = ", ")` and judged each fit on two things: calibration (genomic inflation λ, which should
sit near 1) and precision (the median standard error of the PTSD coefficient,
which should fall if the SVs are removing real noise).

```{r}
#| label: k-sweep-compute
# This is the sweep behind data/05_k_lambda_sweep.csv, and it runs at render.
# Each candidate k gets its own sva() fit plus a full-probe limma fit, but n.sv is
# supplied here, so each fit skips the Buja-Eyuboglu estimation that num.sv() does
# above -- which is what makes the four candidates together cost about 140
# seconds (measured: 136.9 s), rather than the several minutes the single
# estimated fit above took on its own. The peak is comparable to that fit's (~9 GB). The small table
# it writes is committed to the repository, and the chunk below reads it back, so
# the figure and the text quote one set of numbers.

# Genomic inflation: the median observed chi-squared over its null expectation.
# 1 is calibrated, > 1 inflated, < 1 deflated.
lam <- function(p) median(qchisq(1 - p, 1), na.rm = TRUE) / qchisq(0.5, 1)
coefn <- "ptsdCase"          # the PTSD contrast, with Control as the reference level

# k = 0 is the reference point for every SE comparison: the same design without
# any surrogate variables.
fit0 <- eBayes(lmFit(Mcb, mod))
tt0  <- limma::topTable(fit0, coef = coefn, number = Inf, sort.by = "none")
se0  <- sqrt(fit0$s2.post) * fit0$stdev.unscaled[, coefn]

rows <- list(data.frame(
  k = 0, npar = ncol(mod), resid_df = nrow(mdk) - ncol(mod),
  lambda = lam(tt0$P.Value), med_se = median(se0), se_ratio = 1, frac_gain = NA,
  n_p1e5 = sum(tt0$P.Value < 1e-5), n_fdr = sum(tt0$adj.P.Val < 0.05),
  n_bonf = sum(tt0$P.Value < 0.05 / nrow(tt0)),
  r2_slide = NA, r2_smoke_sv = NA, max_r2_sv_ptsd = NA, strata_full_rank = TRUE,
  df_F = 45 - (ncol(mod) - 1), df_M = 42 - (ncol(mod) - 1)))

for (K in k_candidates) {
  set.seed(42)                                  # same seed at every k
  SVk <- sva(Mcb, mod, mod0, n.sv = K)$sv       # SVA's native order, first K
  colnames(SVk) <- paste0("SV", seq_len(K))
  dm  <- cbind(mod, SVk)
  fit <- eBayes(lmFit(Mcb, dm))
  tt  <- limma::topTable(fit, coef = coefn, number = Inf, sort.by = "none")
  se  <- sqrt(fit$s2.post) * fit$stdev.unscaled[, coefn]

  # What chip structure is still left once this design has been fitted: residuals
  # first, then how much of them chip explains.
  R <- t(resid(lm.fit(dm, t(Mcb[diag_probes, ]))))

  # The stratified fits in chapter 07 drop sex and refit per stratum, so check
  # that the design is still full rank there before adopting a k.
  full_rank <- all(sapply(levels(sexk), function(lv) {
    i  <- which(sexk == lv)
    ds <- cbind(model.matrix(~ ptsd + age + smoke + pos + Neu + NK + CD4T + CD8T +
                               Bcell + Mono, data = mdm)[i, , drop = FALSE],
                SVk[i, , drop = FALSE])
    qr(ds)$rank == ncol(ds)
  }))

  rows[[length(rows) + 1]] <- data.frame(
    k = K, npar = ncol(dm), resid_df = nrow(mdk) - ncol(dm),
    lambda = lam(tt$P.Value), med_se = median(se),
    se_ratio = median(se / se0),
    # fraction of probes whose SE improved relative to k = 0
    frac_gain = mean(se < se0),
    n_p1e5 = sum(tt$P.Value < 1e-5), n_fdr = sum(tt$adj.P.Val < 0.05),
    n_bonf = sum(tt$P.Value < 0.05 / nrow(tt)),
    r2_slide = mean_r2(R, slidek),
    # how much of the named smoking covariate the SVs have re-absorbed
    r2_smoke_sv = summary(lm(smoke ~ SVk))$r.squared,
    max_r2_sv_ptsd = max(sapply(seq_len(K), function(j)
      summary(lm(SVk[, j] ~ mdk$ptsd))$r.squared)),
    strata_full_rank = full_rank,
    # residual df left in each sex stratum, which is the real cost of large k
    df_F = 45 - (ncol(dm) - 1), df_M = 42 - (ncol(dm) - 1))
}
ks <- do.call(rbind, rows)
write.csv(ks, "data/05_k_lambda_sweep.csv", row.names = FALSE)
```

```{r}
#| label: k-sweep
# Read the committed sweep back. It is a few hundred bytes and lives in the
# repository, so this works whether or not you ran the chunk above.
ks <- read.csv("data/05_k_lambda_sweep.csv")
knitr::kable(
  ks[, c("k", "npar", "resid_df", "lambda", "med_se", "se_ratio", "r2_smoke_sv", "df_F", "df_M")],
  digits = c(0, 0, 0, 4, 5, 4, 3, 0, 0),
  col.names = c("k", "n par", "resid df", "lambda", "median SE", "SE ratio",
                "R2(smoke~SV)", "resid df F", "resid df M"),
  caption = "Surrogate-variable count sweep. `SE ratio` is the median standard error relative to k = 0; `R2(smoke~SV)` is how much of the smoking proxy the SVs collectively absorb.")
```

Reading down that table:

- **k = 0 is clearly wrong.** λ = 0.85 is badly *deflated*, the signature of
  unmodeled structure inflating residual variance and washing out real
  differences.
- **k = 6 fixes the calibration** (λ = 1.02) and buys most of the precision on
  offer. The median SE drops
  `r sprintf("%.1f%%", 100*(1 - ks$se_ratio[ks$k==6]))` against k = 0, and
  `r sprintf("%.0f%%", 100*ks$frac_gain[ks$k==6])` of probes come out with a
  smaller SE than they had at k = 0 — that second figure is the `frac_gain`
  column, a count of probes that improved, not a share of the improvement. The
  largest median-SE drop anywhere in the sweep is
  `r sprintf("%.1f%%, at k = %d", 100*max(1 - ks$se_ratio, na.rm = TRUE), ks$k[which.max(1 - ks$se_ratio)])`,
  so k = 6 captures about
  `r sprintf("%.0f%%", 100*(1 - ks$se_ratio[ks$k==6])/max(1 - ks$se_ratio, na.rm = TRUE))`
  of what is available — while keeping four more residual degrees of freedom and
  a λ far closer to 1.
- **k = 8, 10, and 15 buy almost nothing and cost real power.** By k = 10, λ has
  drifted up to 1.07, which is still an acceptable lambda, and the SEs have barely moved,
  but the burden of degrees of freedom becomes problematic.
- **The SVs increasingly re-capture the smoking proxy.** R² of `smoke` on the SVs climbs
  from 0.39 at k = 6 to 0.64 at k = 15. Since we specifically want smoking as a
  named, interpretable covariate, letting latent variables absorb it is a loss,
  not a gain.

k = 6 is the smallest k that gets λ to 1 and that leaves the most residual degrees of freedom.



```{r}
#| label: sv-table
# What each retained SV tracks. A surrogate variable is only interpretable
# through its associations, so every candidate explanation gets a column --
# including PTSD, because an SV that tracks the exposure is absorbing signal
# rather than confounding.
r2 <- function(y, x) summary(lm(y ~ x))$r.squared
sv_table <- data.frame(
  SV       = seq_len(ncol(SV)),
  r2_slide = sapply(seq_len(ncol(SV)), function(j) r2(SV[, j], slidek)),
  r2_pos   = sapply(seq_len(ncol(SV)), function(j) r2(SV[, j], pos)),
  r2_smoke = sapply(seq_len(ncol(SV)), function(j) r2(SV[, j], smoke)),
  r2_ptsd  = sapply(seq_len(ncol(SV)), function(j) r2(SV[, j], mdk$ptsd)),
  r2_age   = sapply(seq_len(ncol(SV)), function(j) r2(SV[, j], as.numeric(mdk$age))),
  r2_sex   = sapply(seq_len(ncol(SV)), function(j) r2(SV[, j], sexk)))

knitr::kable(sv_table, digits = 4,
  col.names = c("SV", "R2 slide", "R2 position", "R2 smoke", "R2 PTSD", "R2 age", "R2 sex"),
  caption = "What each surrogate variable tracks. Chip (slide) and array position dominate; association with PTSD stays below 0.05 throughout.")
```

SV5 is strongly chip-linked (R² = 0.64) and SV4 next (R² = 0.34). Note that SV4 and SV5 also carry substantial sex
association (R² = 0.37 and 0.46) — a direct consequence of the chip × sex confound
documented above.

```{r}
#| label: sva-bars
#| fig-cap: "Association (R²) of each surrogate variable with chip (technical), sex (biological), and PTSD (the exposure). The SVs carry the chip structure — SV5 reaches R² = 0.64 — and the sex signal entangled with it, while every SV stays near-orthogonal to PTSD (largest R² = 0.04). This is SVA working as intended: it absorbs unwanted technical variation without touching the signal we want to test."
#| fig-width: 8.4
#| fig-height: 4.4
# Three of the sv_table columns, in long form. The factor levels are set
# explicitly so the legend reads technical -> biological -> exposure, which is
# the order the caption discusses them in.
bars <- rbind(
  data.frame(SV = sv_table$SV, r2 = sv_table$r2_slide, what = "Chip (technical)"),
  data.frame(SV = sv_table$SV, r2 = sv_table$r2_sex,   what = "Sex (biological)"),
  data.frame(SV = sv_table$SV, r2 = sv_table$r2_ptsd,  what = "PTSD (exposure)"))
bars$SV   <- factor(paste0("SV", bars$SV), levels = colnames(SV))
bars$what <- factor(bars$what, levels = c("Chip (technical)", "Sex (biological)",
                                          "PTSD (exposure)"))

p_sv <- ggplot(bars, aes(SV, r2, fill = what)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.7) +
  scale_fill_manual(values = c("Chip (technical)" = "#0F3D43",   # teal_dark
                               "Sex (biological)" = "#B8873F",   # sand
                               "PTSD (exposure)"  = "#8C3A4A"),  # plum
                    name = NULL) +
  # fixed 0-1 axis: the point of the figure is how small the PTSD bars are, which
  # a free axis would hide by rescaling them
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.04))) +
  labs(x = NULL, y = expression(R^2),
       title = "The surrogate variables absorb chip, not the exposure",
       subtitle = sprintf(paste("Buja-Eyuboglu suggested %d SVs on the full filtered matrix;",
                                "the %d carried into the model are shown"),
                          n_sv_be, ncol(SV))) +
  theme(legend.position = "top")
p_sv

ggsave("data/05_sva_bars.png", plot = p_sv, width = 8.4, height = 4.4, dpi = 200)
```

```{r}
#| label: sva-summary
# The two numbers that decide whether the SVA was worth running: the SVs should
# be strongly chip-linked and near-orthogonal to the exposure.
chip_r2 <- sapply(1:ncol(SV), function(j) summary(lm(SV[,j] ~ factor(mdk$slide)))$r.squared)
ptsd_r2 <- sapply(1:ncol(SV), function(j) summary(lm(SV[,j] ~ factor(mdk$ptsd)))$r.squared)
cat(sprintf("Strongest chip-linked SV: SV%d (R2 = %.2f)\n", which.max(chip_r2), max(chip_r2)))
cat(sprintf("Largest SV-PTSD association: %.3f (all SVs are ~orthogonal to the exposure)\n",
            max(ptsd_r2)))
```

We carry these surrogate variables forward as covariates in the
[EWAS model](06_ewas.qmd), alongside sex, age, and cell composition.

::: {.callout-tip}
## Don't over-correct
More correction is not always better. Each surrogate variable spends a degree of
freedom; adding too many, or adding SVs that happen to correlate with your exposure,
removes real signal. `num.sv` gives a principled count; inspect the SV–exposure
associations (as above) and, if any SV correlates strongly with the phenotype, treat
that as a warning that the exposure and a batch are partially confounded.
:::

## Saving what we just computed

The PCA diagnostic, the stratified ComBat matrix, the smoking proxy and the surrogate
variables are all inputs to later chapters, so each one is written out here, by the
code that produced it.

```{r}
#| label: save-05
# 1. The PCA diagnostic. Small, and it is what the variable-association figures
#    are drawn from, so a reader can redraw them without re-running the PCA.
saveRDS(list(pcs = pcs, pve = pve, r2 = pc_r2, md = md, props = props,
             slide = slide, posrow = posrow),
        "data/05_batch_pca.rds")

# 2. The smoking proxy: the score itself, the panel it came from, and the PC1
#    loadings, so that what went into the covariate stays inspectable.
saveRDS(list(smoke = smoke, panel = panel, present = panel_probes,
             pve = panel_pve, loadings = pc$rotation[, 1]),
        "data/05_smoking_proxy.rds")

# 3. The model inputs chapters 06 and 07 read: the surrogate variables, the
#    87-sample metadata and cell proportions they were estimated alongside, and
#    the design record. `k_selected` is what the EWAS uses to slice `SV`.
saveRDS(list(SV = SV, n.sv = ncol(SV), keep = keep, mdk = mdk, propk = propk,
             smoke = smoke, panel_probes = panel_probes, panel_pve = panel_pve,
             excluded_n = excluded_n, tested_probes = tested_probes,
             prev_n_sv = n_sv_be, sv_table = sv_table,
             k_selected = k_selected, k_candidates = k_candidates,
             r2_smoke_sv = summary(lm(smoke ~ SV))$r.squared,
             chip_r2 = chip_r2, pos_r2 = sv_table$r2_pos,
             position_in_model = TRUE,
             design_note = paste("ComBat(slide) within sex strata;",
                                 "array position as fixed covariate; SVA k=6")),
        "data/05_sva.rds")

# 4. The corrected M-value matrix the EWAS is actually fit on. This one is
#    511 MB, which is why it is excluded from git and distributed as a Zenodo
#    tier rather than committed.
saveRDS(Mcb, "data/05_mvals_combat.rds")
```

::: {.callout-note collapse="true"}
## Picking the chapter up from the saved checkpoints instead

The two expensive steps above — the full M-value matrix and the `num.sv`/`sva` pair —
need about 9 GB of memory between them. If that is more than your machine has, the
published checkpoints are the same files this chapter just wrote:

```{.bash filename="Terminal"}
./get_data.sh D_filtered E_model_inputs
```

```{.r filename="RStudio Console"}
sva_o <- readRDS("data/05_sva.rds")
SV    <- sva_o$SV          # 87 x 6 surrogate variables
mdk   <- sva_o$mdk         # the 87 modeled samples
smoke <- sva_o$smoke       # the methylation-derived smoking score
Mcb   <- readRDS("data/05_mvals_combat.rds")   # stratified-ComBat M-values
```

What this route does not give you is the diagnostic reasoning — you inherit k = 6 and
the stratified design as decisions already made, rather than watching the evidence for
them accumulate.
:::

## 7. When you don't have IDATs: batch correction from a processed matrix {#sec-no-idats}

Batch diagnosis and latent-variable correction work on a **β/M-matrix**, so the
non-IDAT path is largely the same — with two caveats:

1. **You may not have the batch annotation.** Chip and position come from the IDAT
   filenames / Sentrix barcodes. If the submitter deposited only a processed matrix,
   check the sample metadata (`pData`) for `Sentrix_ID`, `Sentrix_Position`,
   `plate`, `Slide`, or a scan-date column — many deposit these even without IDATs.
   If they are absent, you **cannot reconstruct** the named batches, so a surrogate 
   variable method is more appropriate.
2. **Diagnose and correct on M-values from the matrix.** PCA + variable-association,
   `sva`, and `ComBat` all take a matrix. Convert deposited β to M-values first
   (see [normalization](02_normalization.qmd)), run the identical SVA recipe, and
   add the surrogate variables to your model:

   ```r
   M <- log2(beta / (1 - beta))          # from a processed beta-matrix
   mod  <- model.matrix(~ exposure + covariates, data = pData)
   mod0 <- model.matrix(~ covariates, data = pData)
   SV   <- sva(M, mod, mod0)$sv
   ```

`ComBat` is likewise matrix-native (`sva::ComBat(dat = M, batch = plate, mod = mod)`),
so if a *clean, recorded, un-confounded* batch variable is present in a
matrix-only deposit, you can correct it directly.

---

**Next:** [EWAS analysis](06_ewas.qmd) — fit the association model for PTSD with
sex, age, cell composition, and surrogate variables as covariates, and interpret the
results honestly.

## References {.unnumbered}

::: {#refs}
:::
