---
title: "Epigenome-Wide Association Studies: a hands-on tutorial"
subtitle: "From raw IDATs to annotated hits, using the Grady Trauma Project (GSE132203)"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
ss <- readRDS("data/sample_sheet.rds")
ss <- ss[order(ss$slide, ss$array_pos), ]
```

## Who this is for

This tutorial is written for rotation students and anyone new to array-based
DNA methylation analysis. It walks the full path from **raw Illumina IDAT
files** to a set of **annotated, genome-wide association results**, running
every step for real on a teaching subset of a public dataset. By the end you
will have:

- read raw two-channel IDATs into R and understood what the numbers mean;
- run the quality-control checks that decide which samples and probes to trust;
- normalized the data and filtered unreliable probes using ancestry-aware masks;
- estimated blood cell composition and diagnosed technical batch structure;
- fit an epigenome-wide model, run it reproducibly as a Snakemake pipeline, and annotated the top CpGs and regions.

## The dataset

We use **GSE132203**, the *Grady Trauma Project* — whole-blood Illumina
MethylationEPIC (v1, 850K) profiles from an adult cohort deeply phenotyped for
trauma exposure and PTSD. It was chosen for teaching because it is unusually
complete:

| Feature | Why it matters for teaching |
|---|---|
| Raw IDATs deposited | Lets us start from the true beginning, not a processed matrix |
| N = 795 (we use a **96-sample subset**) | Big enough to be realistic, small enough to run live |
| Deposited cell-type proportions | An **answer key** to validate our own deconvolution |
| Sentrix barcodes in filenames | Real chip/position structure for the batch-effects lesson |
| Explicit age & sex, multiple exposures | Proper covariate modeling, several candidate outcomes |

## How the teaching subset was chosen

The Grady Trauma Project was selected for this tutorial in specific because I wanted
to be able to provide a walkthrough that shows:

1. How to process from raw IDAT files,
2. Uses whole blood as the tissue source since this is a common tissue used in methylation studies and requires some additional consideration in statistical modeling.
3. Has available phenotype data for a disease or trait association testing, as well as sex and age demographic information.

The resulting balance of the 96 samples chosen for the tutorial subset:

```{r}
#| label: subset-balance
bal <- list(
  PTSD          = table(ss$ptsd, useNA = "ifany"),
  Sex           = table(ss$gender),
  Race          = table(ss$race, useNA = "ifany")
)
bal
cat("\nAge: range", paste(range(ss$age, na.rm=TRUE), collapse="-"),
    "| mean", round(mean(ss$age, na.rm=TRUE), 1), "\n")
cat("Chips carrying both PTSD classes:",
    sum(tapply(ss$ptsd, ss$slide, function(x) length(unique(na.omit(x)))>1)),
    "of", length(unique(ss$slide)), "\n")
```

::: {.callout-note}
## A bonus lesson I can teach with the Grady Trauma Project
This cohort is ~95% African American, which means in addition to the main tutorial
lessons I wanted to achieve, I also get to showcase using **population-specific**
SNP (single nucleotide polymorphism) masks. Differences in genomic architecture is
observed across populations; a variant may be polymorphic in one population, but is
not in another, and vice-versa. SNPs in close proximity to a CpG can impact the
ability for the CpG probes to measure the methylation status accurately. Thus, CpGs
near common SNPs are typically excluded from downstream analyses. However, not all
populations share the same SNPs and masking with the wrong population may cause you
to lose true-positive signals and retain false-positive signals.
:::

## How the notebooks fit together

The preprocessing-through-analysis arc maps onto these notebooks:

1. **[Background](00a_background.qmd)** — what DNA methylation is and how arrays measure it. Skip it if you already work with methylation data.
2. **[Setup](00_setup.qmd)** — downloading the raw IDATs, reading them into R, and what the whole tutorial costs in time, memory, and disk.
3. **[Quality control](01_qc.qmd)** — detection p-values, intensity, predicted sex, SNP identity.
4. **[Normalization](02_normalization.qmd)** — within- and between-array normalization.
5. **[Probe filtering](03_probe_filtering.qmd)** — detection-p, cross-reactive, and ancestry-aware SNP masks.
6. **[Cell composition](04_cell_composition.qmd)** — reference-based deconvolution, validated against deposited proportions.
7. **[Batch effects](05_batch_effects.qmd)** — chip and array-position diagnosis; ComBat, and surrogate variables for what remains.
8. **[EWAS](06_ewas.qmd)** — the association model, adjusted for sex, age, smoking, array position, cell composition, and surrogate variables.
9. **[Snakemake pipeline](07_pipeline.qmd)** — the same analysis as a versioned, one-command reproducible workflow, with BACON and sex-stratified meta-analysis.
10. **[Annotation](08_annotation.qmd)** — mapping top CpGs to genes, regions (DMRs), and regulatory context.

Each notebook loads the object saved by the previous one, so you can work
through them in order or jump to any step and pick up the saved checkpoint.
